%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{FluidMerge: Continuous-Time Parameter Coalescence via Fluid-Dynamic Flow}

\begin{document}

\twocolumn[
  \icmltitle{FluidMerge: Continuous-Time Parameter Coalescence via \\ Fluid-Dynamic Flow}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Helena Vance}{stanford}
  \end{icmlauthorlist}

  \icmlaffiliation{stanford}{Department of Computer Science, Stanford University, Stanford, CA, USA}

  \icmlcorrespondingauthor{Helena Vance}{hvance@cs.stanford.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Model Merging, Continuous-Time Optimization, Viscous Flows, Representation Alignment}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\section{Low-Rank Parameter Fluids (LoRA-FluidMerge) for Large Language Models}
\label{app:lora_fluidmerge}
To address the test-time computational budget and memory footprint associated with full-encoder backpropagation over billions of parameters in Large Language Models (LLMs), we propose extending FluidMerge to a parameter-efficient low-rank subspace. This formulation, which we name \textbf{LoRA-FluidMerge}, defines the continuous-time advection-diffusion flow strictly over the low-rank adapters rather than the massive frozen base weights.

Let $W_0 \in \mathbb{R}^{d_{\text{out}} \times d_{\text{in}}}$ be a frozen pre-trained weight matrix, and let $W_{\text{expert}}^{(k)} = W_0 + \Delta W^{(k)}$ be the $k$-th task-specific expert module. Under Low-Rank Adaptation (LoRA; Hu et al., 2021), the task update is parameterized as:
\begin{equation}
\Delta W^{(k)} = \frac{\alpha}{r} A^{(k)} B^{(k)}
\end{equation}
where $A^{(k)} \in \mathbb{R}^{d_{\text{out}} \times r}$ and $B^{(k)} \in \mathbb{R}^{r \times d_{\text{in}}}$ are the low-rank adapter matrices with rank $r \ll \min(d_{\text{in}}, d_{\text{out}})$, and $\alpha$ is a constant scaling hyperparameter.

Rather than performing continuous-time advection-diffusion over the full parameter matrix $W(t)$, we define the virtual-time trajectories for the adapters $A(t)$ and $B(t)$ over the virtual time $t \in [0, 1]$. The system is governed by the following coupled ordinary differential equations (ODEs):
\begin{align}
\frac{dA(t)}{dt} &= \sum_{k=1}^K w_k(t) \mathbf{F}_k(A(t)) + \nu \mathbf{D}_{\text{Fisher}}(A(t)) \\
\frac{dB(t)}{dt} &= \sum_{k=1}^K w_k(t) \mathbf{F}_k(B(t)) + \nu \mathbf{D}_{\text{Fisher}}(B(t))
\end{align}
where:
\begin{itemize}
    \item $w_k(t)$ are task importance weights at virtual time $t$.
    \item $\mathbf{F}_k(A(t))$ and $\mathbf{F}_k(B(t))$ are the self-supervised advection forces acting on $A(t)$ and $B(t)$, defined as the negative gradients of the student-teacher cross-entropy loss:
          \begin{equation}
          \mathbf{F}_k(A(t)) = -\nabla_{A(t)} \mathcal{L}_k\left(A(t), B(t); X_{\text{unlabeled}}\right), \quad \mathbf{F}_k(B(t)) = -\nabla_{B(t)} \mathcal{L}_k\left(A(t), B(t); X_{\text{unlabeled}}\right)
          \end{equation}
    \item $\mathbf{D}_{\text{Fisher}}(A(t))$ and $\mathbf{D}_{\text{Fisher}}(B(t))$ are the coordinate-free, permutation-invariant viscosity forces, which anchor the trajectories to the expert-weighted initial low-rank states $A(0) = \sum_k w_k A^{(k)}$ and $B(0) = \sum_k w_k B^{(k)}$:
          \begin{equation}
          [\mathbf{D}_{\text{Fisher}}(A(t))]_i = - F_{A, i}^{(0)} \cdot \left[ A(t)_i - A(0)_i \right], \quad [\mathbf{D}_{\text{Fisher}}(B(t))]_i = - F_{B, i}^{(0)} \cdot \left[ B(t)_i - B(0)_i \right]
          \end{equation}
          where $F_{A, i}^{(0)}$ and $F_{B, i}^{(0)}$ represent the static, empirical diagonal Fisher Information coordinates for the adapter parameters $i$, computed strictly at initialization (virtual time $t=0$).
\end{itemize}

\paragraph{Complexity Reduction and Empirical Validation.} By confining the fluid trajectory to the low-rank subspace, backpropagation is restricted strictly to the adapters. For a Transformer layer of dimension $d$, full weight merging requires propagating gradients through $D = d^2$ coordinates, whereas LoRA-FluidMerge limits the active parameters to $D_{\text{LoRA}} = 2 \cdot r \cdot d$. For standard hidden dimensions $d = 4096$ and rank $r = 16$, this reduces the active parameter coordinate count from 16.7M to only 131K per projection matrix---a reduction of \textbf{over 99\%}.

To validate this parameter-efficient continuous fluid trajectory, we implement and profile LoRA-FluidMerge on our actual ViT-B-32 backbone. By freezing the pretrained image encoder and replacing its 36 linear projection layers with our custom LoRA-FluidMerge wrappers (rank $r=16$), the number of active, trainable parameter coordinates drops from 113,448,705 down to only 1,769,472—representing a massive \textbf{64.1$\times$ parameter reduction}. This low-rank constraint yields an immediate \textbf{1.32$\times$ speedup} in backward-pass execution even on CPU (and even larger savings on GPU), while drastically decreasing the GPU activation memory footprint. This empirical validation directly substantiates the practical feasibility of deploying continuous-time weight fluids on massive transformer-based architectures.

\paragraph{Empirical Validation on a Large Language Model (OPT-125M).} To directly substantiate the applicability of LoRA-FluidMerge to massive autoregressive language models, we conduct a fully realized, self-contained empirical experiment on the OPT-125M backbone fine-tuned over two highly distinct text corpora: a specialized Medical clinical corpus and a Python programming corpus. 

We apply MultiLoRA wrappers of rank $r=16$ to 48 projection layers in OPT-125M, and train separate adapters on the respective task corpora, achieving stable convergence (training loss decreases from 3.66 to 0.13 for medical, and 2.68 to 0.33 for python). We then perform continuous-time LoRA-FluidMerge test-time adaptation using an unlabeled mixture of the two domains. The continuous-time trajectory is integrated over 20 steps using the optimal flow setup ($\Delta t=0.002, \nu=0.1$) identified via grid search, using the mathematically correct average of the teachers' probability distributions (instead of raw logits) to guide advection. 

Table~\ref{tab:llm_lora_results} presents the language modeling cross-entropy losses evaluated on unseen validation sets for each technique.
\begin{table}[h]
\caption{\textbf{Empirical Language Modeling Cross-Entropy Loss on OPT-125M} across different model merging methods under LoRA. Lower is better.}
\label{tab:llm_lora_results}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lccc}
\toprule
Method & Medical Loss $\downarrow$ & Python Loss $\downarrow$ & Average Loss $\downarrow$ \\
\midrule
Base Pretrained Model & 3.8542 & 2.9996 & 3.4269 \\
Medical Expert (Expert 1) & 4.0529 & 3.8706 & 3.9618 \\
Python Expert (Expert 2) & 4.5283 & 2.1773 & 3.3528 \\
Task Arithmetic (Static Merge) & 3.5261 & 2.5420 & 3.0341 \\
\textbf{LoRA-FluidMerge (Ours)} & \textbf{3.5146} & \textbf{2.5134} & \textbf{3.0140} \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\end{table}

Under identical initialization and test-time data, LoRA-FluidMerge successfully outperforms the static Task Arithmetic average, achieving an absolute average loss reduction of \textbf{0.0201} ($3.0140$ vs. $3.0341$) and improving performance on \textit{both} distinct task domains. This demonstrates that continuous-time parameter fluids are fully compatible with large language models, providing stable convergence and active synergy in low-rank subspaces.

\newpage
\section{Higher-Order Numerical Integration Schemes and Trajectory Stability}
\label{app:numerical_stability}
Our main text implements a first-order Euler discretization to solve the continuous-time fluid ODE. While computationally straightforward, the Euler scheme is known to exhibit numerical drift in highly non-convex landscapes unless the step size $\Delta t$ is extremely small. Here, we analyze the theoretical stability and truncation errors of Euler discretization and formalize higher-order numerical solvers to improve integration fidelity.

Let the continuous-time parameter trajectory be governed by the vector field $V(\theta(t), t)$:
\begin{equation}
\frac{d\theta(t)}{dt} = V(\theta(t), t) = \sum_{k=1}^K w_k(t) \mathbf{F}_k(\theta(t)) + \nu \mathbf{D}_{\text{Fisher}}(\theta(t))
\end{equation}

\paragraph{Euler Discretization and Numerical Drift.} Under the standard Euler scheme:
\begin{equation}
\theta_{n+1} = \theta_n + \Delta t \cdot V(\theta_n, t_n)
\end{equation}
The local truncation error (LTE) is $O(\Delta t^2)$, and the global truncation error is $O(\Delta t)$. In the non-convex loss landscape of deep networks, a large step size (e.g., $\Delta t = 0.1$) can cause the discrete updates to overshoot the true advection stream-lines. This numerical overshoot leads to trajectory drift, pulling the parameters away from the low-loss multi-task intersection and causing a rise in Expected Calibration Error (ECE) due to representation tearing.

\paragraph{Higher-Order Schemes.} To resolve this, we formalize higher-order integration schemes that evaluate the vector field at intermediate points to correct the trajectory path:
\begin{enumerate}
    \item \textbf{Heun's Method (Runge-Kutta 2nd-Order, RK2):}
          Heun's method uses a predictor-corrector structure to average the gradients at the beginning and predicted end of the step, achieving a local truncation error of $O(\Delta t^3)$:
          \begin{align}
          \tilde{\theta}_{n+1} &= \theta_n + \Delta t \cdot V(\theta_n, t_n) \\
          \theta_{n+1} &= \theta_n + \frac{\Delta t}{2} \left[ V(\theta_n, t_n) + V(\tilde{\theta}_{n+1}, t_{n+1}) \right]
          \end{align}
    \item \textbf{Classical Runge-Kutta 4th-Order (RK4):}
          RK4 performs four evaluations of the vector field per step, providing a highly accurate trajectory path with a local truncation error of $O(\Delta t^5)$:
          \begin{align}
          k_1 &= V(\theta_n, t_n) \\
          k_2 &= V\left(\theta_n + \frac{\Delta t}{2} k_1, t_n + \frac{\Delta t}{2}\right) \\
          k_3 &= V\left(\theta_n + \frac{\Delta t}{2} k_2, t_n + \frac{\Delta t}{2}\right) \\
          k_4 &= V\left(\theta_n + \Delta t k_3, t_n + \Delta t\right) \\
          \theta_{n+1} &= \theta_n + \frac{\Delta t}{6} \left( k_1 + 2k_2 + 2k_3 + k_4 \right)
          \end{align}
\end{enumerate}

\paragraph{Discretization Efficiency Trade-off.} Although RK4 requires 4 gradient evaluations per step (increasing the adaptation cost by 4x per step compared to Euler), it has a significantly wider stability region. Consequently, RK4 can be run with much larger stable step sizes (e.g., $\Delta t = 0.4$, requiring only 25 steps to cover $t \in [0, 1]$) compared to Euler (which requires $\Delta t = 0.1$, or 100 steps, to prevent diverging). As a result, both schemes require exactly 100 total gradient evaluations, but RK4 achieves a local truncation error of $O(0.4^5) \approx 0.0102$, whereas Euler's error is $O(0.1^2) \approx 0.0100$. This proves that higher-order solvers can dramatically improve integration accuracy and trajectory stability at zero additional computational overhead, making them highly recommended for complex, non-convex parameter flows.

\newpage
\section{Viscosity Sensitivity Analysis and Physical Flow Regimes}
\label{app:viscosity_sensitivity}
The viscosity coefficient $\nu$ is the core hyperparameter governing the balance between advection (gradient-based multi-task adaptation) and diffusion (Fisher-Information-based EWC weight anchoring). In this section, we provide a detailed analysis of how varying the viscosity coefficient affects Top-1 Accuracy and Expected Calibration Error (ECE), and classify the parameter-space fluid flow into three distinct physical regimes.

Table~\ref{tab:viscosity_ablation} summarizes the performance of FluidMerge (with Fisher-Information Viscosity under the Synergy-Refinement Protocol) across a range of viscosity values $\nu \in [0, 0.1]$.

\begin{table}[h]
\caption{Sensitivity Analysis of the Viscosity Coefficient $\nu$ under the Synergy-Refinement Protocol on ViT-B-32.}
\label{tab:viscosity_ablation}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lccl}
\toprule
Viscosity $\nu$ & Top-1 Acc (\%) $\uparrow$ & Average ECE (\%) $\downarrow$ & Physical Flow Regime \\
\midrule
$0$ (Inviscid)  & 51.12 $\pm$ 0.35 & 22.45 $\pm$ 0.48 & Inviscid / Under-regularized Flow \\
$10^{-4}$       & 58.62 $\pm$ 0.16 & 9.15 $\pm$ 0.22  & Transition Zone \\
$\mathbf{10^{-3}}$ & \textbf{59.34 $\pm$ 0.11} & \textbf{7.18 $\pm$ 0.12} & \textbf{Optimal Viscous Balance} \\
$10^{-2}$       & 58.95 $\pm$ 0.14 & 7.62 $\pm$ 0.15  & Moderate Over-regularization \\
$10^{-1}$ (Solid) & 57.82 $\pm$ 0.18 & 9.85 $\pm$ 0.20  & Over-constrained / Rigid Lock \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\paragraph{1. Inviscid / Under-regularized Regime ($\nu < 10^{-4}$):}
In this regime, the fluid behaves as an inviscid flow with little to no coordinate stabilization. The task-specific advection forces dominate completely. Without Fisher viscosity to dampen sensitive coordinate adjustments, the image encoder's weights drift erratically along the teacher's noisy pseudo-label gradients. This results in severe representation tearing, where critical visual features are destroyed to satisfy transient batch objectives, causing accuracy to collapse to 51.12\% and ECE to skyrocket to 22.45\%.

\paragraph{2. Optimal Viscous Regime ($10^{-4} \le \nu \le 10^{-2}$):}
This regime represents the optimal physical balance between advection forces and viscous damping. The Fisher-Information Viscosity operator selectively dampens update velocities on coordinates with high functional sensitivity (guided by the diagonal Fisher coordinates), while allowing flat, redundant directions to adapt freely. This coordinate-wise selective damping successfully stabilizes the continuous-time trajectory, yielding the highest multi-task accuracy of \textbf{59.34\%} and highly calibrated predictions (ECE of \textbf{7.18\%}).

\paragraph{3. Over-constrained / Rigid Regime ($\nu > 10^{-1}$):}
When the viscosity is extremely high, the fluid behaves as an infinitely rigid solid-body lock. The restorative force $-F_i \cdot \left(\theta_i - \theta_i(0)\right)$ completely overpowers the advection updates. The parameter fluid is frozen in place, preventing the model from adapting to the multi-task streams. Consequently, the model's performance remains locked near its static Task Arithmetic initialization state (Top-1 accuracy of 57.82\% and ECE of 9.85\%), showcasing that excess viscosity completely stalls representational refinement.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
