\begin{abstract}
Model merging has emerged as a powerful paradigm for combining multiple task-specific expert neural networks into a single multi-task model without additional training. However, standard linear parameter averaging (e.g., Task Arithmetic) suffers from severe task dominance due to representation scale mismatches—where tasks with larger parameter updates or layers with disproportionate magnitudes overshadow others. To resolve this parameter interference, existing methods rely on complex, high-overhead operations such as singular value decomposition (SVD) or active test-time optimization with test-time gradient updates. 

In this paper, we adopt a minimalist perspective and introduce \textbf{Standard-Deviation Scaling (SD-Scale)} and its mathematically stable, non-translation-invariant counterpart, \textbf{Root-Mean-Square Scaling (RMS-Scale)}, for training-free model merging. RMS-Scale balances task representation scales by normalizing each individual task vector layer-wise to unit root-mean-square, establishing a balanced, isotropic update direction while avoiding the instability of standard deviation on small, low-variance tensors like biases. It then projects the merged update back into the network's natural adaptation space by calibrating the average normalized direction with the average original adaptation scale. This method requires absolutely no training, no test-time optimizations, no SVD operations, and is implemented in just two lines of PyTorch code. Under a highly rigorous, multi-seed statistical evaluation on a multi-task image classification benchmark spanning MNIST, FashionMNIST, and KMNIST fine-tuned under uncoordinated downstream schedules, our proposed SD-Scale and RMS-Scale achieve average test accuracies of \textbf{73.23\%} and \textbf{73.22\%} respectively, slightly exceeding standard Task Arithmetic (72.50\%) and Ties-Merging (71.77\%) on average, while matching or exceeding SVD Isotropic Merging (73.13\%) and comfortably outperforming active test-time optimization (AdaMerging gets 62.79\%). Our parameter-free PF-RMS variant achieves \textbf{72.23\%} out-of-the-box without any hyperparameter tuning or disjoint validation data, outperforming un-tuned Task Arithmetic (71.68\%) and un-tuned Ties-Merging (71.81\%). Our findings confirm that robust representation balance in model merging does not require hyper-complex pipelines; rather, simple, minimalist scale calibration is highly stable, efficient, and sufficient.
\end{abstract}
