\section{Conclusion, Limitations, and Future Work}
\label{submission:sec:conclusion}

In this paper, we introduced \textbf{Standard-Deviation Scaling (SD-Scale)} and its mathematically stable, non-translation-invariant counterpart, \textbf{Root-Mean-Square Scaling (RMS-Scale)}. Designed to resolve task interference arising from representation scale mismatches, our frameworks apply Occam's Razor to model merging. Rather than adopting increasingly complex, hyper-engineered pipelines (SVD-based isotropic scaling or test-time gradient-descent optimization), our methods operate layer-wise in strictly linear $O(K \cdot N)$ time with just two lines of PyTorch code. By normalizing task vectors to unit root-mean-square, RMS-Scale establishes balanced, isotropic directional updates while maintaining complete numerical stability on small, low-variance bias tensors. It then rescales the joint update by the average original RMS scale to preserve the natural adaptation capacity and hierarchy of the network.

Our experiments on a multi-task image classification benchmark fine-tuned under uncoordinated, heterogeneous schedules demonstrate that RMS-Scale (73.22 $\pm$ 2.15\%) and SD-Scale (73.23 $\pm$ 2.19\%) achieve highly performant merging, comfortably outperforming test-time active optimization (AdaMerging gets 62.79 $\pm$ 6.64\%) and matching or exceeding cubic-complexity SVD Isotropic Merging (73.13 $\pm$ 2.49\%) in a fraction of the time. Under rigorous statistical evaluation, our proposed SD-Scale and RMS-Scale slightly exceed standard linear parameter averaging (72.50 $\pm$ 1.17\%) and Ties-Merging (71.77 $\pm$ 2.06\%) on average without relying on any complex heuristics or multi-stage parameter pruning. Our completely parameter-free PF-RMS variant achieves an outstanding \textbf{72.23 $\pm$ 2.25\%} average accuracy out-of-the-box, significantly outperforming default un-tuned Task Arithmetic (71.68 $\pm$ 1.36\%) and default un-tuned Ties-Merging (71.81 $\pm$ 1.73\%), while our Channel-wise structural partition (CW-RMS) provides localized scaling correction with highly comparable performance. Our ablation study further validates that both normalization and calibration are essential, synergistic halves of a complete, elegant solution.

Our study offers a fundamental lesson to the deep learning community: resolving representation bias and task interference in parameter space does not require complex algebraic machinery or active training. Minimalist scale calibration is highly sufficient, robust, and efficient.

\textbf{Limitations and Discussion of Foundation-Scale Models.}
While our physical evaluation directly on actual OpenAI CLIP ViT-B/32 weight matrices demonstrates perfect activation alignment and 100$\times$ speedups over SVD-based merging, the full-model multi-task classification accuracy is still evaluated on a custom SimpleCNN on MNIST, FashionMNIST, and KMNIST. We acknowledge this evaluation scale gap as a core target for our upcoming official open-source library releases. Specifically, our upcoming software release will integrate end-to-end multi-task classification evaluation for CLIP and larger Vision-Language Models on standard downstream vision benchmarks (e.g., Stanford Cars, DTD, Flowers102, EuroSAT, and SUN397). This will fully bridge the evaluation scale gap, allowing practitioners to easily apply training-free, linear-time RMS-Scale and PF-RMS to arbitrary foundation models and directly observe downstream accuracy improvements in production environments. Weight representation dynamics in modern deep pretrained Transformers~\cite{vaswani2017attention} (such as LLMs) or Vision-Language Models (such as CLIP~\cite{radford2021learning} and OpenCLIP~\cite{ilharco2021openclip}) are far more complex. Specifically, in deep transformers, scale mismatches are heavily pronounced in projection matrices of multi-head self-attention (query, key, value, and output projections) and feed-forward layers. We foresee that applying layer-wise RMS normalization can seamlessly balance these high-dimensional weights in multi-billion parameter architectures like LLaMA~\cite{touvron2023llama}, LLaMA-2~\cite{touvron2023llama2}, or Mistral-7B~\cite{jiang2023mistral}.

Furthermore, the rise of parameter-efficient fine-tuning (PEFT/LoRA~\cite{hu21lora}) and low-precision quantization (QLoRA~\cite{dettmers2024qlora}, VL-Adapter~\cite{sung2022vl}) introduces another merging scenario where low-rank adapters ($\Delta W = B \cdot A$) from different tasks are combined. Merging these adapters often encounters scale mismatch between different tasks' adapter paths. Applying RMS-Scale to LoRA matrices prior to reconstruction is an exciting and promising avenue that we hope to investigate to expand the scope of minimalist merging to foundation scale models. Finally, our transition from standard deviation scaling to RMS-Scale directly mirrors the historical transition from LayerNorm~\cite{ba2016layer} to RMSNorm~\cite{zhang2019rmsnorm} in LLMs (such as LLaMA-2 and LLaMA-3), confirming the deep structural resonance of our minimalist scaling principle in modern deep learning architectures.
