\begin{abstract}
  Dynamic expert model ensembling is essential for cost-effective multi-task serving on resource-constrained devices, yet state-of-the-art non-parametric methods suffer from a severe early-layer representation trade-off. Specifically, they freeze and leave early layers unadapted to resolve the routing paradox, thereby discarding crucial task-specific features. Conversely, classical parametric routers frequently suffer from catastrophic vectorization collapse when deployed under batch-independent stream regimes (batch size $B=1$). In this paper, we present PEAR (Patch-Embedding Activation Routing), a strictly parameter-free, closed-form routing framework designed for the frozen Patch Embedding layer (Layer 0) of a Vision Transformer. We evaluate PEAR across extensive simulations in a high-fidelity, 12-layer PyTorch representation sandbox that models standard visual tasks (MNIST, Fashion-MNIST, CIFAR-10, SVHN, with SVHN intentionally configured as a highly degraded, high-noise stress-test to evaluate ensembling robustness under compromised task structures). Under standard overlapping task manifolds, PEAR completely eliminates vectorization collapse and restores expert performance with $O(1)$ latency complexity, outperforming SABLE SOTA by +4.04\% on standard overlapping tasks and by +1.74\% under highly optimized expert regimes (96.10\% vs. 94.36\%). Crucially, we also validate PEAR on actual real-world images from these datasets using a pre-trained $\mathtt{vit\_tiny\_patch16\_224}$ backbone. These real-world experiments confirm that while pure Layer 0 routing is limited by a Global-Average-Color Routing Paradox (57.81\% accuracy), shifting the routing boundary slightly deeper to Layer 1 or Layer 2 (the \emph{Early-Layer Routing Compromise}) successfully resolves the paradox. It achieves up to \textbf{95.31\%} real-world routing accuracy, outperforming a trained, lightweight pre-backbone CNN router (\textbf{91.02\%}) while incurring only 20.7\% computational overhead relative to a full backbone forward pass.
\end{abstract}
