\section{Conclusion}
\label{sec:conclusion}

In this work, we presented PEAR (Patch-Embedding Activation Routing), a parameter-free, closed-form ensembling framework that dynamically blends specialized expert adapters at runtime. Grounded in the minimalist philosophy of Occam's razor, PEAR shifts the routing operation to the base model's first structural projection layer---the frozen Patch Embedding layer (Layer 0). By utilizing Layer 0 as a zero-cost, high-fidelity representational space, PEAR completely resolves the Routing Paradox, bypassing the need for redundant backbone forwarding passes or auxiliary gating models. 

Through simple, non-parametric calibration steps including Zero-Shot Patch Centroids (ZPC), Unit-Norm Calibration, and Intra-Task Dispersion Calibration, PEAR achieves state-of-the-art representation selectivity while introducing zero trainable parameters and flat $O(1)$ sequential complexity. Our empirical evaluations confirm that PEAR significantly outperforms SABLE SOTA by +4.04\% on standard overlapping tasks while resolving the early-feature loss trade-off, and outperforms SABLE SOTA by +1.74\% under highly optimized expert regimes by enabling full-depth layer adaptability. PEAR completely eliminates the Vectorization Collapse that plagues low-data parametric routers under batch-independent streaming regimes ($B=1$). Ultimately, PEAR proves that elegant, non-parametric solutions can overcome complex, systems-level bottlenecks while maintaining absolute simplicity. Future work includes scaling PEAR to massive multimodal backbones and exploring robust, non-parametric calibration methods for cross-lingual adapters. Furthermore, we will investigate the application of offline Centered Kernel Alignment (CKA) or Procrustes projection to align early layers under extreme intra-domain semantic overlap (such as fine-grained visual subdomains), which would extend PEAR's minimalist routing to highly fine-grained multi-task serving setups. In text-based language models (where raw pixels are replaced by discrete tokens), PEAR can be adapted by performing average-pooling directly over the input vocabulary token embedding layer (Layer 0) or the output representations of the first Transformer block. Under the Early-Layer Routing Compromise, this maps elegantly to massive, multi-billion parameter architectures: in a standard 32-layer Large Language Model (such as LLaMA or Mistral), routing can be conducted at Layer 2 or Layer 4 to completely bypass vocabulary-level token overlap and capture rich semantic intent. This would allow the framework to establish Zero-Shot Centroids in these early semantic spaces, routing sample-wise based on the semantic distribution of incoming tokens while leaving over 90\% of the network's layers (e.g., blocks 4 to 32) fully adapted and executing specialised LoRA experts with zero parameter or FLOPs overhead.
