\begin{abstract}
Model merging has emerged as a compelling, zero-overhead paradigm for synthesizing multiple task-specific expert neural networks into a single multi-task model without retraining.
However, existing methods are bound to a fundamental dichotomy: they are either static, forcing a rigid compromise across conflicting parameters, or dynamically routed at the batch level, which can introduce transductive dependencies and batch-coupling under mixed-task or non-I.I.D. test-time streams.
In this paper, we propose \textbf{EpiMerge}, a paradigm-shifting model-merging framework inspired by cellular epigenetics.
Just as biological organisms dynamically adapt gene expression in response to environmental stimuli without altering their underlying DNA, EpiMerge uses the input sample itself to dynamically modulate the expression of pre-trained expert parameters.
We introduce highly parameter-efficient \textbf{Epigenetic Reader Heads (ERHs)} that project a global, compact latent representation of each input, generating row-wise and column-wise gating masks via low-rank outer products.
These masks scale the expert task vectors dynamically, reconstructing a unique, sample-specific weight matrix.
Crucially, using vectorized tensor contractions (\texttt{torch.einsum}), EpiMerge executes true sample-wise dynamic merging in parallel across a mixed batch, completely bypassing batch-averaging shortcuts and stream-distribution assumptions.
Across four diverse image classification benchmarks (MNIST, FashionMNIST, CIFAR-10, and SVHN) utilizing a Vision Transformer backbone, EpiMerge demonstrates perfect consistency under non-I.I.D. temporal clustering (Bursty) and extreme small-batch noise ($B=2$) due to its sample-wise independent inference, outperforming uniform model-merging by +20.25\% absolute and exceeding static supervised merging by +22.45\% absolute. We thoroughly characterize the physical trade-offs of this fine-grained gating---including wall-clock latency and peak GPU memory scaling---positioning EpiMerge as a pioneering, biologically-mimetic exploratory proof-of-concept for dynamic multi-task weight modulation.
\end{abstract}
