%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Winner-Take-All Sign Election: A Minimalist Approach to Model Merging}

\begin{document}

\twocolumn[
  \icmltitle{Winner-Take-All Sign Election: \\ A Minimalist Approach to Model Merging}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Jean-Luc Occam}{ias}
  \end{icmlauthorlist}

  \icmlaffiliation{ias}{Institute for Advanced Simplification, Switzerland}

  \icmlcorrespondingauthor{Jean-Luc Occam}{jl.occam@ias.ch}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Machine Learning, ICML}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Confidence-as-Magnitude Foundation}
\label{app:magnitude_confidence}

In this section, we provide additional discussion on the relationship between update magnitude and task confidence in fine-tuned neural networks.
When a pre-trained backbone is adapted to a downstream task via gradient-based optimization (e.g., SGD or AdamW), the weights that undergo the largest absolute changes are those that are being actively reprogrammed to capture the features and nuances of the target domain.
Conversely, weights that undergo near-zero changes are either already suitable for the task or represent dimensions that have minimal relevance to the downstream objective.

When merging multiple experts, parameter-level conflicts (i.e., sign conflicts) represent a state where different tasks demand contradictory representational shifts.
Under the principle of Occam's razor, instead of treating all models equally and executing a convoluted consensus voting loop (which can dilute the most important signals), we argue that the model with the largest absolute update possesses the highest task confidence and necessity at that specific parameter index.
By letting this single maximum-magnitude update dictate the sign of the merged parameter, we guarantee that the most critical domain-specific features are structurally preserved.
Any minor, opposing updates from other tasks are treated as relative noise at that index and are masked out, preventing destructive interference while preserving representation integrity.

\section{Computational and Memory Efficiency Comparison}
\label{app:complexity_analysis}

In this section, we present a detailed comparison of the computational and memory complexity of WTA-Sign versus the leading baseline, \mbox{TIES-Merging} \cite{yadav2023ties}.
Let $K$ represent the number of specialized expert models to merge, and let $D$ denote the total number of parameters in each model.
In modern architectures, $D$ is typically extremely large (ranging from millions of parameters in ViT-B to billions in large language models).

\subsection{TIES-Merging Complexity}
The TIES-Merging pipeline consists of four distinct sequential steps:
\begin{enumerate}
    \item \textbf{Trimming:} To find the top $k\%$ of values by magnitude, the algorithm must sort the absolute values of the parameter updates for each model. Sorting $D$ elements per model takes $O(D \log D)$ time. For $K$ models, this results in a computational complexity of $O(K \cdot D \log D)$.
    \item \textbf{Sign Consensus Voting:} Computing the consensus direction requires collecting sign counts across all models, taking $O(K \cdot D)$ time.
    \item \textbf{Conformity Masking:} Zeroing out non-conforming parameters takes $O(K \cdot D)$ time.
    \item \textbf{Rescaling:} Calculating the energy of the original and trimmed task vectors requires computing norms, which takes $O(K \cdot D)$ time.
\end{enumerate}
Thus, the overall computational complexity of TIES-Merging is dominated by the sorting operation, i.e., $O(K \cdot D \log D)$.
This logarithmic multiplier $\log D$ becomes a major bottleneck as model dimensions scale.

\subsection{WTA-Sign Complexity}
In contrast, our proposed WTA-Sign method relies entirely on standard, element-wise linear operations:
\begin{enumerate}
    \item \textbf{Winner Election:} Finding the argmax over the absolute task vectors takes $O(K \cdot D)$ time, requiring only a single pass over the $K$ models per parameter. No sorting is needed.
    \item \textbf{Sign Election:} Retrieving the sign of the winning expert takes $O(D)$ time.
    \item \textbf{Conformity Masking:} Creating the binary mask based on sign conformity takes $O(K \cdot D)$ time.
    \item \textbf{Conformity Averaging:} Element-wise addition and division of the conforming updates takes $O(K \cdot D)$ time.
\end{enumerate}
The overall computational complexity of WTA-Sign is strictly $O(K \cdot D)$.
By completely eliminating the sorting step, WTA-Sign achieves a massive speedup over TIES-Merging in practice, particularly on high-dimensional model backbones.

\subsection{Memory Comparison}
Both TIES-Merging and WTA-Sign exhibit identical space complexity of $O(K \cdot D)$ to store the task vectors and intermediate mask representations.
However, WTA-Sign does not require storing auxiliary sorted arrays or temporary structures, resulting in a cleaner and more memory-friendly runtime footprint on modern accelerators.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
