\begin{abstract}
Model merging has emerged as a computationally efficient paradigm for multi-task learning, enabling multiple task-specific models to be integrated without retraining on original data. Among adaptive schemes, AdaMerging learns layer-wise merging coefficients in an unsupervised manner using test-time prediction entropy minimization. In this work, we deconstruct AdaMerging to expose two fundamental, interconnected optimization anomalies: (1) the \textbf{Overfitting-Optimizer Paradox}, and (2) the \textbf{Spatial Averaging Paradox}. First, we analyze the role of layer-specific scaling coefficients, demonstrating that while they capture vital architectural hierarchy and hierarchical routing, they are prone to test-time overfitting. We show that these coefficients are structurally specialized to their layers: randomly shuffling them across layers (\emph{Intra-Task Layer Shuffling}) collapses performance. Second, we show that a flat post-hoc \emph{Spatial Average} of these coefficients per task acts as a powerful low-pass regularizer, smoothing away test-time overfitting while preserving the core task-level scales and outperforming static Task Arithmetic, albeit at a $3.09\%$ performance trade-off compared to the unconstrained layer-wise model. Third, we uncover a striking paradox: \emph{direct} test-time optimization of these flat task-wise scales (Task-wise AdaMerging) fails spectacularly, actively degrading performance below its uniform initialization. We explain this **Spatial Averaging Paradox** through multi-task gradient imbalance caused by uncalibrated prediction entropy under low-dimensional bottlenecks: the global optimization is dominated by easy classification tasks, causing destructive parameter interference on harder tasks. Conversely, high-dimensional optimization avoids this trade-off through local layer degrees of freedom, which post-hoc spatial averaging then regularizes. Our work advocates for Occam's razor and scientific precision in model merging, providing a deep, critical analysis of the pitfalls of test-time entropy minimization.
\end{abstract}
