\section{Introduction}
\label{submission:intro}

Multi-task learning (MTL) is a core pursuit of modern artificial intelligence, aiming to construct a single neural network capable of performing multiple diverse tasks simultaneously. While traditional MTL approaches rely on joint fine-tuning or multi-task pre-training, these processes are notoriously resource-intensive, prone to negative transfer, and require simultaneous access to all task-specific datasets. To sidestep these bottlenecks, \emph{model merging} has emerged as an attractive alternative \cite{deep_merge}. By directly combining the weights of independent, task-specific models fine-tuned from a shared pre-trained initialization, model merging achieves multi-task capabilities in seconds of CPU or GPU time without accessing the original training data.

The foundational paradigm of this field is Task Arithmetic \cite{task_arithmetic}, which constructs a merged model by scaling and adding fine-tuned parameter updates (task vectors) to the pre-trained base model. Despite its elegance, naive Task Arithmetic often suffers from performance degradation due to parameter interference and conflicting gradient directions across tasks. To mitigate this, recent literature has introduced increasingly sophisticated merging pipelines. One prominent direction is AdaMerging \cite{AdaMerging_ICLR_2024}, which parameterizes the model-merging process with scaling coefficients (learned either task-wise or layer-wise) and optimizes them in an unsupervised, test-time manner. It uses the average Shannon entropy of the merged model's predictions on a small unlabeled test batch as a surrogate loss function. Under the standard multi-task evaluation protocol of these studies, the shared visual encoder is merged while task-specific classification heads are kept separate. This setup assumes test-time knowledge of the task identity (referred to as \emph{Oracle Routing}) to route inputs to their corresponding heads, isolating the representation quality of the merged backbone.\footnote{We explicitly highlight that this Oracle Routing assumption is standard in the multi-task model merging literature \cite{task_arithmetic, AdaMerging_ICLR_2024, ties_merging}. While it isolates the visual encoder's representation quality from classification head interference, deploying such systems in fully unlabeled environments would require a separate task-identification classifier or a unified multi-task classification head. See Section \ref{sub:preliminaries} for a formal discussion of this boundary condition.}

As these test-time adaptation methods grow in complexity, optimizing hundreds of layer-specific scaling coefficients, we must adopt a skeptical, minimalist lens. Occam's razor suggests that if a complex system can be matched or explained by a simpler one, the simpler system is strictly superior. We raise a critical question: \emph{Does the high-dimensional optimization of layer-wise scaling coefficients capture a fundamental, multi-layer coordination of task representations, or does it merely overfit to the test-time batch?}

In this work, we deconstruct AdaMerging and expose two fundamental, interconnected optimization anomalies:

\textbf{1. The Overfitting-Optimizer Paradox.}
We analyze the role of layer-specific scaling coefficients, demonstrating that while they capture vital architectural hierarchy and representational routing, they are susceptible to test-time overfitting. To deconstruct these scaling dynamics, we design two diagnostic treatments:
\begin{itemize}
    \item \textbf{Intra-Task Layer Shuffling}: We randomly permute the optimized layer-wise coefficients of each task across different layers of the network. We observe a severe accuracy collapse across tasks (from $88.05\%$ down to $78.61\%$). This collapse does not simply imply that the coefficients are random noise; rather, it demonstrates that they are structurally specialized and tailored to the representational hierarchy of the neural network (where early layers capture general features and late layers capture specialized features). Shuffling violates this hierarchy, thereby destroying performance.
    \item \textbf{Spatial Averaging (Spatial Mean)}: We take the optimized layer-wise coefficients for each task and replace them with their flat spatial average, reducing the parameter degrees of freedom from hundreds down to exactly one scalar per task ($T$ parameters total). Crucially, this post-hoc simplification acts as a powerful spatial regularizer. While it incurs a $3.09\%$ performance trade-off compared to the unconstrained layer-wise model ($88.05\%$), it still achieves $84.96\%$, outperforming standard Task Arithmetic ($84.64\%$). This demonstrates that post-hoc spatial averaging acts as a spatial low-pass filter, smoothing away the transductive test-time overfitting component of individual layers while preserving the robust task-level global scales.
\end{itemize}

\textbf{2. The Spatial Averaging Paradox.}
Intriguingly, our deconstruction reveals a striking optimization pathology. While post-hoc \emph{Spatial Averaging} achieves $84.96\%$, \emph{direct} test-time optimization of these flat task-wise scales (Task-wise AdaMerging) fails spectacularly, degrading average accuracy to $81.19\%$---significantly below its uniform initialization of $84.64\%$ (Task Arithmetic). 

We explain this **Spatial Averaging Paradox** through multi-task gradient imbalance caused by uncalibrated prediction entropy under low-dimensional constraints. Prediction entropy is highly sensitive to the scaling of logit magnitudes. Simple classification tasks (such as MNIST or FashionMNIST) have extremely sharp logit distributions; their prediction entropy can be trivially driven to near-zero by scaling up their task vectors, which inflates the logit magnitudes. Harder tasks (such as CIFAR-10 or SVHN) have naturally flatter distributions and higher entropy. 

When optimizing a shared, low-dimensional bottleneck (exactly one scalar per task), the unconstrained joint entropy loss is dominated by easy tasks. The optimizer scales up the easy-task coefficients to drive joint entropy down, creating severe parameter interference that collapses performance on the harder tasks (CIFAR-10 collapses from $89.93\%$ to $81.45\%$; SVHN collapses from $69.94\%$ to $63.71\%$). Under high-dimensional optimization (layer-wise AdaMerging), the optimizer has enough local layer degrees of freedom to minimize entropy without resorting to global scaling trade-offs. Post-hoc spatial averaging then acts as a spatial low-pass filter, preserving the regularized global task scales while smoothing away the high-frequency test-time overfitting.

\subsection{Our Contributions}
Our deconstructive study makes the following contributions:
\begin{enumerate}
    \item We deconstruct the \textbf{Overfitting-Optimizer Paradox} in layer-wise AdaMerging, demonstrating that while the learned coefficients capture vital structural and hierarchical representational routing, they remain prone to test-time overfitting, which post-hoc spatial averaging regularizes.
    \item We discover and mathematically explain the \textbf{Spatial Averaging Paradox} in test-time model merging, detailing how uncalibrated entropy objectives lead to multi-task gradient imbalance and destructive parameter interference on harder tasks under global low-dimensional bottlenecks.
    \item We show that post-hoc \emph{Spatial Averaging} acts as an elegant spatial low-pass filter, resolving this gradient imbalance and achieving $84.96\%$ accuracy, outperforming static Task Arithmetic.
    \item We conduct extensive seed-controlled evaluations of optimization landscape flatness and representational similarity (Linear CKA), clarifying that high CKA is a diagnostic baseline under suppressed task scales rather than an indicator of functional adaptation.
\end{enumerate}

By exposing these optimization paradoxes, our work advocates for a shift in perspective. Rather than blindly building complex, parameter-rich test-time adaptive model merging schemes, we must focus on the inherent math and uncalibrated objectives of weight-space optimization to design robust, scientifically transparent multi-task combinations.
