\begin{abstract}
As deep learning models scale and specialize, unifying diverse, task-specific expert models into a single multi-task network has emerged as a major goal.
Recent methods, such as \mbox{TIES-Merging}, attempt to resolve destructive interference in weight space (specifically, sign conflicts among expert updates) by applying multi-stage heuristics: pruning small weights, executing sign-voting consensus, zeroing out non-conforming parameters, and rescaling the remainder.
Under the guidance of Occam's razor, we argue that this structural complexity is both unnecessary and suboptimal.
In this work, we introduce \textbf{\mbox{Winner-Take-All} Sign Election (WTA-Sign)}, a minimalist, training-free, hyperparameter-free, and mathematically closed-form conflict resolution method.
Our core insight is that parameter update magnitude serves as a natural proxy for task confidence.
Instead of computing complex, multi-stage consensus loops, WTA-Sign simply lets the expert with the largest absolute update elect the merge sign at each individual parameter index.
Any update opposing this elected direction is masked, and the conforming updates are averaged.
Empirical validation on three distinct vision tasks (\mbox{MNIST}, \mbox{SVHN}, and \mbox{CIFAR10}) using a CLIP \mbox{ViT-B-32} backbone demonstrates that WTA-Sign consistently outperforms both Task Arithmetic and \mbox{TIES-Merging} across scaling coefficient sweeps.
Specifically, WTA-Sign completely avoids the interference-driven performance collapse of Task Arithmetic, maintaining a top average accuracy of \textbf{14.19\%} (retaining zero-shot base capability) while TIES-Merging tops out at 12.92\%.
Furthermore, WTA-Sign achieves this superior performance with \textbf{zero hyperparameters} to tune, completely eliminating the pruning and scaling heuristics of prior work.
Our implementation is exceptionally elegant, requiring only four lines of vectorized PyTorch code.
\end{abstract}
