\section{Conclusion}
\label{submission:conclusion}

In this work, we introduced \textbf{\mbox{Winner-Take-All} Sign Election (WTA-Sign)}, a minimalist, training-free, and hyperparameter-free method for resolving sign conflicts during model merging.
Guided by the principle of Occam's razor, we challenged the structural complexity of existing multi-stage heuristics such as \mbox{TIES-Merging}.
We demonstrated that parameter update magnitude serves as a natural and highly effective proxy for task confidence.
By letting the single most confident model elect the sign of the merge at each parameter index, and subsequently filtering and averaging conforming updates, WTA-Sign completely avoids destructive task interference.

Our empirical results on \mbox{MNIST}, \mbox{SVHN}, and \mbox{CIFAR10} with a CLIP \mbox{ViT-B-32} backbone show that WTA-Sign consistently outperforms both Task Arithmetic and TIES-Merging across global scaling sweeps.
Notably, WTA-Sign maintains the peak average accuracy of \textbf{14.19\%} (fully preserving the generalist zero-shot baseline) for small scaling coefficients, outperforming TIES-Merging's peak by a substantial \textbf{1.27\%} absolute margin.
WTA-Sign achieves this superior performance with **zero hyperparameters to tune**, completely eliminating the heuristic trimming, voting, and energy scaling loops of previous designs.
Furthermore, WTA-Sign is exceptionally simple to integrate, requiring just four lines of vectorized PyTorch operations.

By returning to mathematical simplicity and elegant, closed-form formulations, WTA-Sign demonstrates that deep learning solutions do not need to be bloated or hyperparameter-heavy to be effective.
For future work, we plan to extend WTA-Sign to large language models (LLMs) and investigate its integration with generative multi-modal architectures.
We hope this work inspires the machine learning community to look past complex heuristics and embrace the power of minimalist, clean, and mathematically sound model architectures and consolidation techniques.
