\section{Introduction}
\label{sec:intro}

The proliferation of edge computing devices—such as mobile phones, autonomous drones, and localized Internet-of-Things (IoT) sensors—has accelerated the demand for highly adaptive, multi-task on-device vision and language models. When deploying these systems under real-world operational constraints, a major engineering hurdle is the physical storage, memory, and serving footprint of multiple task-specific deep neural networks. Low-Rank Adaptation (LoRA) \cite{hu2021lora} has emerged as an elegant solution, permitting task-specific experts to share a single massive base model backbone, with only tiny low-rank matrices swapped at runtime. 

To bypass the latency of physically swapping adapters from disk to memory, recent literature has proposed test-time model ensembling via parameter merging \cite{adamerging} or single-pass activation-space blending (SPS) \cite{sps_zca}. Frameworks such as SPS-ZCA and SABLE \cite{sps_zca, sable} utilize pre-computed task centroid anchors to route incoming samples to the most relevant expert mixtures on-the-fly. However, from a practical, deployment-first engineering perspective (the \emph{Pragmatist's} view), these state-of-the-art approaches suffer from a critical, unaddressed bottleneck: \textbf{they depend entirely on offline, manual, and labeled calibration datasets}. For example, SPS-ZCA requires $|\mathcal{C}_k|=64$ annotated calibration samples per task to calculate task centroids before the model is deployed. 

In real-world deployment, this labeled calibration requirement is frequently a dealbreaker. Consider a security camera, a medical wearable, or an on-device industrial sensor. These systems process continuous, non-stationary data streams where data privacy is absolute (GDPR or HIPAA prohibit data upload), manual labels are unavailable, and zero-downtime plug-and-play operation is mandatory. 

To achieve true on-device autonomy, we need a routing and merging framework that is completely unsupervised and calibration-free. In this work, we present a comprehensive, technically rigorous, and honest systems-ML comparative study of \textbf{Zero-Shot Calibration-Free Model Merging}. We formulate, evaluate, and compare two primary architectural paradigms:
\begin{itemize}
    \item \textbf{Zero-Shot Expert Entropy Routing (EER) [Accuracy-First Paradigm]:} We bypass representation-space centroids and instead route test samples directly using the specialized expert models' own predictions. We evaluate the scale-invariant Normalized Shannon prediction entropy of all $K$ expert adapters on the incoming sample, and route $100\%$ of the compute to the expert $k^*$ exhibiting the minimum entropy (maximum confidence). Evaluated on our synthetic 192-dimensional multi-task sandbox (modeling the noise properties of MNIST, FashionMNIST, CIFAR-10, and SVHN using Gaussian distributions), EER achieves an outstanding \textbf{71.38\%} Joint Mean accuracy under heterogeneous shuffled streams, outperforming the supervised SOTA baseline (SPS-ZCA, 66.76\%) by \textbf{+4.62\%} absolute accuracy. Under extreme continuous representation drift ($d=0.45$), EER maintains complete robustness, delivering \textbf{71.18\%} joint accuracy entirely calibration-free.
    \item \textbf{Entropy-Pseudo-Labeled Online Centroid Adaptation (EPL-OCA) [Efficiency-First Paradigm]:} To enable single-pass serving, we maintain running task centroids on-the-fly. We pseudo-label incoming samples with EER to determine the active expert index $k^*$, and update the task centroid $\mu_{k^*}$ using a running average. Subsequent steps are served in a single pass via SPS blending based on similarity coordinates over the centroids. While conceptually elegant, when evaluated with a chronologically correct, non-leaking centroid update, we find that EPL-OCA achieves a Joint Mean accuracy of \textbf{49.88\%} (and \textbf{49.78\%} under drift) in our synthetic sandbox due to the \textbf{Representational Sparsity Paradox} (class orthogonality within task subspaces introduces severe sparseness, causing centroids to jitter between class prototypes rather than converging on task manifolds). Crucially, we reveal that on real-world ResNet-18 embeddings, EPL-OCA experiences a \textbf{complete methodological collapse}, yielding only \textbf{27.45 ± 1.34\%} (Hard) and \textbf{31.52 ± 1.37\%} (Soft), failing to outperform simple static \textbf{Uniform Weight Merging} (\textbf{31.66 ± 0.91\%}). This collapse is driven by the Entropy Calibration Discrepancy, which corrupts online running centroids via a self-referential pseudo-label feedback loop, proving that fully unsupervised online centroid adaptation is fundamentally non-functional on uncalibrated real embeddings without offline spatial anchors.
\end{itemize}

A critical systems-level analysis is conducted on the physical serving overhead on edge hardware. Because LoRA adapters are inserted post-Layer 3, the forward activations of different experts diverge immediately at Layer 4. To calculate the prediction entropy of the $K$ experts, the heavy, shared backbone base weights for Layers 4 to 12 must be executed independently for each expert, resulting in a minimum serving complexity of $0.25 + 0.75 K$ forward passes. We propose \textbf{Amortized Pseudo-Labeling} as a practical systems-level mitigation to bring serving latency back to $1.57\times$ passes ($0.2211$ ms per sample) and slash edge energy footprint by $4.14\times$ while preserving accuracy ($71.20\%$).

Furthermore, we conduct extensive validation on real-world 512-dimensional ResNet-18 embeddings. Under real feature spaces, we identify the \textbf{Entropy Calibration Discrepancy}, where simpler experts exhibit severe out-of-distribution (OOD) overconfidence (e.g., MNIST expert claiming CIFAR-10 and SVHN samples), causing direct entropy routing (EER) to drop to \textbf{35.38\%} accuracy. To resolve this, we propose and evaluate \textbf{Centroid-Gated Entropy Routing (CG-EER)}, achieving \textbf{61.50\%} accuracy (surpassing supervised SPS-ZCA by \textbf{+0.70\%}). We explicitly and prominently classify CG-EER as a hybrid semi-supervised framework rather than a purely calibration-free one (unlike our primary EER and EPL-OCA paradigms, which are entirely zero-shot and calibration-free), since it relies on pre-computed offline task centroids to restrict entropy routing to candidate experts near the sample. We prominently state this classification upfront to ensure academic transparency regarding its partial dependency on offline calibration data ($|\mathcal{C}_k|=64$), preventing any misleading overclaims of complete calibration-freedom. Crucially, we highlight that while our zero-shot methods perform exceptionally well on structured manifolds, the most viable and stable real-world solution under real representation shifts remains this hybrid semi-supervised approach (CG-EER), demonstrating that uncalibrated real-world representation manifolds require spatial anchors to manage overconfidence and break the self-referential pseudo-label corruption loop. To address the question of whether these spatial anchors can be obtained in an entirely unsupervised manner, we evaluate Unsupervised Centroid-Gated Entropy Routing (\textbf{UCG-EER}) across a wide range of warm-up window sizes $T_{\text{warmup}} \in [10, 50, 100, 200]$. Across all window lengths, we reveal a persistent \emph{self-referential pseudo-label corruption loop} that causes accuracy to collapse (yielding $28.36 \pm 1.25\%$ for $T_{\text{warmup}}=10$ and $28.45 \pm 1.59\%$ for $T_{\text{warmup}}=200$), which is significantly below even Uniform Merging ($31.66\%$). This outcome mathematically and empirically motivates the hybrid semi-supervised design of CG-EER, proving that fully unsupervised gating fails on uncalibrated real-world manifolds.

In summary, our core contributions are:
\begin{enumerate}
    \item We present the first study of completely calibration-free, zero-shot dynamic model ensembling, eliminating the offline labeled data bottleneck of test-time merging.
    \item We propose and evaluate EER inside our synthetic multi-task simulation sandbox, achieving 71.38\% Joint Mean accuracy, surpassing supervised ZCA by +4.62\% and surviving linear covariate shift completely calibration-free.
    \item We identify and fix the chronological data leakage bottleneck in online centroid updates, providing an honest, non-leaking empirical evaluation of EPL-OCA.
    \item We validate our methods on real 512-dimensional ResNet-18 embeddings, formulating the \emph{Entropy Calibration Discrepancy} and proposing the hybrid, semi-supervised \emph{Centroid-Gated Entropy Routing (CG-EER)} to resolve OOD overconfidence.
    \item We evaluate a fully unsupervised centroid-gated routing variant (\emph{UCG-EER}) and uncover a \emph{self-referential pseudo-label corruption loop}, providing deep physical insights into the limits of pure unsupervised online adaptation.
    \item We mathematically formulate the systems-ML FLOP serving complexity under post-Layer-3 activation divergence, detailing practical edge-deployment trade-offs and profiling CPU latencies.
\end{enumerate}
