\begin{abstract}
Model merging has emerged as a powerful paradigm for consolidating multiple specialized, task-specific expert models into a single multi-task neural network without the prohibitive cost of joint retraining. Existing approaches primarily rely on static, linear weight interpolation or on simple, parameter-heavy feed-forward dynamic routers. In this work, we present \textbf{ChaosMerge} (Chaos-Theoretic Attractor Merging), a novel dynamic model-merging framework that conceptualizes the sequence of a network's layers as discrete time-steps in a non-linear, chaotic Coupled Map Lattice (CML) driven by a Logistic Map. 
While traditional deep chaotic recurrent lattices suffer from catastrophic gradient explosion by design, we resolve this fundamental flaw by introducing a \textbf{Gated Coupled Map Lattice (G-CML)} that utilizes learned layer-wise gating as skip-connections, stabilizing the optimization landscape. Furthermore, we address the common batch-averaging contradiction of dynamic merging by employing \textbf{Task-Specific Dynamic Routing} which executes via task-level centroid features. This preserves task-specific characteristics from being washed out while avoiding sample-by-sample weight-assembly latency at inference time.
Empirical evaluations across visual classification tasks (MNIST, FashionMNIST, CIFAR-10, SVHN) using a Vision Transformer backbone demonstrate that ChaosMerge with only 384 parameters significantly stabilizes optimization, outperforming competitive static merging baselines (such as Task Arithmetic and AdaMerging) and achieving highly competitive performance compared to over-parameterized dynamic routers with nearly $30\times$ more parameters. Our study bridges the gap between non-linear dynamical systems and parameter-space operations, opening new avenues for efficient, regularized dynamic multi-task fusion.
\end{abstract}
