\begin{abstract}
Weight-space model merging has emerged as a powerful, training-free paradigm for creating multi-task networks from fine-tuned specialized experts. However, existing dynamic model merging methods (e.g., QWS-Merge, Linear Routers) suffer from a critical deployment bottleneck: \emph{batch-dependency} and \emph{heterogeneity collapse}. Because these methods compute a single set of merged weights by averaging routing coefficients across the batch dimension, their predictions violate the standard I.I.D. assumption—causing a sample's classification to shift depending on other samples in the batch—and collapse under heterogeneous, mixed-task inference streams. 
To resolve this, we present \textbf{Sparse Low-Rank Dynamic Merging (SLD-Merge)}, a completely batch-independent, parameter-efficient dynamic weight merging framework. SLD-Merge decouples sample-level inference by utilizing offline Singular Value Decomposition (SVD) on expert task vectors to construct lightweight low-rank adapters. At inference time, a bounded cosine-similarity router selects and activates only the single most relevant task adapter per sample (Top-1 hard gating). This sparse parallel design reduces task-specific parameter storage overhead by over \textbf{92.5\%} and adds only \textbf{8.3\%} computational overhead (FLOPs). Furthermore, we introduce \emph{Activation-Space Mean Initialization}, which enables high-quality zero-shot routing without any gradient descent calibration. Evaluating across MNIST, FashionMNIST, CIFAR-10, and SVHN streams under varying batch sizes ($B \in \{1, \dots, 256\}$), SLD-Merge maintains a stable, peak joint accuracy of \textbf{63.87\%}, outperforming strong dynamic baselines by up to \textbf{+8.50\%} while completely eliminating heterogeneity collapse.
\end{abstract}
