\section{Conclusion and Future Work}
\label{sec:conclusion}

In this work, we characterized and addressed the critical limitation of \emph{batch-dependency} and \emph{heterogeneity collapse} in dynamic weight-space model merging. We demonstrated that existing dynamic methods, which average routing coefficients across the batch dimension, fail under heterogeneous streaming inputs and introduce undesirable prediction variations based on batch packaging. 

To resolve these challenges, we proposed \textbf{Sparse Low-Rank Dynamic Merging (SLD-Merge)}, a batch-independent and parameter-efficient dynamic model-merging framework designed for robust real-world deployment. SLD-Merge employs offline SVD on specialized expert task vectors to produce compact, high-fidelity low-rank adapters. During inference, a bounded cosine-similarity router utilizes Top-1 hard gating to route each sample completely independently through only its most relevant adapter path. 

Furthermore, we introduced \textbf{Activation-Space Mean Initialization}, demonstrating that highly competitive dynamic routing can be achieved completely zero-shot, eliminating the need for expensive, hyperparameter-sensitive backpropagation calibration. 

Our empirical results across MNIST, FashionMNIST, CIFAR-10, and SVHN streams confirm that SLD-Merge maintains a stable and peak joint accuracy of \textbf{63.87\%} across all batch sizes, outperforming strong dynamic baselines by up to \textbf{+8.50\%} while reducing the extra task-specific parameter footprint by over \textbf{92.5\%} and adding only \textbf{8.3\%} computational overhead. 

For the Pragmatist, SLD-Merge offers an exceptionally robust, simple-to-integrate, and resource-efficient solution for deploying consolidated multi-task experts in high-frequency online production environments. In future work, we plan to extend SLD-Merge to large language models (LLMs) and profile its execution speed on physical edge NPU and microcontroller hardware.
