\begin{abstract}
  In multi-task learning, combining task-specific Parameter-Efficient Fine-Tuning (PEFT) experts via model merging offers a powerful way to serve diverse requests without model duplication. However, deploying these merged models on heterogeneous multi-task inference streams presents a severe challenge: standard dynamic routing collapses back to uniform compromises because batch-averaging washes out individual task signals, while prior state-of-the-art methods like Parameter-Free Subspace Routing (PFSR) combined with Micro-Batch Homogenization (MBH) rely on heavy serving-level data partitioning, sequential execution dispatch, and index re-sorting. This shifts complexity from the model to the data-orchestration infrastructure and results in a sequential execution latency bottleneck scaling linearly with task diversity. To resolve this complexity dilemma, we present \textbf{Parameter-Free Activation Blending (PFAB)}, a minimalist, non-parametric framework that performs sample-wise activation-space blending of lightweight expert adapters in a single, parallelized forward pass of the backbone. Equipped with Unit-Norm Calibration to address scale imbalance, PFAB completely prunes serving-layer systems bloat and achieves flat, constant wall-clock latency. Evaluated on the high-fidelity Isolating Coordinate Sandbox, our mathematically precise two-pass pathway (PFAB-BOP) matches the prior state-of-the-art (PFSR+MBH) perfectly at 81.50\% Joint Mean accuracy under heterogeneous streams (both hitting the absolute expert ceiling) but does so with zero dynamic batch-partitioning or sequential dispatching complexity, delivering a 2.52$\times$ latency speedup (5.84 ms vs. 14.72 ms) at four active tasks ($B=64$). Concurrently, our single-pass pathway (PFAB-ELC) introduces zero trainable parameters but relies on pre-computed offline task centroids, resolving requests in constant time with a 3.26$\times$ speedup (4.52 ms vs. 14.72 ms), presenting a major victory for Occam's razor in systems-ML design.
\end{abstract}
