\section{Related Work}
\label{sec:related_work}

We position our work at the intersection of model merging, test-time routing, and systems-ML co-design, highlighting how a minimalist architectural shift resolves challenges that previously required heavy database-level orchestrations.

\subsection{Static Model Merging in Parameter Space}
Model merging is an emerging paradigm to synthesize the capabilities of multiple fine-tuned models into a single set of weights without the prohibitive cost of multi-task retraining. Early approaches like Model Soups \cite{wortsman2022robust} and Federated Averaging (\textsc{FedAvg}) \cite{mcmahan2017communication, wang2020federated} demonstrated that averaging weights from models with identical initializations can improve robustness and multi-task generalizability. However, naively averaging models specialized on distinct tasks causes severe destructive interference due to sign disagreements and magnitude disparities in parameter deltas. To combat this, Task Arithmetic \cite{ilharco2022editing} utilizes task-specific vectors to guide weight updates. \textsc{TIES}-Merging \cite{yadav2023ties} introduces a three-step pipeline—pruning redundant near-zero parameters, resolving sign disagreements via majority voting, and scaling the resulting merged vector—to alleviate parameter interference. Similarly, \textsc{DARE} \cite{yu2023language} uses randomized pruning and scaling to achieve super-additive capabilities. Despite these advances, static merging techniques remain bound to a single, global compromise, which inevitably degrades the peak specialized performance of each individual expert.

\subsection{Dynamic Test-Time Routing and Mixture of Experts}
To overcome the limitations of static compromises, recent research has explored dynamic, test-time model merging, Mixture of Experts (\textsc{MoE}) architectures \cite{shazeer2017outrageously, fedus2021switch}, and Low-Rank Mixture of Experts (\textsc{LoRA}-\textsc{MoE}) frameworks \cite{pfeiffer2020adapterhub}. Multi-adapter serving and LoRA-MoE models typically activate and scale multiple parallel adapters concurrently at each layer to handle multi-task requests. Learnable parametric frameworks, such as AdaMerging \cite{yang2023adamerging}, optimize weight-blending coefficients at test-time using entropy minimization. While these methods are effective on clean homogeneous inputs, they require gradient calculations or backward passes, which are slow and unstable in real-time deployment. 

Alternatively, feed-forward parametric routers learn to predict task-routing coefficients directly from representation activations. However, these methods suffer from two severe limitations. First, they require labeled calibration splits, which are often unavailable at test-time, and easily overfit to transductive noise. Second, and more critically, because these routers output a single set of merged model weights for the entire batch to perform inference, they suffer from \emph{heterogeneity collapse} when deployed on mixed-task streams. Under heterogeneous batching, batch-average pooling forces the routing coefficients to flat uniform averages, causing the model to collapse back to a weak static baseline and catastrophically degrade performance across all tasks. 

While recent parameter-efficient Mixture of Experts and synthesis frameworks such as \textsc{LoRA-MoE} \cite{loramoe2023} and \textsc{LoraHub} \cite{lorahub2023} attempt to alleviate multi-task interference via sample-level routing and modular adapter synthesis, they rely on learnable parametric gating networks or multi-stage optimization pipelines that require extensive training data or active few-shot calibration. Conceptually, evaluating parallel LoRA adapters and taking their weighted sum at the activation level is structurally identical to the forward execution pass of multi-LoRA serving and Low-Rank Mixture-of-Experts (LoRA-MoE) serving frameworks, such as S-LoRA \cite{s_lora2024} or Punica \cite{punica2023}. However, while standard LoRA-MoE and multi-LoRA serving frameworks focus on optimizing systems execution with learned parametric gating networks or complex data-orchestration layers, our proposed \textsc{PFAB} focuses on a fundamentally different, minimalist objective. The primary distinction lies in the gating strategy: rather than training complex routing heads or relying on extensive calibration splits, \textsc{PFAB} serves as a non-parametric, calibration-free alternative to learnable LoRA-MoE gating. By shifting dynamic synthesis from parameter-space to activation-space and utilizing frozen classification heads to derive gating coordinates directly, \textsc{PFAB} achieves sample-level routing with zero trainable parameters and zero calibration data, while retaining the high hardware arithmetic intensity of standard single-model deployment.

\subsection{Systems-Serving Complexity and Micro-Batch Homogenization}
To address the vulnerability of dynamic merging to mixed-task streams, researchers have turned to systems-level mitigations. Our predecessor, Parameter-Free Subspace Routing (\textsc{PFSR}) coupled with Micro-Batch Homogenization (\textsc{MBH}) \cite{subspace2026}, resolved heterogeneity collapse by shifting the problem of robustness to the data-serving infrastructure. \textsc{MBH} dynamically partitions the heterogeneous stream into homogeneous micro-batches, compiles separate merged models on-the-fly, dispatches them sequentially, and re-sorts the outputs via scatter-gather operations. While systems-level multi-tenant serving engines like Punica \cite{punica2023} and \textsc{SGMV} \cite{sgmv2023} attempt to optimize multi-LoRA serving through customized CUDA kernels, they require complex systems dependencies, compilation pipelines, and heavy runtime scheduling overhead. 

In contrast, our proposed \textsc{PFAB} represents a radical return to mathematical simplicity. By shifting the blending operation from parameter-space (which is batch-bound) to activation-space (which is sample-bound), we completely prune the database partitioning and sequential dispatching infrastructure. This achieves state-of-the-art multi-task capabilities in a single forward pass of the backbone with constant latency and zero trainable parameters, delivering the ultimate minimalist systems-ML co-design.
