\begin{abstract}
  Model merging has emerged as a highly practical, zero-overhead paradigm to consolidate multiple task-specific expert models into a single unified architecture without expensive multi-task retraining. 
  To adapt blending coefficients dynamically at deployment, Test-Time Adaptation (TTA) via entropy minimization has become a prominent approach. 
  However, in real-world physical environments, test-time input streams are frequently corrupted by sensor noise, weather artifacts, and compression distortions. 
  Under these realistic corruptions, standard adaptive model merging suffers from the \emph{Overfitting-Optimizer Paradox} and is prone to \emph{Noise-Entropy Collapse}, where the first-order optimizer overfits high-frequency transductive noise, causing catastrophic out-of-distribution performance degradation. 
  In this paper, we propose \textbf{FlatMerge}, a highly robust and memory-efficient test-time model merging framework. 
  FlatMerge introduces a dual-regularization mechanism: first, it constrains the optimization search space to a low-degree polynomial of normalized depth to filter high-frequency parameter noise; second, it optimizes the remaining parameters using flatness-aware minimization in the coefficient space to find a flat entropy valley, preventing low-frequency transductive drift. 
  Recognizing the computational and dataset limits of edge deployment, we evaluate FlatMerge inside a highly calibrated numerical simulation of Vision Transformer (ViT-B/32) weight merging, and rigorously validate its physical behavior on real neural networks (MLP and 5-layer CNN experts fine-tuned on MNIST, FashionMNIST, and KMNIST). 
  Our experiments demonstrate that FlatMerge successfully prevents transductive collapse, achieving state-of-the-art robust accuracies (e.g., $85.59\% \pm 0.63\%$ under moderate noise in simulation, and outperforming standard first-order AdaMerging by over $30\%$ absolute on real CNN weights). 
  Importantly, because coefficient-space TTA is formulated using zeroth-order randomized smoothing, ZO-FlatMerge requires zero weight-space backpropagation and zero activation memory caching during adaptation. This resolves critical memory bottlenecks for resource-constrained edge accelerators by trading off a manageable latency overhead.
\end{abstract}
