\begin{abstract}
Test-time dynamic model merging offers a highly flexible, parameter-efficient paradigm for serving multiple specialized expert adapters under heterogeneous query streams. To smooth ensembling weight trajectories and suppress sample-level feature noise, stateful recurrent routing has recently emerged as a powerful technique. However, existing stateful routers rely on an unrealistic deployment assumption: that incoming queries form a continuous, highly correlated single-user stream. In actual production systems, such as large-scale LLM gateways or edge servers, workloads are multi-tenant and heavily interleaved across independent users. In this work, we expose and analyze the \emph{state contamination (cross-talk) bottleneck}, where standard stateful routers bleed memory states across unrelated tenants, causing catastrophic representation bleeding and serving accuracy drops. To resolve this major deployment blocker, we introduce \textbf{Tenant-Decoupled Stateful Routing (TDSR)} as an overarching framework powered by our underlying \textbf{Slot-Kinetics} state decoupling mechanism, which maintains a decoupled pool of active virtual routing states. We propose two highly pragmatic modes: (i) \emph{Explicit Tagging} (metadata available) and (ii) \emph{Implicit Tagless Clustering} (dynamically inferred session context via online cosine similarity of activation coordinates). Under the high-fidelity Analytical Coordinate Sandbox (ICS) evaluated with realistic non-conflated multi-tenant streams, TDSR completely isolates routing states across users. We demonstrate that TDSR Explicit successfully resolves state contamination, achieving up to \textbf{70.60\%} classification accuracy on Orthogonal Manifolds (outperforming contaminated Global PAC-Kinetics by \textbf{+1.90\%} absolute) and \textbf{70.85\%} on Overlapping Manifolds (outperforming contaminated Global PAC-Kinetics by \textbf{+1.75\%} absolute and performing within 0.50\% of the isolated clean-stream baseline). Crucially, TDSR Explicit stabilizes the ensembling trajectories, slashing intra-session routing jitter by up to \textbf{2.4$\times$} relative to stateless SABLE.
\end{abstract}