\section{Conclusion and Future Work}
\label{sec:conclusion}

In this work, we introduced \textbf{Tenant-Decoupled Stateful Routing (TDSR)} as an overarching framework powered by our underlying \textbf{Slot-Kinetics} state decoupling mechanism, offering a highly practical, robust, and zero-overhead solution for stateful test-time model ensembling. While previous stateful ensembling routers provide remarkable representational stability by smoothing routing weight trajectories, they suffer from a severe \emph{state contamination bottleneck} under realistic, multi-tenant interleaved workloads, where memory states are contaminated across unrelated users' requests.

TDSR resolves this bottleneck by maintaining a decoupled pool of active virtual routing states, successfully isolating temporal smoothing within respective user sessions. We proposed both \emph{Explicit Tagging} and \emph{Implicit Tagless Clustering} (dynamically inferred session context via online cosine similarity of activation coordinates, stabilized with fixed orthogonal task-detector centroids).

Through comprehensive experiments inside the high-fidelity Analytical Coordinate Sandbox (ICS), we demonstrated that TDSR Explicit completely eliminates cross-tenant state contamination, achieving robust and stable ensembling tracking. Under rapidly interleaved i.i.d. streams, TDSR Explicit achieves up to \textbf{70.60\%} classification accuracy on Orthogonal Manifolds and \textbf{70.85\%} on Overlapping Manifolds, representing a significant \textbf{+1.90\%} and \textbf{+1.75\%} absolute accuracy improvement over the contaminated Global PAC-Kinetics baseline respectively, completely closing the gap to the strong Oracle clean-stream ceiling. Furthermore, our proposed intra-session jitter analysis demonstrates that TDSR Explicit provides near-Oracle routing stability, slashing high-frequency activation-level noise and reducing intra-session jitter by up to \textbf{2.4$\times$} relative to stateless SABLE.

From a systems-engineering perspective, TDSR is exceptionally well-suited for high-throughput production serving infrastructures. Maintaining small arrays of active states (e.g., $M \times K = 16$ floats) takes virtually zero GPU/CPU memory and permits in-register operations that complete in sub-microsecond scales. Furthermore, by introducing a tenant-specific logical session-step decay, TDSR achieves a scalable, self-cleaning session memory pool, preventing memory leaks and state washout under high-concurrency workloads.

\paragraph{Future Work.}
An exciting avenue of future work is the integration of Slot-Kinetics within hardware-aware active continuous batching engines such as S-LoRA or Punica. Specifically, the stateful routing tensors can be tracked as compact sequence-level metadata alongside the Key-Value (KV) cache. Another promising direction is extending TDSR to support dynamic slot spawning and reclamation, enabling the router to automatically adapt to a highly volatile, time-varying number of active users in real-time edge gateways.