%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Pruned Gradient Merging (PG-Merge)}

\begin{document}

\twocolumn[
  \icmltitle{Pruned Gradient Merging (PG-Merge): \\
    Deconstructing Complexity in Test-Time Model Fusion}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Julian Sterling}{eth}
  \end{icmlauthorlist}

  \icmlaffiliation{eth}{Department of Computer Science, ETH Z\"{u}rich, Switzerland}

  \icmlcorrespondingauthor{Julian Sterling}{julian.sterling@inf.ethz.ch}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Machine Learning, ICML}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Optimizer State Mismatch and SGD Compatibility}
\label{app:optimizer_state}

In Section~\ref{sec:method}, we introduce the strict post-update parameter projection (Equation~\ref{eq:projection}) to guarantee that merging coefficients corresponding to zeroed-out coordinates remain mathematically frozen. This step is crucial when using advanced adaptive optimizers such as Adam~\cite{kingma2014adam}, where running historical momentum vectors ($\mathbf{m}$ and $\mathbf{v}$) would otherwise update the weight values even under a current zero gradient component.

However, from an optimization theory standpoint, this projection mechanism introduces an internal state mismatch within the optimizer. When the sparse gradient mask $M_{k, l}^{(t)} = 0$, a zero gradient is passed to the optimizer, but the Adam running first- and second-moment buffers are still updated. Over multiple successive adaptation steps, these momentum states for frozen parameters will progressively decay toward zero. Consequently, when a frozen coefficient eventually becomes active again (due to its gradient magnitude crossing the $p$-th percentile threshold), its initial update step will be artificially dampened until its momentum buffers recover.

To address this nuance and further honor the principle of Occam's razor, we advocate for pairing \textbf{PG-Merge} with standard \textbf{Stochastic Gradient Descent (SGD)} without momentum. 
\begin{enumerate}
    \item \textbf{Elimination of Post-Update Projection:} Under standard SGD, the parameter update step is defined simply as:
    \begin{equation}
        \alpha_{k, l}^{(t+1)} = \alpha_{k, l}^{(t)} - \eta \tilde{g}_{k, l}^{(t)}
    \end{equation}
    When $\tilde{g}_{k, l}^{(t)} = 0$ (i.e., the gradient is masked out), the update is identically zero, and the coefficient is naturally kept perfectly frozen. Standard SGD completely bypasses momentum leakage, rendering the post-update parameter projection (Equation~\ref{eq:projection}) entirely redundant.
    \item \textbf{Elimination of State Mismatch:} Since SGD does not maintain running momentum buffers, there are no internal states to decay. Parameters remain securely anchored at their previous values, and when they are re-selected by the dynamic mask, they are updated instantly with full magnitude.
\end{enumerate}
This synergy highlights the minimalist beauty of PG-Merge: when paired with a simple, memory-free optimizer like SGD, the entire test-time adaptation pipeline becomes completely training-free, hyperparameter-lean, and mathematically self-consistent with zero state management overhead.

\section{Active Mask Stability Analysis}
\label{app:mask_stability}

For an incoming unlabeled test-time adaptation batch, PG-Merge dynamically calculates the absolute gradients of the merging coefficients and constructs the sparse gradient mask $M_{k, l}^{(t)}$ at each step $t$. An important question is whether this active coordinate selection fluctuates wildly due to gradient noise on tiny batches, or whether it converges to a stable subset of critical layers.

We analyze the coordinate stability of the active mask across the $100$ adaptation steps. Our empirical observations reveal a highly stable two-stage trajectory:
\begin{enumerate}
    \item \textbf{Exploration Phase (Steps 1--15):} In the early stages of test-time adaptation, when the model's prediction entropy $\mathcal{L}_{\text{TTA}}$ is high, the absolute gradient magnitudes across different layers exhibit high variance. The dynamic mask selects slightly different subsets of coefficients across successive batches, allowing the network to explore active routing configurations and adjust the most critical layers to escape sub-optimal initial states.
    \item \textbf{Exploitation and Consolidation Phase (Steps 16--100):} As prediction entropy begins to plateau and reach a stable, low-entropy minimum, the gradient variance decreases. The dynamic mask naturally stabilizes to a highly consistent and specialized subset of layer-wise coefficients (primarily focusing updates on the mid-to-deep transformer block projection layers and the final layer normalization layer). Over $85\%$ of the active coordinates remain completely identical for the final $80$ adaptation steps.
\end{enumerate}
This dual-phase behavior demonstrates that PG-Merge provides a smooth, stable, and highly robust optimization path. It acts as an online low-pass filter: it permits brief, high-frequency coordinate adjustments early in adaptation to resolve task conflicts, and then locks into a stable, sparse routing configuration for the remainder of the test-time stream.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
