\begin{abstract}
Dynamic model merging serves as a highly scalable paradigm for executing multi-task sequential workloads, enabling on-the-fly ensembling of low-rank adapters (LoRAs) without the excessive storage footprint of multiple full-parameter models. State-of-the-art stateful routing methods smooth the ensembling trajectory by modeling routing as a continuous-time chemical kinetics process. However, existing frameworks assume spatial homogeneity across network layers, applying a single global routing coefficient across all network depths. This overlooks a fundamental truth: different layers of deep neural networks extract features at different semantic scales, and therefore process representation spaces at distinct temporal tempos. 

In this work, we introduce \textbf{Layer-Decoupled Stateful Kinetics (LDS-Kinetics)}, a learning-theoretic stateful model merging framework that decouples stateful ensembling dynamics across layers. By partitioning layers into distinct depth-decoupled blocks (or individual layers), LDS-Kinetics maintains separate, independent concentration states that evolve according to block-specific parameters. We address the resulting hazard of overparameterization and transductive overfitting on short calibration streams by formulating a unified PAC-Bayesian complexity penalty across all blocks simultaneously. 

Adhering to a strictly empirical philosophy, we evaluate LDS-Kinetics across extensive multi-dimensional parameter sweeps over 5 independent random seeds inside a 14-layer analytical coordinate sandbox. Across both orthogonal and overlapping task manifold environments, and under both homogeneous and highly heterogeneous query streams, LDS-Kinetics matches or exceeds existing stateful routers. Crucially, our exhaustive ablation sweeps deconstruct the temporal-spatial ensembling dynamics for the first time, demonstrating that deeper layers learn high inertia to act as stable low-pass decision filters, while early layers adapt rapidly to capture transient input transitions.
\end{abstract}
