\begin{abstract}
Model merging has emerged as a powerful paradigm for combining multiple task-specific expert neural networks into a single multi-task model without expensive retraining. Recently, ``quantum-inspired'' techniques, such as Quantum Wavefunction Superposition Merging (QWS-Merge), have claimed state-of-the-art results by modeling merging coefficients as wave-like phase interference in parameter Hilbert space. In this paper, we critically examine these claims from a methodological perspective. We deconstruct QWS-Merge within a controlled, multi-task representation sandbox designed to mathematically isolate and analyze the dynamics of parameter routing. 

Our empirical findings expose a severe vulnerability: under rigorous sandbox validation, the highly complex wave-like formulation of QWS-Merge completely collapses, achieving a Joint Mean accuracy of only \textbf{36.10\%}—performing worse than static uniform merging (\textbf{43.40\%}). In contrast, we show that a simple \textbf{Layer-wise Low-dimensional Classical Router (L3-Router)} avoids this collapse, with our L3-Linear variant achieving a Joint Mean of \textbf{63.10\%} (\textbf{+27.00\%} over QWS-Merge). Most remarkably, our evaluation reveals that the simplest baseline of all—a global, unregularized classical \textbf{Linear Router}—outperforms all multi-layer models, achieving the highest Joint Mean of \textbf{67.20\%}. This serves as a profound methodological warning: researchers frequently introduce architectural complexity or quantum analogies before rigorously evaluating simple global baselines. 

Finally, we audit deployment streams, demonstrating that heterogeneous mixed-task batches cause severe ``heterogeneity collapse'' in linear and quantum routers. We critically deconstruct our own proposed \textbf{L3-Softmax} variant under these mixed-task streams, exposing the **``Robustness-Accuracy Illusion''**. Although L3-Softmax exhibits relative stability under stream shifts (dropping by only \textbf{4.10\%} compared to the Linear Router's \textbf{16.10\%} drop), its absolute accuracy is consistently inferior in all scenarios, showing that its relative robustness is an artifact of Softmax constraints forcing dynamic routing coefficients toward a mediocre, uniform-like average. Our work advocates for extreme baseline transparency and rigorous experimental hygiene in model-merging research.
\end{abstract}
