%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Task-Correlation Prior Regularization (TCPR)}

\begin{document}

\twocolumn[
  \icmltitle{Task-Correlation Prior Regularization (TCPR):\\ Guiding Calibration via Representation Cosine Similarity}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.

  \begin{icmlauthorlist}
    \icmlauthor{Evelyn Sterling}{stanford}
  \end{icmlauthorlist}

  \icmlaffiliation{stanford}{Department of Computer Science, Stanford University, Stanford, CA, USA}

  \icmlcorrespondingauthor{Evelyn Sterling}{esterling@stanford.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Model Merging, Dynamic Routing, Calibration, Prior Regularization}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Experimental Details and Hyperparameters}
\label{appendix:details}

To support scientific transparency, reproducibility, and rigorous empirical validation, we provide the complete architectural specifications and optimization hyperparameters used throughout our experiments.

\subsection{Architectural Specifications}
Our backbone is a Vision Transformer ($\mathtt{vit\_tiny\_patch16\_224}$) with $12$ Transformer blocks, a patch size of $16 \times 16$, and an input resolution of $224 \times 224$ pixels. The hidden dimension is $D = 192$, and the model utilizes $3$ attention heads per layer, totaling approximately $5.7$ million parameters. 

The routing head consists of a single linear projection layer:
\begin{equation}
    W_{\text{route}} \in \mathbb{R}^{192 \times 4}, \quad b_{\text{route}} \in \mathbb{R}^4
\end{equation}
The total number of trainable routing head parameters is exactly $772$. The routing head maps the average patch representation $z(x) \in \mathbb{R}^{192}$ to the routing logits $o(x) \in \mathbb{R}^4$.

\subsection{Task Expert Specialized Training}
Each expert model is fine-tuned from the pre-trained ImageNet checkpoint under identical convergence-guaranteed conditions. The task-specific training details are summarized below:
\begin{itemize}
    \item \textbf{Data Subsets:} We extract a class-balanced subset of exactly $1000$ images per task from each dataset (MNIST, FashionMNIST, CIFAR-10, SVHN).
    \item \textbf{Optimization:} We utilize the AdamW optimizer with a learning rate of $2 \times 10^{-4}$, weight decay of $1 \times 10^{-4}$, and a batch size of $32$.
    \item \textbf{Epochs:} Each expert is trained for exactly $2$ epochs to ensure specialization convergence while preventing parameter divergence from the base progenitor.
\end{itemize}

\subsection{Routing Head Calibration Phase}
The dynamic routing head is optimized on a low-data calibration split to evaluate the methods under severe resource constraints:
\begin{itemize}
    \item \textbf{Calibration Dataset:} Exactly $16$ class-balanced images per task, resulting in a total calibration dataset size of $64$ images.
    \item \textbf{Optimization Budget:} The routing head parameters ($W_{\text{route}}$, $b_{\text{route}}$) are optimized using the Adam optimizer for exactly $100$ steps.
    \item \textbf{Learning Rate and Batch Size:} The learning rate is fixed at $10^{-2}$ with a batch size of $64$ (incorporating the entire calibration set in each step to stabilize gradient estimates).
    \item \textbf{Regularization Coefficients:} The isotropic L2 weight decay factor $\gamma$ is fixed at $10^{-4}$. For the TCPR prior, we swept $\beta \in [10^{-6}, 10^2]$ on a logarithmic scale, identifying $\beta = 1.0$ as optimal for TCPR-Param and $\beta = 0.01$ as optimal for TCPR-Rep.
\end{itemize}

\subsection{Task Similarity Pre-computation}
The task similarity matrices $S^{\text{param}}$ and $S^{\text{rep}}$ are pre-computed offline prior to calibration:
\begin{itemize}
    \item For $S^{\text{param}}$, we calculate the flat cosine similarity of the task vectors $V_k^{(l)}$ at each of the $12$ Transformer blocks and average them.
    \item For $S^{\text{rep}}$, we pass the $64$ calibration samples through the base model and extract the global representation vectors $z(x) \in \mathbb{R}^{192}$ to compute the cross-task similarity.
\end{itemize}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
