\begin{abstract}
The post-hoc merging of task-specific expert neural networks into a unified multi-task model has emerged as a major paradigm for resource-efficient deployment. However, existing ensembling strategies depend on a fundamental, unchallenged additive assumption ($W_{\text{merged}} = W_{\text{base}} + \sum_k \alpha_k V_k$) that leads to parameter coordinate conflicts and destructive interference. Furthermore, we expose a critical vulnerability in current dynamic routing methodologies: under streaming heterogeneous workloads, standard hardware constraints average ensembling coefficients across the batch, leading to a catastrophic \textit{heterogeneity collapse} that flattens expert specialization. In this work, we propose \textbf{Endosymbiotic Holographic Parameter Binding} (EHPB), a novel model merging paradigm drawing from hyperdimensional computing and holographic representations. EHPB treats weight space as a holographic associative memory. Task-specific parameter offsets (task vectors) are bound onto mutually orthogonal high-dimensional bipolar carrier keys and superimposed into a unified holographic weight matrix. At test-time, an input-dependent dynamic unbinding operator demodulates the active expert weights on a sample-by-sample basis, bypassing batch-level ensembling constraints. Empirically, EHPB is completely immune to heterogeneity collapse, retaining its task specialization perfectly. Finally, through rigorous dimension scaling sweeps, we deconstruct a fundamental mathematical constraint of element-wise Hadamard holographic parameter binding, showing that its relative reconstruction error remains scale-invariant. We trace this boundary to the coordinate-isolated nature of Hadamard multiplication versus central-limit-averaging operators like circular convolution, providing an essential theoretical roadmap for future hyperdimensional weight-space ensembling.
\end{abstract}
