\section{Related Work}
\label{sec:related}

\textbf{Model Merging and Soups.} Fusing independent fine-tuned models post-hoc creates multi-task learners without joint retraining. Early approaches like Model Soups~\cite{wortsman2022modelsoups} average weights to improve generalization. Task Arithmetic~\cite{ilharco2022editing} additively combines task vectors $V_k = W_k - W_{\text{base}}$. Git Re-Basin~\cite{ainsworth2022git} aligns permuted network coordinates before averaging, while Fisher Merging~\cite{matena2021merging} weights parameters by diagonal Fisher information. TIES-Merging~\cite{yadav2023ties} prunes small parameter changes and resolves sign conflicts. However, these static merging approaches rely on a uniform linear projection that results in destructive coordinate-wise interference, failing to capture fine-grained task specialization.

\textbf{Dynamic Routing and Test-Time Adaptation.} To mitigate parameter conflicts, dynamic routing can adjust ensembling coefficients $\alpha_k(x)$ based on the input $x$ at test-time. Mixture-of-Experts (MoE)~\cite{jacobs1991adaptive, shazeer2017outrageously, fedus2021switch} routes inputs to specialized sub-networks but requires expensive end-to-end training. Recent post-hoc model merging approaches learn lightweight routing networks at test-time. However, under standard batch-parallel hardware constraints, standard frameworks average ensembling coefficients across the batch dimension. This causes a major, previously undocumented vulnerability: when a batch is heterogeneous (containing samples from multiple tasks), the averaged coefficients flatten expert specialization, causing \textit{heterogeneity collapse}.

\textbf{Hyperdimensional Computing and Holographic Representations.} Hyperdimensional Computing (HDC) or Vector Symbolic Architectures (VSA)~\cite{kanerva2009hyperdimensional} operate on high-dimensional random vectors (typically $D \geq 10^4$) using algebraic operators: addition (superposition), multiplication (binding), and permutation. Holographic Reduced Representations (HRR)~\cite{plate2003holographic} utilize circular convolution as the binding operator to associate features into a distributed vector. While HDC is widely applied to cognitive modeling and associative memories, its application to deep neural network parameter ensembling remains largely unexplored. EHPB bridges this gap by extending hyperdimensional binding from 1D feature vectors to 2D neural network parameter matrices, treating weight space as a holographic associative memory.

Parameter-space modulation and associative superposition have rich roots in older works, notably in \textbf{fast weights} mechanisms~\cite{von1981correlation, ba2016using}, where weights undergo rapid, temporary updates to store associative memory patterns. Similarly, \textbf{hypernetworks}~\cite{ha2016hypernetworks} employ auxiliary networks to generate weight matrices dynamically, and classical \textbf{associative memory networks}~\cite{hopfield1982neural, kohonen1972correlation} store multiple superimposed patterns in a single matrix using correlation-based outer products. EHPB builds upon these foundational paradigms, extending the concept of correlation-based parameter modulation to post-hoc model merging on edge devices, where multi-task experts are superimposed and transcribed dynamically under strict resource constraints.

\textbf{Deconstruction of Over-engineered Metaphors.} Recently, some model merging techniques claim state-of-the-art performance by invoking complex, physical-world metaphors. A prominent example is QWS-Merge~\cite{chosetme2024qws}, which models weights as wave packets in Hilbert space and defines ensembling as quantum phase-interference. As deconstructed by Ashaf \& Jordan~\cite{ashaf2024overengineered}, these complex formulations introduce mathematical instabilities and fail to beat simpler, properly regularized classical baselines. Wave phase-interference equations exhibit erratic gradient behavior during optimization and suffer from coordinate misalignment. Following the visionary researcher philosophy, we reject these over-engineered metaphors in favor of a mathematically transparent, hyperdimensionally sound parameter binding framework.
