\section{Related Work}
\label{sec:related_work}

\paragraph{Parameter-Space Model Merging.}
Model merging combines task-specific expert models training-free \cite{ilharco2022editing, wortsman23}. \emph{Task Arithmetic} \cite{ilharco2022editing} averages weight differences directly but suffers from interference. To resolve this, \emph{TIES-Merging} \cite{ties2023} prunes small updates and elects parameter signs, while \emph{DARE} \cite{dare2024} drops and rescales updates. Other methods dynamically optimize coefficients at test-time, including \emph{AdaMerging} \cite{adamerging2024} and \emph{SyMerge} \cite{symerge2026}. Though successful, these operate in flat Euclidean space, leaving them vulnerable to representation drift since they ignore non-linear network geometry.

\paragraph{Geometry-Preserving and Manifold Learning.}
Restricting parameters to manifolds stabilizes training \cite{harandi2016riemannian}. \emph{Orthogonal Fine-Tuning (OFT)} \cite{oft2023} constrains updates to the orthogonal group $\mathrm{O}(d)$ to guarantee feature stability and prevent forgetting. \emph{OrthoMerge} \cite{yang2026orthomerge} merges orthogonal updates on the manifold of $\mathrm{O}(d)$ by solving the Orthogonal Procrustes problem \cite{procrustes2004} and performing interpolation in the tangent Lie algebra $\mathfrak{so}(d)$ via Cayley transforms. We build upon this geometric framework, subjecting it to a rigorous spectral analysis on curved manifolds.

\paragraph{Spectral Analysis and Representational Isotropy.}
Weight spectra are deeply connected to generalization \cite{pennington2017geometry}; anisotropic updates dominated by a few singular vectors cause representation collapse. \emph{Sharpness-Aware Isotropic Merging (SAIM)} \cite{saim2026} uses sharpness-aware optimization to seek flat minima and performs adaptive isotropic merging via SVD. In Euclidean space, balancing the spectrum to restore isotropy significantly enhances multi-task performance. However, SAIM is restricted to linear weight additions. We present the first attempt to translate representational isotropy to curved manifolds, demonstrating the non-linear distortions that arise under tangent-space spectral modifications.
