\begin{abstract}
Test-time ensembling of multiple expert adapters (e.g., LoRA) offers a powerful paradigm for serving multi-tenant workloads under sequential, streaming queries. However, existing stateless dynamic routers suffer from a severe \emph{routing jitter paradox}, where representation noise causes high-frequency ensembling weight oscillations across layers. Conversely, prior stateful routers, such as those simulating continuous-time biochemical kinetics (ChemMerge) or simple exponential moving averages (Momentum-Merge), smooth these trajectories but introduce severe \emph{inertial drag} (phase/group delay) during task switches, leading to dramatic accuracy collapses under heterogeneous streams. In this work, we propose \textbf{PID-Merge}, a closed-loop, discrete-time Proportional-Integral-Derivative (PID) controlled stateful routing framework. Rooted in classical control theory, PID-Merge treats raw similarity routing weights as the setpoint and active ensembling coefficients as the plant output. By leveraging Proportional responsiveness, Integral smoothing, and Derivative anticipation, PID-Merge achieves exceptional smoothness. Crucially, the Derivative (D) term measures tracking error acceleration, allowing the system to anticipate task transitions and completely suppress phase lag. Our empirical evaluations on the Isolating Coordinate Sandbox (ICS) demonstrate that calibrated PID-Merge achieves up to \textbf{94.82\%} accuracy on challenging heterogeneous overlapping streams, outperforming ChemMerge by \textbf{+6.40\%} and Momentum-Merge by \textbf{+8.65\%} absolute accuracy. Furthermore, its robust training-free (zero-shot) mode achieves \textbf{93.35\%} accuracy under extreme noise and manifold overlap, delivering a highly practical, computationally lightweight ($O(1)$ overhead), and deployment-ready solution for real-world resource-constrained serving environments.
\end{abstract}
