\section{Related Work}
\label{sec:related_work}

Our proposed method lies at the intersection of parameter-efficient adapter serving, model merging, stateful ensembling, and process control theory.

\paragraph{PEFT Serving and Model Merging.}
Low-Rank Adaptation (LoRA) \cite{hu2021lora} has become the standard for tailoring LLMs. SOTA serving engines like Punica \cite{chen2023punica} and S-LoRA \cite{s-lora} hot-swap adapters on a shared base model but assume hard task routing. To blend multiple experts without extra inference costs, static merging methods such as Model Soups \cite{wortsman22soups}, TIES-Merging \cite{yadav2023ties}, and DARE \cite{dare} combine parameters before deployment. However, these techniques yield a single, static model, and are fundamentally unable to adapt to sequential workloads or dynamically adjust blending weights in real-time.

\paragraph{Stateless Dynamic Routing.}
Test-time ensembling blends expert outputs on-the-fly. Stateless routers, such as SABLE \cite{sable} and SPS-ZCA \cite{spszca}, measure activation similarities to pre-computed centroids and route inputs via soft ensembling weights. However, due to representation noise, stateless routing coefficients oscillate wildly from layer to layer—a phenomenon we term the \emph{routing jitter paradox}—which corrupts representations and degrades serving accuracy.

\paragraph{Stateful Dynamic Ensembling.}
To mitigate routing jitter, ChemMerge \cite{chemmerge} models weights via ODE kinetics, which Momentum-Merge \cite{momentum_merge} simplifies as open-loop EMA. PAC-Kinetics \cite{pac_kinetics} optimizes stateful parameters via PAC-Bayes bounds. However, these open-loop stateful methods suffer from \emph{inertial drag}: under rapid task switches, history accumulation delays adaptation, mis-routing queries and collapsing performance, while ODE integrations introduce prohibitive latency overheads.

\paragraph{Control Theory in Deep Learning.}
PID control has been used to schedule learning rates \cite{an2018pidlr}, stabilize GAN training \cite{he2019ganpid}, and formulate stable Neural ODEs \cite{chen2018neuralode}. However, it has not yet been applied to dynamic serving. PID-Merge bridges this gap, replacing fragile chemical metaphors with a classical closed-loop control algorithm that guarantees rapid convergence, simplicity, and high systems-level robustness.