\section{Conclusion, Limitations, and Future Work}
\label{sec:conclusion}

In this work, we introduced \textbf{PID-Merge}, a closed-loop Proportional-Integral-Derivative (PID) controlled stateful routing framework for dynamic, test-time model ensembling. Rooted in classical process control theory, PID-Merge treats raw early-layer similarity weights as the reference setpoint and active layer-wise ensembling coefficients as the controlled plant output. Feeding back the tracking error utilizing Proportional, Integral, and Derivative gains, PID-Merge filters out representation noise while maintaining outstanding responsiveness. Our evaluations on the Isolating Coordinate Sandbox (ICS) show that calibrated PID-Merge achieves up to \textbf{94.82\%} accuracy under overlapping manifolds—yielding up to \textbf{+6.40\%} absolute improvement over SOTA ChemMerge and \textbf{+8.65\%} over Momentum-Merge—while slashing routing jitter and running with negligible latency.

\subsection{Limitations and Honest Scoping}
Despite its strong empirical and theoretical contributions, we transparently note three key limitations of this work:
\begin{itemize}
    \item \textbf{Open-Loop on Neural Representations:} While PID-Merge operates as a closed-loop controller in the ensembling weight trajectory space, it is fundamentally open-loop with respect to the underlying neural representation space. It tracks target weights anchored at early layers but lacks real-time measurement of intermediate representation drift or activation corruption at deeper layers. A fully representation-closed system remains computationally prohibitive for dynamic serving.
    \item \textbf{Sandbox Simulation Noise Limits:} The simulated ICS sandbox injects representation noise only at the initial boundary layer. Consequently, stateless SABLE transitions instantly and appears artificially stable across depth. Physical deep networks generate independent noise at every layer, which is why our physical hardware experiments (using GPT-2) are essential to demonstrate PID-Merge's true layer-wise low-pass filtering.
    \item \textbf{Scale and Stability Enforcement:} Our physical validations are currently limited to a 12-layer GPT-2 backbone. In multi-billion parameter backbones (e.g., 32-layer LLaMA-3), deep-scale phenomena like integrator windup and transition lag are more pronounced, necessitating our proposed conditional integration clamping. Furthermore, while we derive analytical stability bounds via Jury's criterion and enforce them using a soft penalty during calibration, integrating active hard-projection stability boundaries into non-linear parameter search is an ongoing challenge.
\end{itemize}

\subsection{Future Directions}
An exciting immediate avenue for future work is evaluating PID-Merge on multi-billion parameter backbones (e.g., LLaMA-3 8B) with larger, crowded expert pools. Furthermore, we envision developing an online, self-tuning PID controller that monitors rolling query autocorrelation in real-time to dynamically scale gains on-the-fly under shifting workloads, as outlined in our theoretical framework.
