\section{Conclusion}
\label{submission:sec:conclusion}

In this paper, we introduced \textbf{Norm-Preserved Budgeted Task-Vector Pruning (NP-BTVP)}, a training-free weight sparsification and merging framework for specialized expert models. Motivated by a pragmatist philosophy focused on edge-deployment constraints, we presented and evaluated norm-preserved deterministic pruning schemes: global Uniform Pruning (NP-BTVP-U) and layer-wise Adaptive Saliency-Based Pruning (NP-BTVP-S). Crucially, we incorporated norm-preserving rescaling into our pruning masks to mathematically prevent update norm shrinkage and preserve update signal strength.

Empirically, our framework achieves up to \textbf{95\% sparsity (compressing a 115 MB CLIP expert to under 23 MB)} with outstanding multi-task merging performance. Our study uncovers three major insights of high value to the model merging community:
First, we find that training-stage loss landscape flatness (via SAM) does not provide an additional coordinate-aligned pruning buffer under well-converged regimes compared to standard AdamW. Rather, when coupled with our norm-preserving rescaling, both standard AdamW and flatness-aware SAM experts demonstrate an extraordinary and nearly identical level of resilience to heavy sparsification, maintaining high classification performance even under extreme 95\% sparsification. This separates the geometry of dense weight merging from coordinate sparsification, highlighting norm-preserving rescaling as the primary driver behind recovering sparsification performance.
Second, under identical storage budgets, our deterministic rescaled Uniform Pruning (NP-BTVP-U) is highly competitive with DARE-Merging and completely crushes TIES-Merging (by \textbf{3.81\%} at half the parameter budget under SAM), demonstrating that deterministic magnitude selection is highly robust and acts as a powerful regularizer while avoiding stochastic dropout variance.
Third, Saliency Pruning (NP-BTVP-S) is subject to a fundamental double-bind between inter-layer scale imbalance (under global scaling) and local noise amplification (under layer-wise scaling), making the simpler, global NP-BTVP-U the most stable, robust, and pragmatically superior edge-merging solution.

Future work will explore combining our rescaled pruning methods with ultra-low bitwidth quantization (e.g., 4-bit or 2-bit quantization of the active updates) to achieve extreme 40--80$\times$ storage reduction, as well as developing on-device hardware accelerators that can fuse sparse task vectors into dense backbones with zero latency on edge hardware. Additionally, while we evaluated NP-BTVP on vision transformers, the framework is highly relevant to Large Language Models (LLMs) such as LLaMA or OPT. Due to the extreme parameter scale and layer-wise heterogeneity of LLMs, we expect the Saliency Double-Bind to be even more pronounced in language tasks, further cementing global Uniform Pruning (NP-BTVP-U) as the optimal, most stable choice for a broader audience of NLP practitioners. By shedding light on the geometric and scale-preservation properties of weight-space model merging, our work provides a solid, practical foundation for highly compressed, cost-efficient, and robust edge intelligence.