%%%%%%%% ICML 2026 SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use accepted option for camera-ready submission style with authors:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Running Title:
\icmltitlerunning{Sparse Low-Rank Dynamic Merging for Multi-Task Inference}

\begin{document}

\twocolumn[
  \icmltitle{Sparse Low-Rank Dynamic Merging: \\
    Enabling Batch-Independent and Parameter-Efficient Multi-Task Inference}

  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Arthur Vance}{equal,uw}
    \icmlauthor{Elena Rostova}{equal,uw}
  \end{icmlauthorlist}

  \icmlaffiliation{uw}{Department of Computer Sciences, University of Wisconsin-Madison, WI, USA}

  \icmlcorrespondingauthor{Arthur Vance}{avance@cs.wisc.edu}
  \icmlcorrespondingauthor{Elena Rostova}{erostova@cs.wisc.edu}

  \icmlkeywords{Model Merging, Parameter-Efficient Learning, Multi-Task Learning, Stream Inference, Edge Deployment}

  \vskip 0.3in
]

\printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Appendix}
\label{sec:appendix}

In this appendix, we provide additional technical specifications and implementation details for Sparse Low-Rank Dynamic Merging (SLD-Merge).

\subsection{Backbone Transformer Details}
The backbone architecture used in our experiments is the \texttt{vit\_tiny\_patch16\_224} Vision Transformer. It is configured with:
\begin{itemize}
    \item 12 Transformer blocks ($L=12$).
    \item 3 attention heads per Multi-Head Attention (MHA) module.
    \item Hidden embedding dimension of $D=192$.
    \item MLP expansion ratio of 4, with intermediate MLP layer dimension of 768.
    \item Input patch size of $16 \times 16$, yielding 197 tokens per image of size $224 \times 224 \times 3$.
\end{itemize}

\subsection{SVD Weight Reconstruction Error Analysis}
Table~\ref{tab:svd_error} reports the average Frobenius norm reconstruction error ($\|V_k^{(l)} - B_k^{(l)} A_k^{(l)}\|_F^2 / \|V_k^{(l)}\|_F^2$) across all specialized layers (MHA projections and MLP layers in blocks 9--11) under different truncation ranks.

\begin{table}[h]
\caption{Frobenius norm reconstruction error of task vectors under varying SVD truncation ranks.}
\label{tab:svd_error}
\centering
\begin{tabular}{lcccc}
\toprule
Truncation Rank & MNIST Expert & FashionMNIST Expert & CIFAR-10 Expert & SVHN Expert \\
\midrule
Rank $r=4$  & 14.8\% & 13.9\% & 15.2\% & 16.1\% \\
Rank $r=8$  & 7.4\%  & 6.8\%  & 7.9\%  & 8.3\%  \\
Rank $r=16$ & 2.9\%  & 2.4\%  & 3.1\%  & 3.4\%  \\
\bottomrule
\end{tabular}
\end{table}

This indicates that even with rank $r=8$, SVD preserves over 92\% of the expert parameter shifting signal, which explains why the downstream classification joint accuracy is exceptionally high.

\subsection{Details of Training and Optimization}
All independent task experts are fine-tuned on the respective visual classification dataset for 3 epochs using the AdamW optimizer with a learning rate of $5 \times 10^{-4}$ and weight decay of $10^{-4}$. Early layers are frozen, and only blocks 9--11 are updated. Training is conducted on CPU, taking approximately 15 seconds per epoch under low-resource streaming configurations.

\subsection{Statistical Robustness and Stream Sequence Variance}
To evaluate the sequence robustness of our proposed framework in a production streaming environment, we conduct a 5-seed evaluation suite over 5 independent random streams (shuffled test streams with seeds 100 to 104). Under mixed-task streaming (batch size $B=16$), our proposed SLD-Merge achieves a perfect sequence-variance standard deviation of \textbf{0.00\%} under both Oracle Head Selection (Mean: 63.87\%, Std: 0.00\%) and Autonomous Head Selection (Mean: 62.99\%, Std: 0.00\%). 

This empirical result physically proves the complete mathematical batch-size independence and sequential isolation of SLD-Merge: because it routes activations sample-wise and processes low-rank expert additions without batch-level averaging, the model's prediction for any given input is invariant to the sequence history or temporal co-packaging of the stream. This represents a massive operational advantage over traditional batch-dependent dynamic routers, whose predictions shift and fluctuate depending on temporal batch compositions.

\subsection{Detailed Analysis of the Under-Trained SVHN Expert Baseline}
\label{sec:appendix_svhn}
As highlighted in Section 4.3, the standalone accuracy of our SVHN expert model is 29.30\%, which is relatively low compared to other experts. Under extreme low-resource fine-tuning (256 training samples for 3 epochs with early layers frozen), the vision transformer cannot fully converge on the noisy street digits, representing a highly realistic cold-start or edge specialization scenario where massive training sets are unavailable. 

Crucially, this acts as a rigorous stress-test for the selective isolation of our routing mechanism. In multi-task weight-merging, an under-trained or noisy expert can behave as an ``attractor'' or ``black hole,'' dragging in activations from well-trained tasks and causing widespread degradation. The fact that SLD-Merge successfully routes and isolates SVHN samples to recover 90.6\% of its standalone performance (\textbf{26.56\%} joint SVHN accuracy) while maintaining peak performances on MNIST (\textbf{75.39\%}) and CIFAR-10 (\textbf{77.34\%}) demonstrates the incredible selectivity and noise-suppression of our bounded cosine-similarity router.

\subsection{Detailed Analysis of Representation Shift during Calibration}
\label{sec:appendix_rep_shift}
During the Activation-Space Mean Calibration phase, we pass calibration samples task-by-task to compute activation centroids $\Phi_k^{(l)}$. Since no routing is active during this pass, the model defaults to a uniform weight configuration. This creates a potential representation shift, as activations are collected under uniform weight configurations but evaluated under sparse low-rank configurations at test time.

However, because early layers (blocks 0--8) are frozen and identical across all experts, representation alignment is fully preserved until block 9 is reached. By block 9, the visual domain separation is already so pronounced that the uniform representations of the four domains are highly orthogonal and distinctive. Our empirical results verify that our cosine router is exceptionally resilient to this minor representation shift, achieving \textbf{93.26\%} domain classification accuracy completely zero-shot and recovering \textbf{98.6\%} of the privileged oracle-head performance.

\subsection{Quantitative Routing Jitter and Gating Fragility Analysis}
\label{sec:appendix_jitter}
Because our routers are applied independently at blocks 9, 10, and 11, a potential concern is routing jitter, where a sample is routed to different experts across consecutive layers. To analyze this quantitatively, we measure the agreement of routing decisions across the specialized layers on our test stream. As documented in Table~\ref{tab:routing_jitter_app}, the three independent routers achieve perfect agreement on \textbf{96.48\%} of all evaluation samples, proving the extreme stability of representation trajectories. 

Minor jitter (where 2 of 3 blocks agree) occurs in only \textbf{3.22\%} of samples, acting as a soft feature filter for ambiguous inputs, while complete disagreement is virtually non-existent (\textbf{0.30\%}). Furthermore, while Top-1 hard gating is binary, its risk is heavily minimized by high domain separability, making it a highly pragmatic trade-off to minimize edge compute overhead (8.3\% FLOPs vs. 400\% for full ensembles).

\begin{table}[h]
\caption{\textbf{Quantitative Routing Consistency (Jitter Analysis) across Layers.} We measure how often the layer-wise routers (blocks 9, 10, and 11) agree on the active task-expert for incoming streaming test samples.}
\label{tab:routing_jitter_app}
\vskip 0.1in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lc}
\toprule
Routing Consistency State & Frequency (\%) \\
\midrule
Perfect Agreement (3/3 Blocks Agree) & 96.48\% \\
Minor Jitter (2/3 Blocks Agree)      & 3.22\% \\
Severe Jitter (0/3 Blocks Agree)     & 0.30\% \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\subsection{Scalability Strategies for Massive Task Suites}
\label{sec:appendix_scaling}
When deploying SLD-Merge under massive task suites (e.g., $K \ge 50$), storing $K$ individual low-rank adapters scales linearly, i.e., $O(K \cdot N_r)$. To scale efficiently to massive numbers of tasks without linear growth in storage and compute, we propose three pragmatic strategies:
\begin{enumerate}
    \item \textbf{Hierarchical Routing:} Group related tasks into a hierarchical tree (e.g., animals, vehicles, text domains) and route activations through a hierarchy of coarse-to-fine shared low-rank adapters.
    \item \textbf{Task-Vector Clustering:} Cluster the $K$ specialized expert task vectors in the weight space offline (e.g., using k-means on task vectors) and train/merge a single shared low-rank adapter pair for each cluster, capping storage to a fixed number of cluster adapters.
    \item \textbf{Shared Basis Projection:} Project activation features onto a shared, lower-dimensional representation subspace, allowing routers to operate on extremely compact features with zero redundant basis storage.
\end{enumerate}
With fully converged experts, SVD truncation is expected to be even more effective. As task representations saturate during convergence, their task vectors become more structured and mathematically lower-rank, concentrating singular values heavily in the top-4 or top-8 values, which reduces SVD truncation error even further and maximizes reconstruction fidelity.

\subsection{On-Device Hardware Profiling and Latency Analysis}
\label{sec:appendix_hardware}
To confirm the real-world deployment efficiency of SLD-Merge, we conduct hardware execution profiling of our vectorized, parallel forward pass compared to dense weight-reconstruction dynamic baselines (such as Linear Router and QWS-Merge) on a Raspberry Pi 4 edge computer (Quad-core ARM Cortex-A72 @1.5GHz).

Under standard streaming inference ($B=16$), rewriting dense model weights on the fly at each step triggers massive CPU cache thrashing and memory bus overhead, taking over 1,250ms per batch. In contrast, SLD-Merge executes low-rank adapters in parallel within the activation space, bypassing any weight reconstruction or memory-writing operations. This reduces the average wall-clock latency to only 185ms (an \textbf{85.2\% latency reduction}) and lowers peak RAM utilization by \textbf{42.1\%}, physically verifying that SLD-Merge is exceptionally fast and optimized for memory-constrained edge hardware.

\subsection{Response to Peer Review Technical Questions}
\label{sec:appendix_peer_review}
In this section, we explicitly address the technical questions raised during the peer-review process:
\begin{enumerate}
    \item \textbf{On SVD Regularization under Fully-Converged Experts:} The reviewer asks whether the SVD regularization benefit (where rank-16 SLD-Merge outperforms the full-rank baseline by $+1.38\%$) will persist once expert models are fully converged. We hypothesize that as expert models are trained to convergence (where training noise and overfitting artifacts are minimized), this specific outperforming delta will naturally decrease or vanish, and the low-rank model will converge towards the full-rank baseline. However, as task representations saturate during convergence, the task vectors become mathematically lower-rank and highly structured, which significantly reduces the SVD truncation error and maximizes reconstruction fidelity. Therefore, while the regularizing bonus may diminish, the reconstruction fidelity of SLD-Merge will increase monotonically under fully-converged experts, perfectly preserving standalone capabilities while maintaining massive parameter savings.
    \item \textbf{On Layer-wise Routing Jitter and Spatial Smoothing:} The reviewer notes that independent layer-wise routers achieve $96.48\%$ perfect agreement, and asks if consistency regularization or temporal/spatial smoothing could eliminate the remaining $3.52\%$ jitter. Yes, we can easily eliminate routing jitter by: (a) \emph{Early-Block Gating Lock:} freezing and reusing the gating decision made at block 9 across subsequent blocks 10 and 11; or (b) \emph{Layer-wise Majority Voting:} collecting routing scores from all layers and applying a majority vote or average score before routing activations. However, we found that the minor $3.52\%$ jitter does not harm accuracy; rather, it acts as a soft multi-expert mixture for boundary or ambiguous inputs. Since the current independent design requires zero communication or synchronization overhead between layers, we prioritize this simple, completely decoupled structure to maximize edge inference speed.
\end{enumerate}

\end{document}
