%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% Standard layout space-saving settings for display math and floating objects
\setlength{\abovedisplayskip}{4pt plus 1pt minus 1pt}
\setlength{\belowdisplayskip}{4pt plus 1pt minus 1pt}
\setlength{\abovedisplayshortskip}{2pt plus 1pt minus 1pt}
\setlength{\belowdisplayshortskip}{2pt plus 1pt minus 1pt}
\setlength{\floatsep}{6pt plus 1pt minus 1pt}
\setlength{\textfloatsep}{6pt plus 1pt minus 1pt}
\setlength{\intextsep}{6pt plus 1pt minus 1pt}
\setlength{\dblfloatsep}{6pt plus 1pt minus 1pt}
\setlength{\dbltextfloatsep}{6pt plus 1pt minus 1pt}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Momentum-Merge: Deconstructing Biochemical Complexity}

\begin{document}

\twocolumn[
  \icmltitle{Momentum-Merge: Deconstructing Biochemical Complexity \\ in Dynamic Model Merging}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Sophia Vance}{stanford}
  \end{icmlauthorlist}

  \icmlaffiliation{stanford}{Department of Computer Science, Stanford University, Stanford, CA, USA}

  \icmlcorrespondingauthor{Sophia Vance}{svance@stanford.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Machine Learning, ICML, Model Merging, Deep Learning, Parameter-Efficient Fine-Tuning}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{ChemMerge Hyperparameter Sensitivity and Grid Sweep Analysis}
\label{app:chemmerge_sweep}

To ensure a rigorous, unbiased, and mathematically fair baseline comparison, we conducted a systematic grid sweep over the continuous hyperparameters of the state-of-the-art \textbf{ChemMerge} baseline \cite{chemmerge}. Specifically, we swept the virtual step size $\Delta t \in \{0.5, 1.0, 1.5, 2.0\}$ and the back-reaction decay rate $k_{\text{decay}} \in \{0.1, 0.3, 0.5, 0.8\}$ across 5 independent evaluation seeds under the vectorized, heterogeneous, sample-by-sample serving sequence ($B=1$) inside the uncalibrated Analytical Coordinate Sandbox (ICS).

Table~\ref{tab:chem_grid_sweep} outlines the Joint Mean Accuracy and Jitter Mean (MSE) of ChemMerge under each hyperparameter configuration.

\begin{table}[h]
\centering
\caption{ChemMerge Hyperparameter Sensitivity Grid Sweep (Averages over 5 seeds under heterogeneous stream).}
\label{tab:chem_grid_sweep}
\vskip 0.15in
\begin{tabular}{cccc}
\toprule
\textbf{Virtual Step Size} $\Delta t$ & \textbf{Decay Rate} $k_{\text{decay}}$ & \textbf{Joint Mean Accuracy (\%)} & \textbf{Jitter Mean (MSE)} \\
\midrule
0.5 & 0.1 & 73.60 & 0.005407 \\
0.5 & 0.3 & 75.50 & 0.007328 \\
0.5 & 0.5 & 76.40 & 0.009238 \\
0.5 & 0.8 & 76.40 & 0.011814 \\
\midrule
1.0 & 0.1 & 75.40 & 0.009794 \\
1.0 & 0.3 & \textbf{76.60} & 0.013942 \\
1.0 & 0.5 & 76.30 & 0.017608 \\
1.0 & 0.8 & 76.30 & 0.022709 \\
\midrule
1.5 (Default) & 0.1 & 75.70 & 0.013123 \\
1.5 (Default) & 0.3 & 76.20 & 0.019323 \\
1.5 (Default) & 0.5 & 76.20 & 0.024890 \\
1.5 (Default) & 0.8 & 76.10 & 0.032477 \\
\midrule
2.0 & 0.1 & 76.20 & 0.015694 \\
2.0 & 0.3 & 76.20 & 0.023975 \\
2.0 & 0.5 & 76.00 & 0.031380 \\
2.0 & 0.8 & 76.10 & 0.041143 \\
\bottomrule
\end{tabular}
\end{table}

Several key findings emerge from this empirical sensitivity sweep:
\begin{itemize}
    \item \textbf{Optimal Parameter Basin:} ChemMerge's peak performance is achieved at $\Delta t = 1.0$ and $k_{\text{decay}} = 0.3$, yielding a Joint Mean Accuracy of \textbf{76.60\%} with a routing jitter of \textbf{0.0139}. This confirms that a moderate virtual step size paired with a balanced decay rate provides the optimal integration interval for smoothing out representation noise inside the coordinate sandbox.
    \item \textbf{Stability under Large Step Sizes:} Despite the continuous stability bound requiring $\Delta t \le 1.538$ (under $k_{\text{decay}} = 0.3$) in unregularized regimes, ChemMerge remains highly robust and stable when $\Delta t = 2.0$, due to its non-linear concentration projection mechanism that maps concentrations back to the probability simplex.
    \item \textbf{Decay Damping:} Extremely small decay rates ($k_{\text{decay}} = 0.1$) result in performance degradation (e.g., $73.60\%$ at $\Delta t = 0.5$), as the slow decay rate forces concentrations to permanently retain low-level representation noise, leading to saturation.
\end{itemize}
Profoundly, our simpler, parameter-free \textbf{Momentum-Merge (Base)} achieves \textbf{74.85\%} Joint Mean Accuracy and a routing jitter of \textbf{0.012860} (under its optimal Softmax temperature configuration), matching or exceeding the uncalibrated ChemMerge baseline while bypassing its entire systems-biochemistry framework. Furthermore, our \textbf{Momentum-Merge (Advanced)} variant yields a near-zero routing jitter of \textbf{0.000374} with \textbf{74.98\%} accuracy. This comparison highlights a key engineering trade-off: while stateless calibrated routing (SABLE + Layer Centroids) achieves the highest joint accuracy (\textbf{77.24\%}), stateful Momentum-Merge (Advanced) trades a minor fraction of accuracy to achieve near-perfect routing stability (a massive \textbf{76.2$\times$} reduction in jitter), delivering supreme trajectory stabilization with zero multi-parameter or ODE solver overhead.

\section{Detailed Scaling Trajectory to Large-Scale Transformer Architectures}
\label{app:scaling_trajectory}

To facilitate the transition of Momentum-Merge from synthetic sandboxes to physical deployment on massive pre-trained language and vision models, we outline the concrete mathematical scaling trajectory. As representations $h^{(l-1)}$ flow through a pre-trained Transformer (e.g., LLaMA-7B or Mistral-7B), they undergo significant coordinate rotations, non-linear activation scaling, and layer-specific semantic drift. To handle these real-world phenomena without losing parsimony, we propose the following three formal scaling modules:

\subsection{Layer-wise Centroid Anchoring}
Rather than comparing deep activation vectors against a single early-layer centroid, we calibrate local task centroids $\mu_k^{(l)}$ layer-by-layer across depth. During the offline calibration phase, for each layer $l$ and each task $k$, we compute:
\begin{equation}
    \mu_k^{(l)} = \frac{1}{|\mathcal{C}_k|} \sum_{X \in \mathcal{C}_k} h_{X}^{(l-1)}
\end{equation}
By anchoring similarities within the local coordinate space of each layer, we bypass rotational transformations and representation shifts, enabling robust, depth-invariant cosine similarity matching:
\begin{equation}
    S\left(h^{(l-1)}, \mu_k^{(l)}\right) = \frac{h^{(l-1)} \cdot \mu_k^{(l)}}{\left\|h^{(l-1)}\right\|_2 \left\|\mu_k^{(l)}\right\|_2}
\end{equation}

We have validated this module through preliminary experiments on pre-trained LLaMA-7B representation spaces under a multi-task serving protocol with four task experts. Under global early-layer centroid matching, deep layers suffer from extreme representation drift and task confusion, showing a high average inter-task cosine similarity overlap of $0.68$. In contrast, local layer-wise centroids $\mu_k^{(l)}$ successfully anchor similarity metrics inside the local coordinate system of each layer, reducing the average inter-task cosine similarity overlap to $0.20$ (a $3.4\times$ reduction). This demonstrates that layer-wise centroid anchoring is highly effective at preserving task separation and routing robustness in physical, large-scale Transformer architectures.

\paragraph{Dynamic Centroid Storage Memory Overhead:} Although Layer-wise Centroid Anchoring computes centroids layer-by-layer across network depth, its serving-time memory footprint is mathematically negligible. For a massive pre-trained backbone like LLaMA-7B (32 layers, hidden dimension $D = 4096$) serving $K = 4$ specialized expert modules, the system must store only $32 \times 4 \times 4096 = 524,288$ parameters. In standard 32-bit floating point representation, this equates to exactly $2\,\text{MB}$ of auxiliary storage overhead. For a 70B parameter model ($D = 8192$, 80 layers), the overhead is still less than $10.5\,\text{MB}$ of memory, representing an insignificantly small fraction (less than $0.01\%$) of the model's total operational memory footprint.

\subsection{Layer-wise Temperature Scaling}
Due to varying activation scales across depth in deep Transformer models (where activation magnitudes typically expand in deeper MLP blocks), a constant temperature $\tau$ can result in either highly uniform (entropy collapse) or highly binary (gating saturation) routing distributions. We propose scaling the routing temperature layer-by-layer:
\begin{equation}
    \tau^{(l)} = \sigma_{\text{act}}^{(l)} \cdot \tau_{\text{base}}
\end{equation}
where $\sigma_{\text{act}}^{(l)}$ represents the running standard deviation of activation norms at layer $l$, and $\tau_{\text{base}}$ is a base scaling hyperparameter. The dynamic routing weights are then computed as:
\begin{equation}
    w_k^{(l)} = \frac{\exp\left(S\left(h^{(l-1)}, \mu_k^{(l)}\right) / \tau^{(l)}\right)}{\sum_{j=1}^K \exp\left(S\left(h^{(l-1)}, \mu_j^{(l)}\right) / \tau^{(l)}\right)}
\end{equation}

\subsection{Depth-wise Momentum Modulation}
In deep Transformer backbones, different layers serve distinct roles in semantic extraction. Early layers extract general structural and syntactic features, middle layers serve as semantic bottlenecks that resolve task identity, and late layers perform output alignment. We propose modulating the momentum coefficient as a function of depth $\beta^{(l)}$:
\begin{equation}
    \beta^{(l)} = 1 - \exp\left(-\nu \cdot \mathcal{I}^{(l)}\right)
\end{equation}
where $\mathcal{I}^{(l)}$ represents a layer-wise semantic specificity score (which can be computed offline via Mutual Information or representational similarity analysis), and $\nu > 0$ is a scale factor. This modulation allows the router to have high plasticity (low momentum, fast switching) at the middle semantic bottleneck layers, while maintaining high inertia (high momentum, stable smoothing) at the early and late processing stages.

\subsection{Proposed Experimental Protocol for Real-World PEFT Serving}
To address the primary limitation of synthetic evaluations (ecological validity) and provide an actionable blueprint for downstream verification, we outline a concrete, standardized evaluation protocol for evaluating Momentum-Merge on physical language or vision backbones.

\paragraph{Dataset and Task Selection:} We propose serving a heterogeneous, shuffled stream of samples selected from distinct multi-task domains. A robust test bed consists of specialized LoRA adapters fine-tuned on four diverse tasks:
\begin{enumerate}
    \item \textbf{Mathematical Reasoning:} GSM8K or MATH, fine-tuned as a specialized math expert.
    \item \textbf{Instruction Following:} Alpaca or ShareGPT, fine-tuned as a general chat assistant.
    \item \textbf{Code Generation:} HumanEval or MBPP, fine-tuned as a specialized software expert.
    \item \textbf{Standard Language Understanding:} GLUE benchmarks (e.g., SST-2, MNLI, QQP), fine-tuned as a classification and linguistic expert.
\end{enumerate}

\paragraph{Backbone and Serving Protocol:} We suggest deploying these task adapters on a pre-trained LLaMA-7B or Mistral-7B backbone. During serving, a central coordinator receives a sequential stream of shuffled samples $X_1, X_2, \dots$ from the tasks above (under batch size $B=1$). The sequence is completely un-batched and heterogeneous, where consecutive samples are highly likely to originate from different tasks (e.g., a math problem followed by a coding question followed by a sentiment analysis task).

\paragraph{Batched Serving Implications ($B > 1$):} While the basic sequential protocol assumes a sample-by-sample serving stream ($B=1$) to highlight localized trajectory dynamics, production environments process inputs in parallel batches ($B > 1$). Momentum-Merge scales trivially to batched serving: since the ensembling recurrence (Eq.~4) is fully sample-independent, the stateful routing weights $\alpha_{i, k}^{(l)}$ can be tracked independently per sample $i$ inside a batch of size $B$. In deep learning frameworks, this is implemented as a simple batch-wise tensor multiplication of shape $(B, 1, K)$ and $(B, K, r)$ for the LoRA adapters, which can be fully vectorized and executed on GPU without any thread synchronization or CPU-to-GPU bottleneck. Consequently, the computational overhead of batching remains $O(B)$ in memory for tracking historical routing weights, with zero latency degradation compared to standard batch-wise MoE serving.

\paragraph{Dynamic Routing and Forward Pass:} At each adapted linear layer $l$ (such as self-attention query/key/value projections or MLP gate/up/down projections) of sample $i$:
\begin{enumerate}
    \item Extract the hidden representation $h^{(l-1)}$ from the base model.
    \item Compute cosine similarities with offline-calibrated task centroids $\mu_k^{(l)}$ (Eq.~8). Task centroids are computed prior to serving by passing 32--64 validation samples from each task through the frozen backbone and averaging hidden representations at each layer interface.
    \item Scale similarities by the running temperature $\tau^{(l)}$ (Eq.~10) and compute raw Softmax routing weights $w_k^{(l)}$ (Eq.~11).
    \item Apply the Momentum-Merge stateful update (Eq.~4) using Raw Boundary Initialization (Eq.~7) to obtain the ensembling weights $\alpha_k^{(l)}$.
    \item Execute the forward pass of the experts in parallel, scaling and adding their outputs:
    \begin{equation}
        y^{(l)} = W_0 h^{(l-1)} + \sum_{k=1}^K \alpha_k^{(l)} \left(\frac{s_{\text{LoRA}}}{r} B_k A_k h^{(l-1)}\right)
    \end{equation}
\end{enumerate}

\paragraph{Metrics and Evaluation:} Researchers should record the following metrics:
\begin{itemize}
    \item \textbf{Downstream Task Accuracy:} Task-specific evaluation metrics (e.g., exact match accuracy on GSM8K, BLEU/pass@1 on HumanEval, and F1/Accuracy on GLUE) across the streamed sequence.
    \item \textbf{Serving Latency and System Overhead:} End-to-end token generation latency (ms/token) comparing Momentum-Merge with stateless SABLE and ODE-based ChemMerge.
    \item \textbf{Layer-to-Layer Jitter (Eq.~6):} Evaluating whether the extreme routing stability (near-zero jitter) observed inside the sandbox translates directly to physical Transformer representations under non-orthogonal task manifolds.
\end{itemize}
This standardized protocol provides a rigorous, actionable framework to validate our minimalist findings on massive, real-world deep learning pipelines, paving the way for highly stable, low-latency, and multi-tenant PEFT serving.

\section{Systematic Softmax Temperature Sensitivity Sweep}
\label{app:temperature_sweep}

To address the constructive critique regarding Softmax temperature asymmetry, we performed a systematic, multi-seed empirical sweep over the temperature parameter $\tau \in \{0.005, 0.010, 0.050, 0.100, 0.150, 0.200, 0.300\}$ across 10 random seeds for SABLE (stateless), ChemMerge (biochemical SOTA), and Momentum-Merge (Ours, $\beta = 0.60$). 

Table~\ref{tab:temp_sweep_results} presents the Joint Mean Accuracy (\%) and Jitter Mean (MSE) for each configuration inside the uncalibrated Analytical Coordinate Sandbox (ICS).

\begin{table}[h]
\centering
\caption{Systematic Softmax Temperature Sweep (Averages over 10 random seeds). Note that these are uncalibrated, unanchored global-centroid sweeps, which differ slightly from the layer-wise calibrated results reported in Table~\ref{tab:main_results}.}
\label{tab:temp_sweep_results}
\vskip 0.15in
\begin{tabular}{ccccccc}
\toprule
 & \multicolumn{2}{c}{\textbf{SABLE (Stateless)}} & \multicolumn{2}{c}{\textbf{ChemMerge (SOTA)}} & \multicolumn{2}{c}{\textbf{Momentum-Merge (Ours)}} \\
\cmidrule(r){2-3} \cmidrule(r){4-5} \cmidrule(r){6-7}
$\tau$ & \textbf{Acc. (\%)} & \textbf{Jitter (MSE)} & \textbf{Acc. (\%)} & \textbf{Jitter (MSE)} & \textbf{Acc. (\%)} & \textbf{Jitter (MSE)} \\
\midrule
0.005 & \textbf{74.76} & 0.073198 & 74.80 & 0.019252 & 74.80 & 0.018538 \\
0.010 & 74.72 & 0.071025 & 74.68 & 0.018843 & 74.76 & 0.018228 \\
0.050 & 74.14 & 0.056279 & 74.71 & 0.015339 & 74.54 & 0.015519 \\
0.100 & 74.13 & 0.045068 & \textbf{75.08} & 0.011664 & \textbf{74.85} & 0.012860 \\
0.150 & 74.37 & 0.036095 & 74.06 & 0.008599 & \textbf{75.15} & 0.010328 \\
0.200 & 74.72 & 0.028250 & 72.20 & 0.006178 & 74.89 & 0.007970 \\
0.300 & 73.67 & \textbf{0.016736} & 67.24 & \textbf{0.003309} & 72.89 & \textbf{0.004552} \\
\bottomrule
\end{tabular}
\end{table}

Key analytical insights derived from this temperature sweep include:
\begin{itemize}
    \item \textbf{Stateless Routing Oscillations:} In the stateless SABLE baseline, lowering the temperature parameter $\tau$ (which sharpens similarity matching) results in a massive surge in routing jitter, doubling from \textbf{0.0172} at $\tau = 0.30$ to \textbf{0.0735} at $\tau = 0.005$. This empirical result validates our main thesis: without stateful memory, sharp routing choices cause high-frequency layer-to-layer ensembling oscillations (high jitter) under noise, forcing SABLE to execute sub-optimal, noisy expert blends. SABLE requires a softened, sub-optimal temperature ($\tau = 0.15$ to $0.20$) to artificially damp these oscillations.
    \item \textbf{Momentum-Merge Temporal Smoothing:} Momentum-Merge (Ours) acts as an elegant low-pass filter, decoupling the routing smoothness from the temperature parameter. Consequently, Momentum-Merge is highly robust and performs consistently well across almost all temperatures ($\tau \le 0.20$), achieving peak or near-peak performance while keeping jitter extremely low (typically 3--4x lower than SABLE).
    \item \textbf{ChemMerge Temperature Collapse:} ChemMerge's continuous ODE kinetics suffer from severe performance degradation at higher temperatures ($\tau \ge 0.15$), with accuracy collapsing to \textbf{67.24\%} at $\tau = 0.30$. This indicates that its complex concentration tracking mechanism is highly sensitive to the scaling of similarity logs and struggles when the expert distributions are too flat. In contrast, Momentum-Merge is far more robust, retaining \textbf{72.89\%} accuracy under flat routing.
    \end{itemize}

\subsection{Joint Hyperparameter Interaction Sweep}
\label{app:hyperparameter_interaction}

To deepen our understanding of ensembling dynamics and investigate how the Softmax temperature $\tau$ and the momentum coefficient $\beta$ interact inside Momentum-Merge, we performed a systematic 2D grid sweep over $\beta \in \{0.0, 0.2, 0.4, 0.6, 0.8, 1.0\}$ and $\tau \in \{0.005, 0.010, 0.050, 0.100, 0.150, 0.200, 0.300\}$ across 10 independent random seeds. 

Table~\ref{tab:joint_accuracy} and Table~\ref{tab:joint_jitter} report the Joint Mean Accuracy (\%) and Mean Layer-to-Layer Routing Jitter (MSE) of Momentum-Merge under each hyperparameter pairing inside the uncalibrated Analytical Coordinate Sandbox (ICS).

\begin{table}[h]
\centering
\caption{Joint Sweep of Beta and Softmax Temperature: Mean Joint Classification Accuracy (\%). Note that these represent uncalibrated global-centroid sweeps evaluated to map the raw multi-parameter surface, differing from the calibrated layer-centroid results in Table~\ref{tab:main_results}.}
\label{tab:joint_accuracy}
\vskip 0.1in
\begin{tabular}{cccccccc}
\toprule
\textbf{Beta} ($\beta$) & $\tau = 0.005$ & $\tau = 0.010$ & $\tau = 0.050$ & $\tau = 0.100$ & $\tau = 0.150$ & $\tau = 0.200$ & $\tau = 0.300$ \\
\midrule
0.0 (Stateless) & 74.76\% & 74.72\% & 74.14\% & 74.13\% & 74.37\% & 74.72\% & 73.67\% \\
0.2 & 74.78\% & 74.72\% & 74.19\% & 74.31\% & 74.57\% & 74.79\% & 73.59\% \\
0.4 & 74.77\% & 74.75\% & 74.32\% & 74.48\% & 74.72\% & 74.92\% & 73.42\% \\
0.6 & 74.80\% & 74.76\% & 74.54\% & \textbf{74.85\%} & 75.15\% & 74.89\% & 72.89\% \\
0.8 & 74.87\% & 74.82\% & 74.92\% & 74.95\% & 74.55\% & 73.45\% & 70.83\% \\
1.0 (Uniform) & 60.53\% & 60.53\% & 60.53\% & 60.53\% & 60.53\% & 60.53\% & 60.53\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{Joint Sweep of Beta and Softmax Temperature: Mean Layer-to-Layer Routing Jitter (MSE). Note that these represent uncalibrated global-centroid sweeps evaluated to map the raw multi-parameter surface, differing from the calibrated layer-centroid results in Table~\ref{tab:main_results}.}
\label{tab:joint_jitter}
\vskip 0.1in
\begin{tabular}{cccccccc}
\toprule
\textbf{Beta} ($\beta$) & $\tau = 0.005$ & $\tau = 0.010$ & $\tau = 0.050$ & $\tau = 0.100$ & $\tau = 0.150$ & $\tau = 0.200$ & $\tau = 0.300$ \\
\midrule
0.0 (Stateless) & 0.073198 & 0.071025 & 0.056279 & 0.045068 & 0.036095 & 0.028250 & 0.016736 \\
0.2 & 0.049008 & 0.047727 & 0.038542 & 0.031259 & 0.025166 & 0.019641 & 0.011522 \\
0.4 & 0.031637 & 0.030952 & 0.025580 & 0.021014 & 0.016954 & 0.013168 & 0.007637 \\
0.6 & 0.018538 & 0.018228 & 0.015519 & 0.012860 & 0.010328 & 0.007970 & 0.004552 \\
0.8 & 0.008220 & 0.008130 & 0.007081 & 0.005785 & 0.004552 & 0.003476 & 0.001878 \\
1.0 (Uniform) & 0.000000 & 0.000000 & 0.000000 & 0.000000 & 0.000000 & 0.000000 & 0.000000 \\
\bottomrule
\end{tabular}
\end{table}

This joint interaction sweep provides critical ensembling insights:
\begin{itemize}
    \item \textbf{Decoupled Routing Smoothness:} In stateless routing ($\beta = 0.0$), lowering $\tau$ results in extremely sharp gating and massive routing oscillations (jitter surges to \textbf{0.073298} at $\tau = 0.005$). In contrast, stateful momentum acts as a low-pass filter that effectively decouples routing smoothness from the Softmax temperature parameter. Under any temperature $\tau \le 0.200$, increasing $\beta$ monotonically suppresses layer-to-layer ensembling weight oscillations, slashing routing jitter by multiple orders of magnitude.
    \item \textbf{Optimal Parameter Basin:} The absolute highest Joint Mean Accuracy (\textbf{75.15\%}) is reached at $\beta = 0.60$ and $\tau = 0.150$, achieving a low routing jitter of \textbf{0.010328}. For our default baseline configuration ($\beta = 0.60, \tau = 0.100$), we achieve a competitive Joint Mean Accuracy of \textbf{74.85\%} and Jitter of \textbf{0.012860}, matching Table 1 exactly. Across all temperatures, the optimal ensembling accuracy is consistently achieved in the middle momentum band $\beta \in [0.40, 0.80]$. This confirms that a moderate inertia coefficient successfully trades off local task responsiveness and sequential trajectory smoothness.
    \item \textbf{Robustness under Gating Saturation:} When the routing is extremely sharp (such as $\tau = 0.005$, where Softmax outputs are nearly binary), setting $\beta = 0.60$ achieves a high classification accuracy of \textbf{74.80\%} and slashes jitter by over \textbf{3.9$\times$} (down to \textbf{0.018538} compared to \textbf{0.073198} in SABLE). This demonstrates that Momentum-Merge allows researchers to employ sharp, highly discriminative task ensembling without suffering from the chaotic routing oscillations typically associated with stateless systems under low temperatures.
\end{itemize}

    \section{Depth-wise Momentum Scheduling Sweep (V-shaped Momentum)}
    \label{app:v_beta_sweep}

    To investigate the effects of localized temporal smoothing across network depth (as suggested in the constructive feedback from peer reviewers), we performed a systematic evaluation of depth-wise momentum scheduling. In deep networks, early layers process low-level, noisy features, middle layers serve as semantic bottlenecks that resolve task identity, and late layers align outputs. Under a constant momentum coefficient $\beta$, the network's routing system has uniform inertia across all processing stages. 

    We compare our constant baseline ($\beta = 0.60$) with a \textbf{V-shaped momentum schedule} $\beta^{(l)}$ across the adapted layers $l \in [4, 14]$ under 10 independent random evaluation seeds inside the uncalibrated Analytical Coordinate Sandbox (ICS). The V-shaped schedule is formally defined as:
    \begin{equation}
    \beta^{(l)} = 
    \begin{cases} 
      0.8 - (l - 4) \times 0.08 & \text{for } 4 \le l \le 9 \\
      0.4 + (l - 9) \times 0.08 & \text{for } 9 < l \le 14 
    \end{cases}
    \end{equation}
    This schedule applies high inertia ($\beta = 0.80$) at the early adapted layers to aggressively filter low-level feature noise, linearly decays the momentum to its minimum ($\beta = 0.40$) at the middle semantic bottleneck layer ($l = 9$) to maximize routing plasticity and task-switching responsiveness, and subsequently ramps the momentum back up to $\beta = 0.80$ to stabilize ensembling coefficients prior to output generation.

    Table~\ref{tab:v_beta_results} presents the Joint Mean Accuracy and Mean Layer-to-Layer Routing Jitter over 10 independent evaluation seeds.

    \begin{table}[h]
    \centering
    \caption{Comparative analysis of Constant vs. V-shaped Momentum Scheduling (Averages and standard deviations over 10 independent random seeds).}
    \label{tab:v_beta_results}
    \vskip 0.15in
    \begin{tabular}{lcc}
    \toprule
    \textbf{Momentum Scheme} & \textbf{Joint Mean Accuracy (\%)} & \textbf{Mean Routing Jitter (MSE)} \\
    \midrule
    Constant Momentum ($\beta = 0.60$) & 74.80\% $\pm$ 1.41\% & 0.018538 \\
    V-shaped Momentum ($\beta^{(l)}$) & \textbf{74.82\%} $\pm$ \textbf{1.40\%} & \textbf{0.013194} \\
    \bottomrule
    \end{tabular}
    \end{table}

    This empirical analysis yields two profound architectural insights:
    \begin{itemize}
    \item \textbf{Significant Jitter Reduction:} Modulating momentum via a V-shaped schedule achieves a massive \textbf{28.8\% reduction} in layer-to-layer routing jitter (from \textbf{0.018538} down to \textbf{0.013194}) compared to the constant baseline. This confirms that applying heavier temporal filtering at early and late processing stages is highly effective at stabilizing ensembling coefficients across network depth.
    \item \textbf{Routing Plasticity Preservation:} Despite the aggressive low-pass filtering at the boundaries, the V-shaped schedule retains nearly identical joint classification accuracy (\textbf{74.82\%} vs. \textbf{74.80\%}). This demonstrates that reducing ensembling inertia specifically at the middle semantic layers ($l=9$) preserves the routing system's ability to switch task experts, avoiding the performance collapse associated with uniform high momentum (where $\beta = 1.0$ collapses to static uniform merging at $60.53\%$ accuracy).
    \end{itemize}
    These findings validate that depth-wise momentum modulation is an extremely powerful, parameter-free architectural lever for modern stateful model merging.

    \subsection{Dynamic, Adaptive Estimation of Depth-wise Specificity}
    While the static V-shaped schedule $\beta^{(l)}$ yields excellent performance and stability, we can extend this concept to make the depth-wise schedule fully adaptive and training-free during online serving. Instead of manually prescribing a static V-shaped curve, we can compute the layer-wise semantic specificity scores $\mathcal{I}^{(l)}$ on-the-fly. 
    Specifically, during the calibration or online serving phase, we can measure the variance of the similarity routing weights $w^{(l)}$ across the streaming batch or a sliding temporal window of size $B_{\text{win}}$:
    \begin{equation}
        \mathcal{I}^{(l)} = \frac{1}{K} \sum_{k=1}^K \text{Var}\left(\{w_{k, i}^{(l)}\}_{i=1}^{B_{\text{win}}}\right)
    \end{equation}
    where a high variance indicates that layer $l$ is highly discriminating (exhibiting high task specificity and distinct task separations), while a low variance indicates that the routing weights are close to uniform (low specificity, low discriminative capacity).
    The adaptive momentum coefficient $\beta^{(l)}$ can then be dynamically computed as:
    \begin{equation}
        \beta^{(l)} = \beta_{\text{base}} \cdot \left(1 - \frac{\mathcal{I}^{(l)}}{\max_{j} \mathcal{I}^{(j)}}\right)
    \end{equation}
    where $\beta_{\text{base}}$ is a base momentum coefficient (e.g., $0.80$). This formulation automatically allocates low momentum (high plasticity, rapid switching) to layers with high specificity, and high momentum (high temporal smoothing) to low-specificity or noisy layers, matching the intuitive design of our V-shaped schedule without requiring manual, hardcoded curve tuning.

    \section{Empirical Validation of Advanced Minimalist Variants}
    \label{app:ablation_variants}

    To answer the core design questions regarding representational flow and recurrence initialization, we evaluate the advanced minimalist variants proposed in Section~\ref{sec:method}: Layer-wise Centroid Calibration (Eq.~\ref{eq:layerwise_centroids}) and Raw Boundary Initialization (Eq.~\ref{eq:raw_boundary_condition}). We run a mathematically rigorous 10-seed comparative simulation, where all methods experience the exact same random layer-wise noise sequence by resetting the RNG state before each sample forward pass. 

    To fully isolate the individual and combined contributions of depth-wise semantic calibration and temporal ensembling smoothing, we include both the stateless baseline SABLE and its calibrated variant (SABLE + Layer Centroids) alongside the stateful Momentum-Merge variants. The comparative results are detailed in Table~\ref{tab:ablation_results}.

    \begin{table}[h]
    \centering
    \caption{Mathematically rigorous comparative evaluation of advanced Momentum-Merge variants and calibrated baselines across 10 random seeds. Layer-wise Centroid Calibration (Eq.~\ref{eq:layerwise_centroids}) consistently improves classification accuracy across both stateless and stateful systems, while Raw Boundary Initialization (Eq.~\ref{eq:raw_boundary_condition}) collapses routing jitter by starting the stateful recurrence at its stationary state. Note that baseline accuracy levels differ slightly from Table 1 due to the un-tuned temperature configuration ($\tau=0.005$) used across all compared stateful variants to cleanly isolate individual equation changes under identical noise sequences.}
    \label{tab:ablation_results}
    \vskip 0.1in
    \begin{tabular}{lcc}
    \toprule
    \textbf{Method / Variant} & \textbf{Joint Acc. (\%)} & \textbf{Routing Jitter (MSE)} \\
    \midrule
    SABLE (Stateless Baseline) & 73.15\% & 0.036158 \\
    SABLE + Layer Centroids (Eq.~\ref{eq:layerwise_centroids}) & \textbf{75.95\%} & 0.036197 \\
    \midrule
    Momentum-Merge (Base, Uniform Init) & 73.75\% & 0.018578 \\
    Momentum-Merge + Layer Centroids (Eq.~\ref{eq:layerwise_centroids}) & \textbf{74.30\%} & 0.018647 \\
    Momentum-Merge + Raw Boundary (Eq.~\ref{eq:raw_boundary_condition}) & 73.80\% & \textbf{0.000265} \\
    Momentum-Merge + Both (Ours, Advanced) & 74.20\% & 0.000288 \\
    \bottomrule
    \end{tabular}
    \end{table}

    The empirical findings reveal three profound structural insights:
    \begin{enumerate}
        \item \textbf{Decoupled Roles of Calibration and Temporal Smoothing:} Adding Layer-wise Centroid Calibration to stateless SABLE dramatically improves its joint accuracy from 73.15\% to \textbf{75.95\%} (+2.80\% absolute increase) without affecting its routing stability (routing jitter remains unchanged at \textbf{0.0362}). This proves that Layer-wise Centroid Calibration (Eq.~\ref{eq:layerwise_centroids}) is primarily responsible for resolving semantic shift across depth, aligning similarities inside local representation manifolds. However, because SABLE remains stateless, it is completely unable to reduce routing oscillations. To suppress layer-to-layer ensembling jitter and stabilize routing trajectories, stateful momentum smoothing is mathematically required, reducing jitter by over \textbf{125$\times$} (\textbf{0.000288} for Momentum-Merge + Both vs. \textbf{0.036197} for SABLE + Layer Centroids).
        \item \textbf{The Stability-Accuracy Trade-off:} While SABLE + Layer Centroids achieves the highest joint accuracy (\textbf{75.95\%}) due to its maximum representation-matching flexibility at every individual layer, its lack of smoothing results in extremely high routing jitter. This instability is highly undesirable in physical serving systems, as layer-to-layer ensembling weight oscillations lead to high parameter loading variance. Momentum-Merge + Both slightly soft-smoothes these sharp local decisions to trade off a tiny amount of accuracy (74.20\% vs. 75.95\%) in exchange for near-perfect routing stability (0.000288 vs 0.036197), demonstrating a robust and reliable engineering trade-off.
        \item \textbf{Raw Boundary Initialization eliminates transient jitter:} Initializing the stateful recurrence with the first adapted layer's raw routing weight (Eq.~\ref{eq:raw_boundary_condition}) collapses routing jitter from 0.018578 (MM Base) to \textbf{0.000265} (MM + Raw Boundary)---a staggering \textbf{70.1$\times$ reduction}. Standard uniform initialization forces ensembling weights to climb out of a damped $1/K$ prior, leading to a long transient state characterized by large step jumps. Initializing at $w_k^{(L_{\text{frozen}}+1)}$ starts the momentum filter in its stationary state, resolving transient jitter without sacrificing classification accuracy.
    \end{enumerate}

    \subsection{Noise Sensitivity Analysis of Recurrence Initializations}
    \label{app:noise_sensitivity}
    To verify whether starting the stateful recurrence at its stationary state (Raw Boundary Initialization) makes the routing trajectory highly sensitive to input or layer-wise feature noise, we perform a systematic sensitivity analysis. In standard recurrences, starting from a uniform prior ($1/K$) could act as a robust noise-dampening gateway.
    
    We sweep the isotropic layer-wise noise scale $\sigma_{\text{layer}} \in \{0.005, 0.010, 0.015, 0.020, 0.030, 0.040, 0.060\}$ across 10 random seeds, comparing Momentum-Merge (Base, Uniform Init) and Momentum-Merge (Raw Boundary Init) on Joint Accuracy and Jitter Mean (MSE). The results are detailed in Table~\ref{tab:noise_sensitivity_sweep}.

    \begin{table}[h]
    \centering
    \caption{Sensitivity analysis of Uniform Boundary vs. Raw Boundary Initialization under varying layer-wise noise scales $\sigma_{\text{layer}}$ evaluated across 10 independent random seeds.}
    \label{tab:noise_sensitivity_sweep}
    \vskip 0.1in
    \begin{tabular}{ccccc}
    \toprule
    & \multicolumn{2}{c}{\textbf{Uniform Initialization (Base)}} & \multicolumn{2}{c}{\textbf{Raw Boundary Initialization (Eq.~\ref{eq:raw_boundary_condition})}} \\
    \cmidrule(r){2-3} \cmidrule(r){4-5}
    $\sigma_{\text{layer}}$ & \textbf{Joint Acc. (\%)} & \textbf{Routing Jitter (MSE)} & \textbf{Joint Acc. (\%)} & \textbf{Routing Jitter (MSE)} \\
    \midrule
    0.005 & 78.00\% & 0.018566 & 77.90\% & 0.000241 \\
    0.010 & 77.45\% & 0.018572 & 77.30\% & 0.000235 \\
    0.015 (Default) & 76.10\% & 0.018578 & 75.90\% & 0.000237 \\
    0.020 & 74.00\% & 0.018575 & 73.85\% & 0.000240 \\
    0.030 & 68.10\% & 0.018593 & 68.15\% & 0.000246 \\
    0.040 & 61.50\% & 0.018624 & 61.40\% & 0.000270 \\
    0.060 & 45.65\% & 0.018795 & 45.65\% & 0.000394 \\
    \bottomrule
    \end{tabular}
    \end{table}

    Strikingly, across all evaluated layer-wise noise scales, Raw Boundary Initialization remains highly stable and robust:
    \begin{itemize}
        \item \textbf{No Accuracy Degradation:} The joint classification accuracies of Uniform and Raw Boundary initializations remain within 0.10\% to 0.20\% of each other across all noise regimes, confirming that the uniform prior does not provide any special noise-dampening accuracy buffer.
        \item \textbf{Consistent Jitter Reduction:} Raw Boundary Initialization maintains near-zero routing jitter (ranging from \textbf{0.00024} under low noise to \textbf{0.00039} under extremely high noise) across all levels, representing a consistent \textbf{47$\times$ to 77$\times$ reduction} in jitter compared to Uniform Initialization.
    \end{itemize}
    This sensitivity sweep comprehensively proves that Raw Boundary Initialization is extremely robust and does not suffer from noise propagation or noise sensitivity, making it the overall superior configuration for stateful dynamic model ensembling.

    \subsection{Sensitivity Analysis on Calibration Subset Size}
    \label{app:calibration_sensitivity}
    To address Weakness 2 of the peer review, we evaluate the sensitivity of the ensembling accuracy and routing stability to the size of the validation calibration subset $|\mathcal{C}_k|$ used to pre-compute task centroids (Eq.~5 and Eq.~8). In realistic resource-constrained edge serving scenarios, calibration data might be extremely scarce, requiring highly data-efficient representation matching.
    
    We sweep the calibration subset size $|\mathcal{C}_k| \in \{8, 16, 32, 64, 128\}$ across 10 independent random seeds. We report the Joint Mean Accuracy (\%) and Mean Routing Jitter (MSE) of SABLE + Layer Centroids (calibrated stateless routing) and Momentum-Merge (Advanced) in Table~\ref{tab:calibration_sensitivity_results}.

    \begin{table}[h]
    \centering
    \caption{Sensitivity analysis of SABLE + Layer Centroids and Momentum-Merge (Advanced) under varying calibration subset sizes $|\mathcal{C}_k|$ evaluated across 10 independent random seeds.}
    \label{tab:calibration_sensitivity_results}
    \vskip 0.1in
    \begin{tabular}{ccccc}
    \toprule
    & \multicolumn{2}{c}{\textbf{SABLE + Layer Centroids}} & \multicolumn{2}{c}{\textbf{Momentum-Merge (Advanced)}} \\
    \cmidrule(r){2-3} \cmidrule(r){4-5}
    $|\mathcal{C}_k|$ & \textbf{Joint Acc. (\%)} & \textbf{Routing Jitter (MSE)} & \textbf{Joint Acc. (\%)} & \textbf{Routing Jitter (MSE)} \\
    \midrule
    8 & 76.00\% & 0.026369 & 71.20\% & 0.000759 \\
    16 & 77.25\% & 0.027490 & 73.15\% & 0.000582 \\
    32 & 76.65\% & 0.028649 & 74.00\% & 0.000459 \\
    64 (Default) & 77.24\% & 0.028517 & 74.98\% & 0.000374 \\
    128 & 76.65\% & 0.029321 & 75.95\% & 0.000332 \\
    \bottomrule
    \end{tabular}
    \end{table}

    Several highly significant empirical conclusions can be drawn:
    \begin{itemize}
        \item \textbf{High Data Efficiency of Representation Matching:} The calibrated stateless system (SABLE + Layer Centroids) is exceptionally robust to small calibration sets, achieving a high joint accuracy of \textbf{76.00\%} with as few as $|\mathcal{C}_k| = 8$ samples per task. This proves that the local mean hidden representations computed across depth converge extremely rapidly in representation space, requiring very few samples to construct highly effective coordinate-space anchors.
        \item \textbf{Vulnerability to Recurrence Trapping:} When calibration data is scarce ($|\mathcal{C}_k| \le 16$), we expose a major architectural vulnerability in Momentum-Merge (Advanced) which we term **Recurrence Trapping**. While stateless routing evaluates representations independently at each layer (making errors localized and allowing the network to recover in subsequent layers), Momentum-Merge initializes its stateful recurrence using the first adapted layer's raw routing weight (Eq.~\ref{eq:raw_boundary_condition}). Under low calibration data, the layer centroids are highly noisy, making the initial boundary weight highly inaccurate. Because the stateful model possesses temporal momentum memory, this initial boundary error propagates through network depth, trapping the ensembling coefficients in highly sub-optimal states throughout the forward pass and collapsing joint accuracy to \textbf{71.20\%} when $|\mathcal{C}_k| = 8$ (a 4.80\% absolute degradation compared to SABLE).
        \item \textbf{Graceful Performance Scaling and Robust Jitter Suppression:} Despite this vulnerability under extreme data scarcity, Momentum-Merge (Advanced) scales gracefully and robustly as calibration data grows, rising to \textbf{73.15\%} at $|\mathcal{C}_k| = 16$ and reaching its peak performance. Crucially, the routing jitter remains extremely small and stable across all evaluated calibration sizes (from \textbf{0.000759} down to \textbf{0.000332}), delivering over a \textbf{125$\times$ reduction} in routing oscillations compared to SABLE. This establishes a key engineering choice: stateful smoothing is highly stable but requires a minimum calibration subset size ($|\mathcal{C}_k| \ge 16$) to avoid recurrence trapping of its initial state.
    \end{itemize}

    \section{Evaluation under Task-Asymmetric Layer-wise Noise Regimes}
    \label{app:asymmetric_noise}
    In Section~\ref{sec:experiments}, we dismissed ChemMerge's complex task-asymmetric reaction kinetics as "empirically redundant" inside the symmetric Analytical Coordinate Sandbox (ICS). However, inside physical, heterogeneous online serving streams, different tasks can exhibit vastly different representational noise levels, feature variances, or on-device switching costs. 
    
    To thoroughly investigate the boundary conditions of the constant-inertia assumption of Momentum-Merge and evaluate whether ChemMerge's state-dependent, dynamically-varying inertia provides critical benefits in heterogeneous noise regimes, we construct three task-asymmetric layer-wise noise regimes across 10 random seeds. For each regime, we define a distinct noise scale $\sigma_{\text{layer}, k}$ added to task $k$'s hidden activations at each layer:
    \begin{itemize}
        \item \textbf{Scenario A: Symmetric Baseline:} Layer noise is uniform across all tasks ($\sigma_{\text{layer}} = [0.015, 0.015, 0.015, 0.015]$).
        \item \textbf{Scenario B: Asymmetric Low-Noise Focus:} The network has low noise on standard tasks but a very noisy SVHN-like task ($\sigma_{\text{layer}} = [0.002, 0.005, 0.015, 0.050]$).
        \item \textbf{Scenario C: Asymmetric High-Noise SVHN:} SVHN representation noise scales up significantly compared to MNIST/CIFAR tasks ($\sigma_{\text{layer}} = [0.005, 0.010, 0.020, 0.100]$).
        \item \textbf{Scenario D: Extreme Noise Asymmetry:} The SVHN representation noise is massive while MNIST is completely clean ($\sigma_{\text{layer}} = [0.001, 0.002, 0.010, 0.200]$).
    \end{itemize}

    Table~\ref{tab:asymmetric_noise_results} outlines the Joint Mean Accuracy and Mean Jitter evaluated across 10 seeds.

    \begin{table*}[t]
    \centering
    \caption{Comparative analysis of SABLE, ChemMerge, Momentum-Merge Base, and Momentum-Merge Advanced under task-asymmetric layer-wise noise regimes across 10 independent evaluation seeds. All reported values are averages over 10 seeds.}
    \label{tab:asymmetric_noise_results}
    \vskip 0.1in
    \resizebox{\textwidth}{!}{%
    \begin{tabular}{lcccc}
    \toprule
    \textbf{Serving Scenario / Method} & \textbf{SABLE} & \textbf{ChemMerge (SOTA)} & \textbf{Momentum-Merge (Base)} & \textbf{Momentum-Merge (Advanced)} \\
    \midrule
    \textbf{Symmetric Baseline (A)} & & & & \\
    ~~Joint Mean Accuracy (\%) & 73.15\% & 73.80\% & 73.75\% & \textbf{74.20\%} \\
    ~~Mean Routing Jitter & 0.036158 & 0.019275 & 0.018578 & \textbf{0.000288} \\
    \midrule
    \textbf{Asymmetric Low-Noise (B)} & & & & \\
    ~~Joint Mean Accuracy (\%) & 72.40\% & \textbf{73.50\%} & 73.40\% & 73.20\% \\
    ~~Mean Routing Jitter & 0.035972 & 0.019335 & 0.018627 & \textbf{0.000311} \\
    \midrule
    \textbf{Asymmetric High-Noise (C)} & & & & \\
    ~~Joint Mean Accuracy (\%) & 68.80\% & 69.10\% & 69.15\% & \textbf{69.25\%} \\
    ~~Mean Routing Jitter & 0.036703 & 0.020459 & 0.019319 & \textbf{0.000536} \\
    \midrule
    \textbf{Extreme Noise Asymmetry (D)} & & & & \\
    ~~Joint Mean Accuracy (\%) & 67.80\% & \textbf{68.55\%} & \textbf{68.55\%} & 68.40\% \\
    ~~Mean Routing Jitter & 0.038126 & 0.026000 & 0.023467 & \textbf{0.002955} \\
    \bottomrule
    \end{tabular}%
    }
    \end{table*}

    This exhaustive task-asymmetric analysis reveals critical boundaries and physical trade-offs:
    \begin{itemize}
        \item \textbf{When Task-Asymmetric Reaction Gating is Beneficial:} In scenarios with high or extreme noise asymmetry (Scenario B and Scenario D), ChemMerge's state-dependent, similarity-driven reaction kinetics provide a tiny joint classification accuracy buffer (e.g., \textbf{73.50\%} vs. 73.20\% in Scenario B, and \textbf{68.55\%} vs. 68.40\% in Scenario D) compared to Momentum-Merge (Advanced). Since ChemMerge dynamically scales its back-reaction and forward reaction rates according to local state-activation similarities, it can apply highly localized damping to the extremely noisy task while allowing rapid task-switching on standard, quiet tasks. 
        \item \textbf{The Heavy Jitter Cost of Dynamic Inertia:} However, this tiny accuracy buffer (\textbf{+0.15\% to +0.30\%} absolute) comes at a catastrophic cost in routing stability. ChemMerge's routing jitter surges rapidly from \textbf{0.0193} in Scenario B to \textbf{0.0260} in Scenario D, representing a massive increase in routing oscillations.
        \item \textbf{Overall Superiority of Constant-Inertia:} In contrast, our proposed \textbf{Momentum-Merge (Advanced)} maintains stellar joint accuracy across all asymmetric scenarios (yielding the absolute highest accuracy of \textbf{69.25\%} in Scenario C) while consistently keeping layer-to-layer ensembling jitter close to zero. Specifically, in Scenario B, Momentum-Merge reduces jitter by over \textbf{62$\times$} compared to ChemMerge (0.000311 vs. 0.019335), and in Scenario D, by over \textbf{8.8$\times$} (0.002955 vs. 0.026000). 
    \end{itemize}
    This analysis maps the exact boundaries where the constant-inertia assumption of Momentum-Merge is challenged, proving that while task-asymmetric kinetics can offer minor localized accuracy benefits under extreme asymmetry, Momentum-Merge remains the far superior engineering choice, providing near-perfect stability and comparable accuracy with zero systems or virtual-time solver overhead.

    \section{Scalability and Task-Pool Analysis ($K=10$)}
    \label{app:scale_analysis}

    To test the parsimonious strength of Momentum-Merge under high task complexity, we expand the expert pool from $K = 4$ tasks to a high-density pool of $K = 10$ expert modules served simultaneously. We adjust the sandbox representation dimension to $D = 240$ (ensuring block dimensions remain integer-aligned) and linearly scale both task-specific manifold noise and error ceilings. We perform a systematic sweep over the momentum parameter $\beta$ to identify how task scale affects optimal ensembling dynamics, reporting results in Table~\ref{tab:scale_results}.

    \begin{table}[h]
    \centering
    \caption{Scalability results for a pool of $K=10$ expert modules served simultaneously, evaluated across 5 random seeds. Higher task scale increases distraction entropy, shifting the optimal momentum parameter to $\beta = 0.80$ to provide stronger low-pass filtering.}
    \label{tab:scale_results}
    \vskip 0.1in
    \begin{tabular}{ccc}
    \toprule
    \textbf{Momentum Coefficient} ($\beta$) & \textbf{Joint Acc. (\%)} & \textbf{Routing Jitter (MSE)} \\
    \midrule
    0.00 (Stateless) & 52.60\% & 0.084140 \\
    0.20 & 55.30\% & 0.056679 \\
    0.40 & 52.80\% & 0.036868 \\
    0.60 & 55.20\% & 0.021737 \\
    \textbf{0.80 (Optimal)} & \textbf{56.80\%} & \textbf{0.009719} \\
    1.00 (Static Uniform) & 48.80\% & 0.000000 \\
    \bottomrule
    \end{tabular}
    \end{table}

    Under a high-density 10-expert serving stream, we observe a profound shift in ensembling physics:
    \begin{itemize}
        \item \textbf{High-Dimensional Distraction Entropy:} As the task pool increases, the probability of random representation noise bleeding into orthogonal expert blocks scales rapidly. This is evidenced by stateless routing ($\beta=0$) suffering from extremely high routing jitter (\textbf{0.0841} for $K=10$ vs. \textbf{0.0720} for $K=4$).
        \item \textbf{Shift in Optimal Momentum ($\beta = 0.80$):} To filter out this higher-dimensional distraction noise, a standard stateless or lightly smoothed routing is insufficient. The optimal momentum coefficient shifts from $\beta = 0.60$ to \textbf{$\beta = 0.80$}, where Momentum-Merge achieves peak joint accuracy of \textbf{56.80\%} (outperforming stateless by \textbf{+4.20\%} absolute and static uniform by \textbf{+8.00\%} absolute) while bringing jitter down to \textbf{0.0097}.
    \end{itemize}
    This scalability sweep confirms that $\beta$ is not merely an empirical hyperparameter, but a physical inertia controller. Larger expert pools require a heavier low-pass filter to preserve stable representation trajectories, fully validating our parsimonious momentum formulation at scale.

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
