\begin{abstract}
While model merging has emerged as a powerful training-free paradigm for combining specialized expert models, deploying these models on resource-constrained edge hardware remains a major challenge due to their dense, full-sized parameter footprints. In this paper, we conduct a rigorous post-mortem and limitation-mapping study of joint model merging and pruning on edge hardware. We present \textbf{ZipMerge}, a framework designed to co-optimize layer-wise merging coefficients and magnitude-pruning boundaries at test-time using an unsupervised minimum entropy objective on tiny calibration sets, and through its extreme stress-test we expose the fundamental limits of both linear weight-space operations and adaptive test-time compression. We evaluate two optimization paradigms: (1) first-order gradient descent via a Straight-Through Estimator (STE), and (2) zero-order search via a 1+1 Evolution Strategy (ES). 

Through extensive evaluation of a compact Vision Transformer (\texttt{vit\_tiny\_patch16\_224}) backbone across four highly disparate visual classification tasks (MNIST, FashionMNIST, CIFAR-10, and SVHN), we expose the fundamental limits of both linear weight-space merging and test-time adaptive compression under extreme domain shift. Our experiments reveal that all merged configurations (including Uniform, AdaMerging, and ZipMerge) suffer from complete representation collapse, performing at the level of random guessing (10\% to 14\% accuracy). We show that the simple, unoptimized decoupled baseline, Prune-then-Merge (P-then-M), consistently outperforms joint test-time optimization under high task conflict because pre-merging pruning acts as a spatial regularizer that removes orthogonal parameter noise. 

Furthermore, we analyze how unconstrained unsupervised test-time adaptation on tiny calibration sets triggers the \emph{Overfitting-Optimizer Paradox}, where entropy minimization drives down training loss while destroying generalizable features. We translate these empirical boundary failures into actionable architectural guidelines for edge deployment, advocating for domain-aligned task selections, parameter-efficient adapters (PEFT), and explicit test-time regularizations.
\end{abstract}
