\begin{abstract}
As parameter-space model merging becomes a dominant paradigm for multi-task consolidation, a recent trend has emerged: introducing complex, highly over-engineered mathematical metaphors to achieve dynamic, input-dependent routing of task-specific weights. A prominent example is Quantum Wavefunction Superposition Merging (QWS-Merge), which models parameters as quantum eigenstates and employs wave phase-interference equations to prevent representation collapse. In this work, we adopt a critical, methodological lens to analyze and deconstruct QWS-Merge under a strictly fair and converged evaluation pipeline. Using a compact Vision Transformer (\texttt{vit\_tiny\_patch16\_224}) backbone trained to high convergence across MNIST, FashionMNIST, CIFAR-10, and SVHN, we make several key findings: (1) We clarify the paradigm comparison between offline-calibrated dynamic routing (QWS-Merge) and online Test-Time Adaptation (AdaMerging); while they achieve close joint accuracy ($83.97 \pm 0.53$\% vs.\ $89.30 \pm 0.00$\%), QWS-Merge operates as a pure forward pass with zero active optimization or latency overhead during inference. (2) We show that the reported "collapse" of classical linear routing is a pure methodological artifact of unregularized baseline optimization. Applying standard L2 regularization to a simple, classical Linear Router completely resolves the SVHN collapse, boosting SVHN accuracy to \textbf{91.73} $\pm$ \textbf{3.71}\% (outperforming QWS-Merge by \textbf{+12.00\%}) and increasing the joint mean accuracy from $78.13 \pm 2.54$\% to \textbf{82.80} $\pm$ \textbf{4.85}\%. (3) We identify a Softmax-based zero-sum competitive bottleneck during calibration and introduce the Softmax-free \textbf{BSigmoid-Router}, which replaces Softmax with independent Sigmoids to achieve a highly stable \textbf{83.73} $\pm$ \textbf{1.93}\% joint homogeneous and \textbf{83.96} $\pm$ \textbf{2.27}\% heterogeneous stream accuracy. (4) We show that our proposed \textbf{GLS-Router}, once controlled for layer-wise capacity, exhibits extreme sensitivity to calibration seeds (standard deviation of 24.30\% on SVHN). This reveals a major scientific finding: QWS-Merge's complex wave projection equations function as an essential structural regularizer that stabilizes weight-routing and prevents task-sacrificing behavior under extremely tight offline calibration budgets.
\end{abstract}
