\section{Introduction}
\label{sec:intro}

The paradigm of modular deep learning has emerged as a cornerstone of modern on-device artificial intelligence. By decoupling task-specific knowledge from a shared pre-trained backbone using Parameter-Efficient Fine-Tuning (PEFT) techniques, such as Low-Rank Adaptation (LoRA) \cite{hu2021lora}, practitioners can serve hundreds of concurrent tasks on resource-constrained edge hardware. In dynamic streaming environments, however, incoming inference requests are inherently heterogeneous, consisting of mixed-task inputs processed in arbitrary batch configurations. 

This setting introduces a severe deployment challenge. Standard static weight-space merging methods (e.g., Task Arithmetic \cite{ilharco2022editing}, TIES-Merging \cite{yadav2023ties}, or DARE \cite{yu2023dare}) collapse when forced to evaluate multiple distinct tasks simultaneously within a single vectorized forward pass—a phenomenon known as \textit{heterogeneity collapse}. To bypass this, state-of-the-art dynamic routing frameworks (such as Parameter-Free Subspace Routing (PFSR) \cite{pfsr2025} combined with Micro-Batch Homogenization (MBH) \cite{mbh2025}) partition incoming streams on-the-fly and route them to isolated expert parameters. While this preserves task-specific accuracy, it imposes a prohibitive linear latency penalty, scaling sequentially as $O(K)$ where $K$ is the number of active tasks. On resource-constrained edge CPUs, such linear scaling completely negates the throughput advantages of batch vectorization.

To resolve the latency bottleneck, Single-Pass Activation-Space Dynamic Blending (SPS) \cite{chatterjee2024robustness} executes the heavy pre-trained backbone exactly once, blending the lightweight adapter activations sample-wise inside a single forward pass, which restores a constant $O(1)$ latency. In parallel, Zero-Shot Centroid Alignment (ZCA) computes cosine similarity coordinates against task centroids pre-computed from a small calibration set at an early layer, resolving the late-stage routing paradox (which would otherwise require running the heavy model twice). 

Despite these empirical successes, we, as theorists, argue that these existing dynamic routers (e.g., SABLE, SPS-ZCA) suffer from three fundamental limitations:
\begin{enumerate}
    \item \textbf{Lack of Learning-Theoretic Foundations:} The temperature parameters $\boldsymbol{\tau}$ and scaling coefficients are typically set via empirical sweeps or grid searches. There are no formal, provably correct bounds to guarantee that these parameters will generalize to unseen, out-of-distribution, or temporally shifted streams.
    \item \textbf{Vulnerability to Heteroscedastic Noise:} In real-world multi-task deployments, representation spaces are highly heteroscedastic—different task manifolds exhibit vastly different spatial variances and noise profiles (e.g., clean digits in MNIST vs. highly cluttered natural scenes in SVHN). Standard cosine similarity-based routers scale down the signal components of high-noise tasks, leading to severe task confusion and routing collapse.
    \item \textbf{Representation Anisotropy and Fragmentation:} Deep feature representations are notoriously anisotropic, confined to a narrow, low-dimensional cone, and can exhibit extreme sparsity (representation fragmentation). Heuristic routers easily overfit to this noise when calibrated on small sample sets.
\end{enumerate}

To address these limitations, we propose \textbf{PAC-ZCA}, a mathematically rigorous, learning-theoretic framework for dynamic model merging. Our approach reformulates dynamic routing under the PAC-Bayesian learning framework \cite{mcallester1999pac}. We model the sample-wise routing decision as a randomized, strictly temperature-only Gibbs policy (Softmax routing) over task-agnostic, unsupervised Subspace Energy Projection (SEP) features. We define a Gaussian prior and posterior over the log-temperature parameters $\ln \boldsymbol{\tau}$ in parameter space, deriving a mathematically valid PAC-Bayesian bound on out-of-sample routing risk. We prove that minimizing the parameter-space KL-divergence acts as a powerful complexity penalty, bounding the router's Lipschitz constant and preventing deterministic overfitting by implicitly regularizing output routing entropy.

Rather than relying on empirical heuristics, PAC-ZCA dynamically solves for the optimal, task-specific temperature parameters $\boldsymbol{\tau}^*$ by minimizing our derived PAC-Bayesian generalization bound via differentiable gradient-based optimization over a tiny calibration split (16 samples per task). Our primary contributions are:
\begin{itemize}
    \item \textbf{Theoretical Foundation:} We establish the first mathematically sound PAC-Bayesian learning framework for dynamic model-merging routing, proving generalization bounds that link calibration sample complexity directly to out-of-sample routing error.
    \item \textbf{Parameter-Space Regularization:} We define our bound over the router's parameter space (the log-temperatures), bypassing data-dependency flaws and establishing a formal duality between parameter complexity and routing entropy.
    \item \textbf{Subspace Energy Projection:} We formulate Subspace Energy Projection (SEP), a task-agnostic, unsupervised early-layer dimensionality reduction step, and show how it generalizes to real-world, non-orthogonal distributed feature manifolds using Principal Component Analysis (PCA).
    \item \textbf{Empirical Vindication:} Under extreme heteroscedastic noise and representation fragmentation inside a 14-layer analytical Coordinate Sandbox, PAC-ZCA (under Block features) achieves \textbf{64.16\% $\pm$ 2.23\%} joint classification accuracy on decoupled calibration splits. This represents a massive \textbf{+23.70\%} improvement over standard SABLE on raw coordinates, and successfully stabilizes ensembling variance compared to standard unregularized Temp-Only ERM. Furthermore, on real images (MNIST, Fashion-MNIST, CIFAR-10) using a pre-trained ResNet-18 backbone, PAC-ZCA (under UN-PCA) achieves \textbf{70.87\% $\pm$ 2.20\%} joint accuracy, outperforming standard SABLE (65.67\%) and strictly outperforming standard unregularized ERM (69.47\% $\pm$ 2.21\%) while maintaining highly stable ensembling performance.
\end{itemize}
