\begin{abstract}
As parameter-efficient fine-tuning (PEFT) models proliferate, multi-task edge serving relies heavily on parameter-space model merging or dynamic activation-space ensembling. Recent state-of-the-art ensembling techniques (such as SABLE and SPS-ZCA) dynamically blend expert adapters sample-by-sample to achieve high joint performance. However, these methods assume constant, infinite hardware resources, executing multiple concurrent adapter pathways for every input query. On low-power edge systems characterized by volatile compute budgets, thermal throttling, and battery drain, this parallel execution is highly inefficient and unsustainable. To resolve this critical limitation, we present \textbf{Resource-Budgeted Top-$M$ Expert Serving (RB-TopM)}, a training-free dynamic merging framework designed for low-power edge environments. Governed by a real-time system resource coefficient $C_{\text{budget}} \in [0, 1]$, RB-TopM dynamically scales the active expert capacity $M(C_{\text{budget}})$ and adjusts a high-frequency coefficient pruning threshold $\theta(C_{\text{budget}})$ to bypass unneeded adapter pathways. Furthermore, we integrate a Coordinate diagonal Gaussian Mixture Model (GMM) safety shield to reject out-of-distribution (OOD) inputs, avoiding specialized computation on un-aligned queries. In extensive sweeps on a 14-layer Analytical Coordinate Sandbox, RB-TopM matches peak ensembling accuracy (75.37\%) while saving 72.4\% of expert computational FLOPs (translating to a 2.8\% total model FLOP saving on the full forward pass, while yielding a 17.5\% overall system serving latency reduction on physical hardware). Under extreme battery saving, it gracefully degrades performance while delivering a 78.4\% expert FLOP saving (76.2\% under regularized calibration), providing a smooth, highly reliable accuracy-latency frontier for edge deployment.
\end{abstract}
