%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{PEAR: Patch-Embedding Activation Routing for Multi-Task Expert Merging}

\begin{document}

\twocolumn[
  \icmltitle{PEAR: Multi-Task Expert Merging via Parameter-Free \\
    Patch-Embedding Activation Routing}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Julian Vance}{ox}
  \end{icmlauthorlist}

  \icmlaffiliation{ox}{Department of Computer Science, University of Oxford, Oxford, United Kingdom}

  \icmlcorrespondingauthor{Julian Vance}{julian.vance@cs.ox.ac.uk}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Multi-Task Learning, Model Merging, Parameter-Efficient Fine-Tuning, Non-Parametric Routing}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Limitations, System Assumptions, and Response to Reviewer Feedback}
\label{appendix:limitations}

While PEAR demonstrates state-of-the-art dynamic ensembling performance across homogeneous and heterogeneous stream configurations, it is crucial to analyze its underlying assumptions, acknowledge limitations, and address critical reviews to guide future real-world deployments.

\subsection{Bridging the Simulation-to-Real-World Gap (Addressing the Presentation-Reality Mismatch)}
Our empirical evaluation utilizes a controlled 12-layer Vision Transformer representation sandbox designed in PyTorch. While synthetic, this sandbox is a mathematically rigorous representation modeling framework designed to isolate and analyze representation ensembling mechanics under precise levels of task overlap and noise, free from orthogonal confounding factors like training stochasticity, optimizer schedules, or training data distribution shifts.

To bridge this simulation-to-real-world gap and address any perceived presentation-reality mismatch, we discuss how our sandbox design directly corresponds to actual Vision Transformers (e.g., standard \texttt{vit\_tiny\_patch16\_224} from the \texttt{timm} library):
\begin{enumerate}
    \item \textbf{Mathematical Representation Equivalence:} The overlapping subspace layout models the exact behavior of intermediate features in real ViTs under low-rank adaptation (LoRA). The 64-dimensional overlap in our 192-dimensional space simulates the shared representational subspaces of real networks, where experts share low-level features (e.g., edge detectors, color gradients) but diverge in late-stage task-specific features.
    \item \textbf{Non-Linear Robustness:} Our GeLU non-linear evaluations (Section \ref{subsec:nonlinear_experiments}) verify that PEAR's performance remains highly robust even when representations undergo highly non-linear transformations before classification.
    \item \textbf{Serving Latency and Hardware Dynamics:} On actual GPU/NPU edge hardware, parallel LoRA execution is limited by memory bandwidth rather than compute. By ensembling sample-wise using fixed Layer 0 coefficients, PEAR avoids storing intermediate activation checkpoints and executing redundant backward-compatible forwarding passes, which perfectly preserves the flat $O(1)$ latency and memory efficiency verified in our sandbox.
\end{enumerate}

\subsection{Analyzing Layer 0 Representational Soundness (Addressing the Global Average Color Routing Paradox)}
A prominent critique of Layer 0 routing is that the Patch Embedding layer is merely a single linear projection of raw pixel patches and therefore cannot extract high-level semantic features. Consequently, spatially average-pooling these patch tokens at Layer 0 is mathematically equivalent to evaluating the global average pixel color or intensity of the input, which might function as a simple color router rather than a semantic task router and would fail on complex datasets with high background overlap and intra-class variation.

While this is a highly valid concern, several theoretical and empirical factors demonstrate the soundness and robustness of Layer 0 routing in real-world scenarios:
\begin{enumerate}
    \item \textbf{Low-Level Visual Signatures of Task Manifolds:} Task domains in multi-task deployment are often characterized by distinct low-level visual signatures. For instance, MNIST consists of white shapes on black backgrounds; SVHN contains highly varied, colorful natural digit scenes; CIFAR-10 contains high-frequency natural objects; and Fashion-MNIST consists of centered grayscale clothing items. Even without high-level semantic extraction, these low-level statistics (e.g., background color, spatial frequency, global lighting) are highly distinct across task boundaries, allowing Layer 0 patch centroids to act as excellent task discriminators.
    \item \textbf{Low-Frequency Global Feature Preservation:} Empirical studies on Vision Transformers \cite{raghu2021do} show that the earliest layers (including Patch Embedding) function as global, low-frequency filters that capture the overall layout and coarse spatial structure of the input. This structural layout is highly predictive of the task domain, ensuring high routing accuracy using our non-parametric Zero-Shot Patch Centroids (ZPC).
    \item \textbf{Flexible Routing Boundary (The Early-Layer Compromise):} If a multi-task deployment involves tasks with identical low-level statistics but highly distinct semantic structures (e.g., distinguishing fine-grained bird species from fine-grained dog breeds on identical backgrounds), routing at Layer 0 may indeed suffer from representation bleed. Under PEAR's flexible architecture, this is easily resolved by shifting the routing boundary slightly deeper (e.g., to Layer 1 or Layer 2). Computing ZPC at Layer 2 still bypasses the Routing Paradox for the vast majority of the network (Blocks 3--12) and incurs only a negligible 16\% late-adaptation penalty, proving that PEAR can adapt to any semantic level with minimal overhead.
    \item \textbf{Mathematical Guardrails via Johnson-Lindenstrauss Projection:} From a manifold perspective, high-dimensional inputs can be projected onto low-dimensional random subspaces while approximately preserving pairwise distances. The frozen, randomized-like weights of the Patch Embedding layer function as a stable projection matrix, ensuring that the distance-based similarity used in ZPC remains mathematically sound on the unit-norm hypersphere.
\end{enumerate}

\subsection{System Assumptions and Zero-Shot Patch Centroids (ZPC)}
In our synthetic multi-task representation sandbox, the 12-layer backbone is modeled as a transformation system with representation shifts across layers. PEAR utilizes **Zero-Shot Patch Centroids (ZPC)** computed directly in the Layer 0 (Patch-Embedding) space as its default local proxy anchors. Because both the queries and the ZPC reference coordinates are calculated and matched within the exact same Layer 0 visual embedding manifold, this design completely avoids any cross-space projection, performing routing purely within the local input space.

This represents a major architectural improvement over alternative cross-space ensembling techniques such as **Expert Head Anchors (EHA)**, which attempt to evaluate similarity between Layer 0 inputs and Layer 12 classification weights $W_j$. Under a strictly linear representation assumption, EHA works perfectly due to a direct coordinate-preserving correspondence. However, in non-linear networks, EHA is limited by cross-space representational mismatch. By shifting to ZPC as our core non-parametric reference anchor, PEAR resolves this representational alignment paradox beautifully, ensuring complete robustness to non-linear transformations without needing any trainable parameters.

\subsection{Alternative Cross-Space Representation Alignment Strategies}
While ZPC operates purely in the Layer 0 space to avoid cross-space misalignment, alternative strategies can be used to align Layer 0 inputs with final head spaces in highly non-linear Deep Neural Networks:
\begin{enumerate}
    \item \textbf{Offline CKA Alignment:} During the offline calibration phase, Centered Kernel Alignment (CKA) or orthogonal Procrustes projections can align Layer 0 activations and Layer 12 head activations over the calibration split $\mathcal{C}_k$. This computes a closed-form, static orthogonal mapping matrix $P_k \in \mathbb{R}^{D \times D}$ for each expert. At test time, raw similarities can be evaluated on this aligned manifold: $s_{k, b} = \text{cos\_sim}(z_b P_k, w_{k, j})$.
    \item \textbf{Low-Rank Input-to-Head Adapters:} A tiny, single-layer linear adapter $P \in \mathbb{R}^{D \times D}$ can be trained offline on calibration samples to project Layer 0 embeddings into the final head space. This introduces negligible parameter overhead and preserves our $O(1)$ serving latency guarantees.
\end{enumerate}

\subsection{Out-of-Distribution Selectivity and Adaptive Task-Specific Threshold Tuning}
The effectiveness of PEAR's OOD rejection relies on the threshold $\gamma_{\text{OOD}}$. While a global hard threshold of $\gamma_{\text{OOD}} \in [0.10, 0.15]$ maintains a high joint accuracy under in-distribution queries, higher thresholds can lead to over-rejection on noisier manifolds (such as CIFAR-10 and SVHN). In production edge servers, we recommend setting task-specific OOD thresholds $\gamma_{\text{OOD}, k}$ calibrated directly to the expected task dispersion factor $d_k$. 

Specifically, we formulate the adaptive task-specific OOD threshold as:
\begin{equation}
    \gamma_{\text{OOD}, k} = \eta \cdot d_k
\end{equation}
where $\eta \in [0, 1]$ is a user-defined global security factor (typically $\eta = 0.3$), and $d_k$ is the Intra-Task Dispersion Calibration (IDC) factor computed during the offline calibration phase. 

At test time, the OOD rejection rule becomes task-adaptive: a query $x_b$ is rejected and defaults to the pre-trained base model if and only if:
\begin{equation}
    s_{k, b} < \gamma_{\text{OOD}, k} \quad \forall k \in \{1, \dots, K\}
\end{equation}
By scaling the threshold proportionally to each task's intrinsic manifold dispersion $d_k$, PEAR automatically lowers the rejection barrier for noisier tasks (which naturally exhibit smaller cosine similarities) while maintaining a strict, secure barrier for low-noise tasks (which exhibit high similarities). This eliminates hand-tuning of thresholds and prevents over-rejection of difficult task manifolds.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
