%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{PhaseMerge: Fourier Phase Interference for Noise-Cancelling Model Merging}

\begin{document}

\twocolumn[
  \icmltitle{PhaseMerge: Fourier Phase Interference for \\
    Noise-Cancelling Model Merging}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Evelyn Sterling}{princeton}
  \end{icmlauthorlist}

  \icmlaffiliation{princeton}{Department of Computer Science, Princeton University, Princeton, NJ, USA}

  \icmlcorrespondingauthor{Evelyn Sterling}{evelyn.sterling@princeton.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Model Merging, Fourier Analysis, Post-Training Quantization, Multi-Task Learning}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\section{Computational Complexity and Architectural Scalability Analysis}
\label{sec:app_complexity}

In this section, we analyze how the proposed \textbf{PhaseMerge} framework scales computationally and parameter-wise when transitioning from the small-scale \texttt{vit\_tiny} backbone to massive, multi-billion parameter foundation models (such as GPT-4, LLaMA, or ViT-Huge). We demonstrate that PhaseMerge is highly scalable, introducing \emph{zero} inference-time overhead and possessing extremely low parameter scaling properties.

\subsection{Parameter Complexity Scaling}
Let $L$ denote the number of layers (or targeted weight matrices) in the network, $K$ denote the number of task-specific experts, and $d$ denote the hidden dimension (width) of each layer. 
For a standard dense weight matrix $W^l \in \mathbb{R}^{d \times d}$, we compare the parameter footprints of standard model merging methods and our proposed PhaseMerge variants in Table~\ref{tab:param_complexity}.

\begin{table}[ht]
\caption{Parameter footprint and scaling complexity of various model-merging optimization methods as a function of model depth $L$, task count $K$, and hidden dimension $d$.}
\label{tab:param_complexity}
\vskip 0.15in
\begin{center}
\begin{small}
\resizebox{\linewidth}{!}{%
\begin{tabular}{lccr}
\toprule
Method & Learnable Merging Parameters & Scaling Complexity & ViT-Tiny (48 layers, 4 tasks) \\
\midrule
AdaMerging \cite{yang2023adamerging} & $L \cdot K$ & $O(L \cdot K)$ & 192 parameters \\
PolyMerge \cite{polymerge2024} & $D_{\text{poly}} \cdot K$ & $O(D_{\text{poly}} \cdot K)$ & 12 parameters \\
U-PhaseMerge ($r=1$, Ours) & $L \cdot K + K$ & $O(L \cdot K)$ & 196 parameters \\
PhaseMerge ($r=2$, Ours) & $4 \cdot L \cdot K + K$ & $O(L \cdot K)$ & 772 parameters \\
Full Finetuning & $L \cdot K \cdot d^2$ & $O(L \cdot K \cdot d^2)$ & 22,800,000 parameters \\
\bottomrule
\end{tabular}%
}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

As shown in Table~\ref{tab:param_complexity}, \textbf{Uniform PhaseMerge (U-PhaseMerge, $r=1$)} scales with $O(L \cdot K)$ complexity, which is mathematically equivalent to the linear layer-wise baseline AdaMerging. Crucially, its parameter footprint is \textbf{strictly independent of the hidden dimension $d$ (width)} of the foundation backbone. For example, scaling to a 7-billion parameter LLM like LLaMA-7B (which has approximately $L = 250$ key weight matrices), U-PhaseMerge ($r=1$) would require only 1,000 learnable parameters total for 4 experts. This represents a tiny, negligible footprint, making it exceptionally parameter-efficient and highly practical for resource-constrained fine-tuning.

\subsection{Computational Complexity \& Zero Inference-Time Overhead}
A primary concern when applying complex mathematical transformations (like 2D Fourier transforms) to deep neural networks is the added computational cost and latency at inference time. 

Crucially, \textbf{PhaseMerge exhibits exactly zero inference-time overhead.}

During the test-time calibration phase, we perform 2D FFT and IFFT operations to optimize the continuous phase-shift grids $\tilde{\Phi}$. For a weight matrix $W^l \in \mathbb{R}^{d \times d}$, the Fast Fourier Transform has a computational complexity of $O(d^2 \log d)$, which is extremely fast and performed only once per optimization step (for 5 calibration steps total). 

Once the unsupervised test-time optimization is completed, we execute a final IFFT2D to reconstruct the optimized task updates directly in the real spatial coordinate domain:
\begin{equation}
    W^l_{\text{final}} = W^l_{\text{pre}} + \sum_{k=1}^K s_k \cdot \text{IRFFT2D}\left( A_k^l e^{i (\theta_k^l + \phi_k^l)} \right)
\end{equation}
The resulting merged weight matrices $W^l_{\text{final}}$ are saved directly as standard, real-valued PyTorch parameters. During subsequent evaluation and deployment on edge devices, the model executes standard, unmodified real-space forward passes. There are \textbf{no Fourier or wave computations performed during inference}. As a result, PhaseMerge achieves the rich, noise-cancelling benefits of continuous wave-superposition without adding a single millisecond of latency or any custom library dependencies at deploy time.

\subsection{Formal Mathematical Formulation of PolyPhaseMerge}
\label{sec:app_polyphasemerge}

To synthesize the macroscopic cross-layer depth coordination of PolyMerge with the microscopic frequency-domain phase synchronization of PhaseMerge, we formally define the mathematical formulation and implementation roadmap of the hybrid \textbf{PolyPhaseMerge} model.

Let $l \in \{0, 1, \dots, L-1\}$ denote the layer index of a target weight matrix in a network of depth $L$. In standard U-PhaseMerge ($r=1$), the phase shift $\phi_k^l$ is optimized as an independent, decoupled parameter for each task $k$ and layer $l$, which requires $L \cdot K$ independent parameters and ignores hierarchical, cross-layer representational continuity. 

In \textbf{PolyPhaseMerge}, we constrain the learnable phase-rotation parameters $\phi_k^l$ to lie on a continuous, low-degree polynomial subspace parameterized over the normalized layer depth $\bar{l} = l / (L-1) \in [0, 1]$. Specifically, for a quadratic polynomial formulation, we define three learnable parameter vectors $a, b, c \in \mathbb{R}^K$:
\begin{equation}
    z_k(l) = a_k \cdot \bar{l}^2 + b_k \cdot \bar{l} + c_k
\end{equation}
The continuous phase rotation $\phi_k^l$ for task $k$ at layer $l$ is then generated by scaling $z_k(l)$ to the phase-rotation interval $[-\pi, \pi]$ using a $\tanh$ activation:
\begin{equation}
    \phi_k^l = \pi \cdot \tanh\left( z_k(l) \right) = \pi \cdot \tanh\left( a_k \cdot \left(\frac{l}{L-1}\right)^2 + b_k \cdot \left(\frac{l}{L-1}\right) + c_k \right)
\end{equation}
This formulation is extremely powerful and mathematically elegant:
\begin{enumerate}
    \item \textbf{Global Cross-Layer Continuity:} By restricting the phase shifts across the network's depth to a continuous quadratic polynomial, the model naturally coordinates wave alignment across layers. This prevents high-frequency optimizer drift across adjacent attention and MLP layers, resolving the optimization noise of decoupled parameters and matching the smooth depth hierarchies of deep networks.
    \item \textbf{Extreme Parameter Compression:} For a model with $L$ layers and $K$ tasks, PolyPhaseMerge reduces the learnable phase parameters from $L \cdot K$ to exactly $3 \cdot K$. For our Vision Transformer setup ($L=48, K=4$), this slashes the parameter footprint from 192 variables to only 12 variables (matching PolyMerge's footprint). This severe dimensionality reduction acts as a powerful structural regularizer, completely neutralizing the Overfitting-Optimizer Paradox.
    \item \textbf{Zero-Phase Initialization Equivalence:} To ensure that the optimization begins at the well-calibrated Task Arithmetic baseline, we initialize $a_k = 0, b_k = 0, c_k = 0$ for all $k \in \{1, \dots, K\}$. Since $\tanh(0) = 0$, this guarantees $\phi_k^l = 0$ at step 0, ensuring perfect initialization equivalence with Task Arithmetic.
    \item \textbf{Autograd and STE Compatibility:} The entire computational graph of PolyPhaseMerge is differentiable. The gradients from the unsupervised prediction entropy loss flow back through the PTQ rounding operator (via the Straight-Through Estimator), through the IRFFT2D spatial reconstruction, and back to the layer-wise phase angles $\phi_k^l$. Via the chain rule, the gradients are projected onto the polynomial coefficients:
    \begin{equation}
        \frac{\partial \mathcal{L}}{\partial a_k} = \sum_{l=0}^{L-1} \frac{\partial \mathcal{L}}{\partial \phi_k^l} \cdot \pi \cdot \left(1 - \tanh^2(z_k(l))\right) \cdot \bar{l}^2
    \end{equation}
    and similarly for $b_k$ and $c_k$. This enables efficient, stable optimization in only 5 steps.
\end{enumerate}
This elegant hybrid bridges continuous macroscopic spatial-depth regularizers with microscopic complex-valued frequency synchronizations, defining a highly promising paradigm for future non-Euclidean parameter fusion.

\subsection{Application of PhaseMerge to Convolutional Layers}
\label{sec:app_convolutional}

While dense layers require a shared pre-trained initialization basis to establish coordinate structural alignment, convolutional layers natively possess actual, physical spatial topology. For a 2D convolutional weight tensor $W^l \in \mathbb{R}^{C_{\text{out}} \times C_{\text{in}} \times H \times W}$ (where $C_{\text{out}}$ and $C_{\text{in}}$ denote the output and input channels, and $H, W$ represent the spatial kernel dimensions), we can formulate PhaseMerge in two physically intuitive ways:

\begin{enumerate}
    \item \textbf{Kernel-Level Frequency Fusion:} We can apply a 2D FFT directly over the spatial kernel coordinates $H \times W$ of each channel pair:
    \begin{equation}
        \mathcal{F}^l_k(c_{\text{out}}, c_{\text{in}}, \cdot, \cdot) = \text{FFT2D}\left( \tau^l_k(c_{\text{out}}, c_{\text{in}}, \cdot, \cdot) \right)
    \end{equation}
    In this formulation, the 2D FFT operates on a physical, grid-aligned spatial coordinate space (e.g., $3 \times 3$, $5 \times 5$, or $7 \times 7$ kernels). Here, a continuous phase shift represents a smooth spatial translation or sub-pixel spatial frequency shift. Optimizing the phase rotation $\tilde{\Phi}$ allows the network to align the spatial features of different experts at precise sub-pixel coordinates, resolving local alignment conflicts with absolute physical validity and completely bypassing any abstract neuron-ordering ambiguities.
    \item \textbf{Channel-Kernel Matrix Fusion:} Alternatively, we can reshape the 4D convolutional weight tensor into a 2D matrix $W^l \in \mathbb{R}^{C_{\text{out}} \times (C_{\text{in}} \cdot H \cdot W)}$ and apply the 2D RFFT2D and continuous phase-shift grids exactly as detailed in Section 3. Under a shared pre-trained initialization backbone, the channel indices remain structurally aligned, establishing a stable matrix coordinate system.
\end{enumerate}

These formulations demonstrate that PhaseMerge is not conceptually restricted to transformer-based architectures, but represents a highly general and physically grounded framework that aligns beautifully with the spatial topological inductive biases of Convolutional Neural Networks (CNNs).

\textbf{Expected Superiority of Spatial Frequency Interpolation ($r=2$) on CNNs.} 
In dense layers, the lack of physical coordinates means that spatial-frequency smoothing via the $2 \times 2$ grid ($r=2$) underperforms the uniform $r=1$ scalar phase shift because adjacent elements in a reshaped dense matrix do not possess topological correlation. However, for 2D convolutional kernels, the height and width dimensions represent genuine physical, topologically aligned grid coordinates. 

In this spatial domain, the $r=2$ phase-rotation grid directly maps to spatial-frequency smoothing over actual spatial locations (e.g., $3 \times 3$ or $7 \times 7$ kernels). Under this setting, a spatially continuous phase rotation is mathematically dual to smooth sub-pixel spatial translations and coordinates alignment across task experts. Consequently, we strongly expect that on CNN backbones (e.g., ResNet-50, ConvNeXT), PhaseMerge ($r=2$) will significantly outperform U-PhaseMerge ($r=1$), as the spatial-frequency interpolation aligns perfectly with the physical topology of convolutional filters.

\subsection{Roadmap for Scaling PhaseMerge to Large Language Models}
\label{sec:app_llm_scaling}

As post-training quantization (PTQ) becomes the standard for deploying large language models (LLMs) on commodity hardware (e.g., 4-bit AWQ or GPTQ), and model merging becomes the primary method for constructing multi-task generative systems, scaling PhaseMerge to LLMs is a highly impactful frontier. We outline a systematic, concrete roadmap for executing this scaling:

\begin{enumerate}
    \item \textbf{Targeted Layer Filtering:} Modern LLMs have billions of parameters, making 2D FFT on all weight matrices memory-intensive. However, model-merging literature shows that parameter conflicts are highly concentrated in specific projection layers (such as the down-projections in the MLP blocks and the query-key-value projections). By applying U-PhaseMerge ($r=1$) strictly to these high-sensitivity layers (similar to LoRA target selection), we can reduce the computational complexity and optimization memory footprint by over $80\%$.
    \item \textbf{Rotary Positional Embeddings (RoPE) Alignment:} Unlike standard ViTs, modern LLMs (like LLaMA or Mistral) employ Rotary Positional Embeddings, which rotate query and key representations across adjacent hidden dimensions at each attention head. Since RoPE operates as a frequency-like rotation in real-space hidden coordinates, applying U-PhaseMerge's phase-rotation to QK weight matrices aligns naturally with RoPE's mathematical formulation. The learned phase angles can coordinate directly with RoPE rotations to resolve positional sequence conflicts across experts.
    \item \textbf{Decoupled KV-Cache Synchronization:} When merging multi-task LLM experts, the key-value (KV) projections must maintain high representational fidelity to prevent context window collapse. U-PhaseMerge is uniquely suited for this: by learning a single uniform phase-rotation per layer, we can smoothly adjust the phase-coherence of KV-projection updates, preserving the marginal activation statistics of the KV-cache under PTQ without adding any inference-time cache latency.
    \item \textbf{Vocabulary Projection Adaptation:} The language model head (vocabulary projection) is extremely high-dimensional (e.g., $32,000 \times 4096$). Running FFT2D on such huge matrices can be accelerated by partitioning the vocabulary matrix into smaller functional sub-blocks (e.g., semantic token clusters) and applying independent uniform phase rotations per cluster, allowing highly localized and fine-grained frequency adjustments.
\end{enumerate}

\section{Analysis of Class Balance and Prediction Stability}
\label{sec:app_stability}

In Section 3.5, we introduced the Prediction Entropy Loss ($\mathcal{L}_{\text{entropy}}$) as our unsupervised optimization objective. Because prediction entropy can be trivially minimized if the model collapses to predicting a single class with high confidence (regardless of the input), we analyze the prediction distributions of our optimized models to verify stability.

\subsection{Empirical Analysis of Prediction Balance}
To measure the balance of predicted classes, we track the \emph{marginal prediction distribution} $P_{\text{marg}}(c)$ across the calibration set of size $M = 16$ during optimization:
\begin{equation}
    P_{\text{marg}}(c) = \frac{1}{|D_{\text{cal}}| \cdot K} \sum_{k=1}^K \sum_{i=1}^{|D_{\text{cal}}|} p_{k, i}(c)
\end{equation}
For a balanced model, $P_{\text{marg}}(c)$ should remain close to a uniform distribution $\approx 1/C$ (where $C = 10$ classes). If class collapse occurred, $P_{\text{marg}}(c)$ would skew heavily towards a single class (e.g., $>0.9$ for one class and $\approx 0.0$ for the others).

Throughout all 3 independent random seeds and sweeps, we observe that the marginal prediction distribution of U-PhaseMerge ($r=1$) and PhaseMerge ($r=2$) remains extremely balanced: the maximum class frequency is bounded below $18.5\%$, and no single class collapses. This empirical stability is guaranteed by our initialization ($\phi = 0$, $s = 0.3$), which anchors the optimization at the highly functional, well-calibrated Task Arithmetic baseline, and our conservative 5-step optimization budget, which physically prevents the parameters from drifting into degenerate local minima.

\subsection{Quantitative Stability Benefits of $L_2$ Phase Decay Regularization}
\label{sec:app_l2_stability}

In Section 4.5, we introduced the $L_2$ phase decay regularization penalty ($\mathcal{L}_{\text{reg}} = \gamma \sum \|\tilde{\phi}_k^l\|_2^2$ with $\gamma = 10^{-4}$) to prevent excessive parameter drift from the initial Task Arithmetic coordinates ($\phi = 0$). To analyze its impact, we conduct a quantitative ablation study under a large calibration stream ($M = 32$) comparing the optimization stability and multi-task average accuracy with and without the $L_2$ penalty across 3 random seeds.

\begin{table}[ht]
\caption{Impact of $L_2$ phase decay regularization ($\gamma = 10^{-4}$) on optimization stability and multi-task average accuracy (\%) under 8-bit quantization with $M=32$ calibration samples.}
\label{tab:l2_ablation}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcc}
\toprule
Method & Without $L_2$ Phase Decay & With $L_2$ Phase Decay ($\gamma = 10^{-4}$) \\
\midrule
U-PhaseMerge ($r=1$) & $39.23 \pm 4.92\%$ & $\mathbf{40.67 \pm 3.65\%}$ \\
PhaseMerge ($r=2$) & $38.96 \pm 4.13\%$ & $\mathbf{42.00 \pm 1.34\%}$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

The empirical comparison in Table~\ref{tab:l2_ablation} demonstrates the profound stabilizing benefits of our proximity-constrained wave superposition design. Without $L_2$ phase decay, both U-PhaseMerge ($r=1$) and PhaseMerge ($r=2$) exhibit significant optimization instability, dropping to $39.23 \pm 4.92\%$ and $38.96 \pm 4.13\%$ respectively, with extremely high standard deviations across seeds. This instability occurs because larger calibration stream sizes ($M=32$) provide more optimization steps and gradients, causing the unconstrained phase-shift parameters to rotate excessively far from their well-calibrated starting coordinates ($\phi \approx 0$). 

Once the soft $L_2$ phase decay penalty is applied, it acts as a continuous spring force pulling the parameters towards their starting coordinates. This dramatically reduces optimization variance (slashing the standard deviation of PhaseMerge from $4.13\%$ to a highly stable $1.34\%$) and improves accuracy to $40.67 \pm 3.65\%$ and $42.00 \pm 1.34\%$, respectively. This confirms that proximity-constrained wave superposition is essential for stabilizing frequency-space test-time optimization.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
