\section{Conclusion and Future Work}
\label{sec:conclusion}

In this work, we introduced \textbf{ChebyMerge} (Stable and Optimal Continuous Subspace Model Merging), a mathematically rigorous framework designed to resolve the limitations of existing dynamic model-merging techniques. We exposed the \emph{Overfitting-Optimizer Paradox}, where unconstrained test-time model merging (e.g., AdaMerging) suffers from severe representation collapse due to transductive overfitting on high-frequency sampling noise. We further identified that standard continuous subspace approaches (e.g., PolyMerge) suffer from severe numerical ill-conditioning because their standard monomial basis yields Gram matrices with condition numbers that scale exponentially as $\mathcal{O}(4^d)$.

By projecting the spatial layer-wise merging coefficients onto an orthogonal subspace spanned by Chebyshev polynomials of the first kind, ChebyMerge achieves:
\begin{enumerate}
    \item \textbf{Continuous, Minimax-Optimal Uniform Approximation}: Ensuring bounded representation error under the supremum norm ($L_\infty$) for smooth sensitivity profiles.
    \item \textbf{Exceptional Numerical Conditioning}: Bounding the condition number of the optimization Gram matrix to a tiny constant ($\approx 2.95$ for cubic degree), representing up to a \textbf{3,527$\times$ improvement} over standard monomials, which enables smooth, isotropic, and stable gradient flow.
    \item \textbf{Implicit Boundary Sensitivity Matching}: Naturally concentrating representational resolution at the highly sensitive early and deep layers of neural networks.
\end{enumerate}

Extensive evaluations across four datasets and thirty independent random seeds under highly non-convex, coupled Rastrigin stress-tests confirmed that ChebyMerge completely averts representation collapse, filters out transductive local noise, and consistently stabilizes on-the-fly optimization. 

\subsection{Future Directions: Piecewise Cubic B-Spline Manifolds}
As deep learning scales toward ultra-deep Large Language Models (LLMs) with $L \ge 32$, $80$, or even hundreds of layers, representing the entire model via a single global polynomial can lead to numerical stiffness or require excessively high degrees $d$ that compromise continuous noise-filtering properties. 

In future work, we plan to construct \textbf{Piecewise Continuous Cubic B-Spline Manifolds} with continuous first and second derivative ($C^2$) boundary conditions. By dividing ultra-deep networks into localized parameter blocks and parameterizing each block via a localized Chebyshev spectral projection, we can model sharp, localized layer transitions while strictly limiting the global search dimension. This hybrid formulation will guarantee both global smoothness and local expressivity, paving the way for scalable, theoretically grounded, and numerically flawless multi-task consolidation in foundation models.
