\begin{abstract}
Dynamic model ensembling serves parameter-efficient adapters (e.g., LoRAs)
dynamically on resource-constrained edge devices to handle heterogeneous
streaming tasks. While stateful ensembling smooths routing weight trajectories
across depth, it relies on heuristic constant feedback step sizes ($\eta$)
to warp intermediate representations toward static early-stage centroids. Under
mixed workloads, this causes a representational backward-shift that pulls
refined late-layer activations backward toward noisy early-stage coordinates,
degrading performance. This paper introduces \textbf{Lyapunov-Stable Active
Representation Coupling (L-ARC)}, a training-free closed-loop control system
that provides formal control-theoretic guarantees for continuous-depth feature
propagation. We model the representation similarity error as a system-level
candidate Lyapunov function and analytically derive a local \textbf{Dissipation
Guard} that computes sample-specific, layer-specific feedback rates on-the-fly,
ensuring that representation warping is strictly error-decreasing (dissipative).
To eliminate computational overhead, we propose \textbf{Entropy-Triggered
Gating (ET-L-ARC)}, which dynamically evaluates the Dissipation Guard only under
routing uncertainty. Under Static Centroids (Setting A), L-ARC achieves a Joint
Mean Accuracy of \textbf{74.38\% $\pm$ 0.31\%} and Semantic Similarity of
\textbf{0.7937 $\pm$ 0.0059}, while collapsing the relative latency overhead to
only \textbf{99.85\%} ($0.06$ ms of absolute overhead per sample). Under
transient routing failures (Setting C), L-ARC maintains an outstanding
\textbf{73.97\% $\pm$ 0.39\%} accuracy and \textbf{0.7813 $\pm$ 0.0075} similarity,
driven by our control-theoretic \textbf{Entropy-Gated Concentration Gating
(ECG-Reset)} state-space shield. Finally, under Confident Router Bias (Setting D),
we propose \textbf{Representation-Agreement State Correction (RASC)} to resolve
the state-locking failure of stateful ensembling, achieving a stellar accuracy of
\textbf{73.59\% $\pm$ 0.39\%} and Semantic Similarity of \textbf{0.7467 $\pm$
0.0082}, outperforming previous heuristics. This establishes L-ARC as an
exceptionally robust and mathematically sound framework for dynamic serving on
resource-constrained edge hardware.
\end{abstract}
