\begin{abstract}
Weight-space model merging has emerged as a powerful paradigm for combining specialized neural network experts without the prohibitive costs of multi-task retraining. However, existing approaches operate strictly in real-valued, static Euclidean space, blending parameter matrices via linear interpolation. This static assumption makes merged models highly vulnerable to destructive parameter interference and the ``Overfitting-Optimizer Paradox'' during test-time calibration. Furthermore, these configurations can degrade under post-training quantization (PTQ) schema shifts, as optimized weights overfit to discrete rounding boundaries. In this work, we present a novel paradigm: \textbf{PhaseMerge}, a frequency-domain model merging framework that projects task-specific parameter updates into complex-valued Fourier space. By representing fine-tuning updates in Fourier space, we introduce a differentiable phase-rotation mechanism. We explore two parameterizations: a uniform scalar phase-shift per layer ($r=1$), representing U-PhaseMerge, and a spatial bilinear continuous phase-shift grid ($r=2$), representing PhaseMerge, both constrained by a strict symmetry-preserving frequency mask. Operating in a shared coordinate basis established by a common pre-trained initialization backbone, these methods act as frequency-domain matrix-basis regularizers. They enable constructive reinforcement of aligned task features and active phase-cancellation of conflicting parameter updates and high-frequency quantization noise. Systematic evaluations on Vision Transformers across conflicting vision tasks demonstrate that PhaseMerge significantly outperforms static baselines under FP32 and PTQ regimes. Crucially, the uniform $r=1$ variant exhibits exceptional robustness under extreme 4-bit quantization. Our work shifts the model-merging paradigm from rigid real-space averaging to frequency-domain phase optimization.
\end{abstract}