\begin{abstract}
Parameter-Efficient Fine-Tuning (PEFT) has enabled the efficient deployment of task-specific expert adapters, yet dynamically ensembling these experts on a heterogeneous, sample-by-sample serving stream remains a fundamental challenge. State-of-the-art stateful routing systems, such as ChemMerge, model representational trajectories using complex systems of non-equilibrium biochemical kinetics, Arrhenius reaction rates, and continuous-time Ordinary Differential Equations (ODEs). This introduces high system-level complexity, virtual-time discretization, and multiple uninterpretable hyperparameters. In this paper, we deconstruct this complex physical metaphor through the lens of Occam's razor. We mathematically demonstrate that ChemMerge's stateful kinetics are equivalent to a simple Exponential Moving Average (EMA) under standard discretization. Guided by this minimalist perspective, we propose \textbf{Momentum-Merge}, a training-free and single-parameter constant EMA update on ensembling weights across network depth. Evaluating on a heterogeneous serving stream within the Analytical Coordinate Sandbox (ICS), we expose a fundamental Accuracy-Stability trade-off: stateless calibrated routing (SABLE + Layer Centroids) achieves the highest joint accuracy (\textbf{77.24\%}), but suffers from massive layer-to-layer routing weight oscillations (routing jitter). Adding stateful smoothing acts as a low-pass filter that trades a small fraction of accuracy to achieve high routing stability. Under this trade-off, our proposed basic Momentum-Merge achieves a joint classification accuracy of \textbf{74.85\%} and reduces layer-to-layer routing jitter to \textbf{0.012860} (a \textbf{5.7$\times$} reduction over SABLE). Furthermore, our advanced variant reaches \textbf{74.98\%} joint accuracy while virtually eliminating routing oscillations, reducing jitter to a near-zero \textbf{0.000374}---representing a staggering \textbf{195.7$\times$} reduction over tuned SABLE and \textbf{41.1$\times$} over tuned ChemMerge SOTA. Our results demonstrate that a simple constant momentum filter is mathematically and empirically superior to convoluted physical kinetics for trajectory smoothing, establishing that in dynamic model merging, simpler is strictly better for ensembling stability.
\end{abstract}
