\section{Related Work}
\label{sec:related_work}

\subsection{Parameter-Efficient Fine-Tuning and Model Merging}
Parameter-Efficient Fine-Tuning (PEFT), such as Low-Rank Adaptation (LoRA) \cite{hu2021lora}, Prefix-Tuning \cite{li2021prefix}, and Prompt-Tuning \cite{lester2021power}, adapts massive pre-trained language and vision models to downstream tasks by tuning only a tiny fraction of parameters. To serve multiple task adapters simultaneously, static weight merging has been widely explored. Methods like FedAvg \cite{fedavg}, Model Soups \cite{wortsman2022soup}, Task Vectors \cite{taskvectors}, Ties-Merging \cite{tiesmerging}, and DARE \cite{dare} merge adapter weights directly into the base model. Though effective for cooperative tasks, static merging suffers from representation blurring and parameter interference under heterogeneous conditions, causing performance collapse \cite{gitrebasin}.

\subsection{Dynamic Routing and Mixture of Experts}
To circumvent static merging limits, dynamic ensembling structures have been proposed. Inspired by Mixture of Experts (MoE) \cite{shazeer2017outrageously, fedus2022switch, lepikhin2020gshard}, these methods load specialized task adapters and use routing layers to compute sample-wise ensembling weights on-the-fly. When task labels are unknown, stateless similarity routers like SABLE \cite{sable}, SPS-ZCA \cite{spszca}, and Q-SPS \cite{qsps} match intermediate representations to pre-computed centroids. However, making routing decisions independently at each layer makes them sensitive to noise, causing high routing jitter and cascading representation drift \cite{qmerge}.

Crucially, routing consistency and trajectory smoothing across network depth are known challenges in the sparse MoE literature. Prior work has highlighted that independent, layer-wise sparse gating decisions can cause high-frequency token-routing oscillations, degrading performance and training stability \cite{zoph2022stmoe, clark2022unified}. This has led to the exploration of depth-consistent gating networks that share or regularize routing choices across layers. Our work extends these insights to the training-free model ensembling setting, stabilizing ensembling trajectories across layers to prevent representational drift.

Furthermore, we identify a fundamental physical mechanism supporting stateful smoothing: standard residual and skip connections (present in all modern Transformer architectures) already act as natural low-pass filters on activation representations. Propagating intermediate states via $h^{(l)} = h^{(l-1)} + F(h^{(l-1)})$ implicitly carries task-relevant information forward across depth. This underlying representational continuity explains why a simple, state-independent momentum filter on ensembling weights is so effective at preserving task-expert specialization without requiring complex biochemical or virtual-time ODE integrations.

\subsection{Stateful and Continuous-Time Networks}
To stabilize ensembling trajectories, stateful routing frameworks have emerged. These methods draw inspiration from Neural Ordinary Differential Equations (Neural ODEs) \cite{chen2018neural} and Deep Equilibrium Models (DEQs) \cite{bai2019deep}, which model representation flow as continuous-time dynamical systems. Specifically, SOTA ChemMerge \cite{chemmerge} tracks ensembling coefficients as chemical concentrations governed by biochemical kinetics. Though stabilizing ensembling weights and dampening noise, continuous stateful routing introduces overhead from virtual stepping ($\Delta t$), Arrhenius rates, and ODE solvers. Our work deconstructs this continuous-time stateful paradigm, showing that its complexity is redundant and can be replaced by a simple classical momentum update.

\subsection{Minimalism and Parsimony in Deep Learning}
Our work aligns with the philosophy of parsimony in deep learning. Prior work advocates for stripping redundant components to build simpler, reproducible, and interpretable models \cite{mitchell1980need, duda2000pattern}. Examples include matching complex schedules with simple momentum \cite{polyak1964some, nesterov1983method}, or dualizing representation learning as exponential moving averages \cite{grill2020byol, he2020moco}. By applying Occam's razor to stateful model merging, we continue this tradition, proving that a single standard equation outperforms a highly convoluted physical framework.
