\begin{abstract}
Dynamic test-time model ensembling across heterogeneous, non-stationary task streams has emerged as a critical capability for adaptive neural serving.
Existing stateful routing paradigms, however, are fundamentally bottlenecked by their reliance on flat Euclidean state updates.
By performing temporal smoothing or biochemical kinetics calculations in unconstrained spaces and projecting onto the probability simplex via post-hoc Softmax normalization, these methods introduce severe representational lag (hysteresis), violate geometric scale constraints, and suffer from high-frequency routing jitter.
In this work, we propose \textbf{Unitary Geodesic Routing (UGR)}, a geometric alternative that rejects flat Euclidean update spaces in favor of curved manifold flow.
Operating directly on the $(K-1)$-dimensional hypersphere $\mathbb{S}^{K-1}$, UGR leverages the square-root homeomorphism from Information Geometry to map coordinates to the probability simplex natively.
This guarantees exact simplex constraints with zero mathematical artifacts, representing a closed-form Fisher-Rao geodesic flow.
We formulate state updates as closed-form, Rodrigues-like geodesic rotations (spherical interpolations) along the shortest great-circle path, bypassing costly matrix exponentials.
To reconcile the stability-plasticity dilemma, we introduce Torque-Driven Adaptive Agility, which dynamically accelerates state transitions in proportion to representational mismatch (torque), and Spatial-Temporal Geodesic Coupling to smoothly propagate ensembling trajectories across queries.
Our state representation, updates, and projection are Softmax-free, eliminating representational scale compression.
We evaluate UGR across both a high-fidelity synthetic sandbox and a multi-task text classification Mixture-of-Experts routing benchmark using the \texttt{20newsgroups} dataset.
On the synthetic task stream, UGR achieves a Joint Mean Accuracy of \textbf{75.08\%} (outperforming the SOTA continuous biochemical kinetics baseline by \textbf{+5.43\%}) while slashing routing jitter ($L \ge 5$) by \textbf{2.10$\times$}.
On the text classification stream, UGR delivers a Joint Mean Accuracy of \textbf{92.25\%} ($\pm$ 0.90\%), outperforming the SOTA continuous router by a massive \textbf{+21.60\%} absolute margin while reducing routing jitter by \textbf{16.8$\times$}, establishing a new state-of-the-art Pareto frontier for adaptive test-time ensembling.
\end{abstract}
