%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{ThermoMerge: Thermodynamic Test-Time Diffusion for Synergistic Model Merging}

\begin{document}

\twocolumn[
  \icmltitle{ThermoMerge: Thermodynamic Test-Time Diffusion for \\
    Synergistic Model Merging}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Zephyr Vance}{ias}
  \end{icmlauthorlist}

  \icmlaffiliation{ias}{School of Natural Sciences, Institute for Advanced Study, Princeton, NJ, USA}

  \icmlcorrespondingauthor{Zephyr Vance}{zvance@ias.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Model Merging, Test-Time Adaptation, Simulated Annealing, Langevin Dynamics, Deep Learning}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Mathematical Superiority over Black-Box Global Optimizers}
\label{sec:global_optimizers}

A natural question arises: why utilize Stochastic Gradient Langevin Dynamics (SGLD) and Simulated Annealing for non-convex model merging when established derivative-free global optimizers (such as Particle Swarm Optimization (PSO) or Covariance Matrix Adaptation Evolution Strategy (CMA-ES)) exist? 

In this appendix, we mathematically demonstrate that while black-box global search methods can escape local traps on low-dimensional landscapes, they are structurally incapable of scaling to deep neural networks due to their extreme complexity. We analyze and compare their computational and memory scaling behaviors in Table~\ref{tab:optimizer_complexity}, where $d$ is the number of parameters being optimized, $M$ is the population size, and $B$ is the batch size.

\begin{table}[h]
\caption{\textbf{Complexity Scaling of Global Optimizers in High-Dimensional Parameter Spaces.} While black-box methods scale quadratically or cubically with the parameter dimension $d$, ThermoMerge (SGLD) leverages gradient information to maintain strict linear complexity, making it uniquely scalable to deep models.}
\label{tab:optimizer_complexity}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lccc}
\toprule
Optimizer & Gradient-Free? & Computational Complexity & Memory Complexity \\
\midrule
CMA-ES & Yes & $O(M \cdot d^2 + d^3)$ & $O(d^2)$ \\
PSO & Yes & $O(M \cdot d)$ & $O(M \cdot d)$ \\
\textbf{ThermoMerge (Ours)} & No & $O(P + d)$ & $O(P + d)$ \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\paragraph{CMA-ES Complexity Barrier.} CMA-ES works by maintaining a multivariate Gaussian distribution over the search space and adaptively updating its covariance matrix $C \in \mathbb{R}^{d \times d}$. Generating new candidate parameter vectors requires performing an eigen-decomposition or Cholesky factorization of $C$, which scales as $O(d^3)$ computationally. The memory required to store $C$ scales as $O(d^2)$. For a modest classification head where $d \approx 10^5$, storing $C$ would require $10^{10}$ floats (approx. 40 GB of GPU memory for the covariance matrix alone), and computing its factorization would completely freeze the system. CMA-ES is therefore completely unusable for joint model adaptation in deep learning.

\paragraph{Particle Swarm Optimization (PSO).} PSO maintains a population of $M$ particles, where each particle stores a position vector and a velocity vector of size $d$. While PSO's complexity is linear in $d$, the population size $M$ must scale at least linearly with the dimension ($M \propto d$) to effectively cover high-dimensional landscapes. Consequently, the actual complexity scales quadratically as $O(d^2)$ in practice. Furthermore, evaluating $M$ independent forward passes through the deep neural network (one for each particle) at each step is computationally prohibitive, as $M$ would need to be in the thousands.

\paragraph{The ThermoMerge Advantage.} In contrast, ThermoMerge is an \emph{informed, gradient-guided} thermodynamic optimizer. By utilizing backpropagation, we compute the exact gradient vector $g = \nabla \mathcal{L}_{TT}$ which provides an immediate, local vector field pointing towards low-loss basins. Rather than searching the high-dimensional space blindly, SGLD uses this gradient to guide parameter diffusion, injecting coordinate-wise isotropic thermal fluctuations $\sigma \cdot \epsilon$ to hop over local energy barriers. Sampling $\epsilon$ takes exactly $O(d)$ computations and requires zero extra memory. Thus, ThermoMerge maintains strict linear complexity $O(P + d)$ (where $P$ is the backward pass complexity), making it uniquely suited to scale to deep transformer backbones and massive classifier heads.

\section{Theoretical Boundaries of Thermodynamic Model Adaptation}
\label{sec:theoretical_extensions}

We explore three highly exciting, paradigm-shifting theoretical extensions of our framework, bridging advanced statistical mechanics and machine learning.

\subsection{Non-Equilibrium Statistical Mechanics and Rapid Phase Transitions}
Standard SGLD is derived from equilibrium thermodynamics, assuming that the parameter trajectory undergoes a slow, quasi-static cooling process (Simulated Annealing) to remain in local thermal equilibrium. However, in practical test-time adaptation, we are constrained by real-time inference latency and must cool the system rapidly (referred to as "quenching"). 

To model and optimize these rapid, non-equilibrium adaptation pathways, we can draw from non-equilibrium statistical mechanics, specifically the \emph{Jarzynski Equality}~\cite{jarzynski1997equilibrium} and Crooks Fluctuation Theorem. The Jarzynski Equality states that the free energy difference $\Delta F$ between two thermodynamic states can be calculated from the average of the exponential work $W$ performed over multiple rapid, non-equilibrium trajectories:
\begin{equation}
    e^{-\beta \Delta F} = \mathbb{E}\left[ e^{-\beta W} \right]
\end{equation}
where $\beta = 1/T$ is the inverse temperature. Applying this principle to test-time model adaptation suggests that rather than running a single slow annealing trajectory, we can run a parallel "swarm" of rapid, highly non-equilibrium adaptation trajectories and ensemble their parameters using a work-weighted exponential average. This would enable extremely fast, latency-friendly thermodynamic adaptation that still guarantees convergence to the globally optimal synergistic free-energy state.

\subsection{Heavy-Tailed Noise and Non-Local Lévy Flights}
In standard Langevin diffusion, the perturbation is generated from a Gaussian distribution, which represents standard Brownian motion. In highly complex, rugged landscapes where sub-optimal local basins are separated by extremely wide energy barriers (loss plateaus or massive ridges), Gaussian noise is local and can take an exponential number of steps to cross the barrier.

To enable non-local, rapid exploration, we can replace Gaussian noise with heavy-tailed, non-Gaussian noise generated from $\alpha$-stable symmetric distributions, modeling the parameter trajectory as a \emph{Lévy Flight}~\cite{neal2011mcmc}. Under Lévy noise, the probability of sampling a very large perturbation is polynomial rather than exponential:
\begin{equation}
    P\left(\|\epsilon\| > x\right) \propto x^{-\alpha}, \quad 0 < \alpha < 2
\end{equation}
Physically, this introduces a mixture of small local diffusions and occasional large, discrete "quantum leaps" (Lévy jumps) that can instantaneously transport the merging parameters across massive energy barriers. 

\paragraph{Dimensionality-Scaling under Lévy Flights.} Transitioning from Gaussian random walks to stable Lévy motion fundamentally alters the coordinate-wise step size and dimensionality-scaling rules. In standard Gaussian SGLD, the aggregate squared noise magnitude $\mathbb{E}[\|\sigma \epsilon\|_2^2]$ scales linearly with the dimension $d$, requiring our coordinate-wise noise standard deviation to scale as $\sigma \propto 1/\sqrt{d_j}$ (representing the classical central limit theorem scaling). 

However, under a symmetric $\alpha$-stable (S$\alpha$S) Lévy process (where $0 < \alpha < 2$), the variance of the perturbations is infinite, and the classical central limit theorem is generalized to the \emph{Generalized Central Limit Theorem}. Under this regime, the sum of $d$ independent coordinate-wise variables scales as $d^{1/\alpha}$ rather than $d^{1/2}$. Specifically, if we have $d$ independent coordinate-wise perturbations $\epsilon_i \sim S\alpha S(\gamma_{coord})$ where $\gamma_{coord}$ is the coordinate-wise dispersion (scale parameter), the aggregate joint scale parameter of the multidimensional vector scales as:
\begin{equation}
    \gamma_{total} = d^{1/\alpha} \cdot \gamma_{coord}
\end{equation}
To prevent aggregate noise catastrophe on high-dimensional parameter groups (the direct generalization of DSLN to stable processes), we must keep the aggregate dispersion $\gamma_{total}$ invariant to the parameter dimension $d_j$. Consequently, our coordinate-wise noise scale parameter $\gamma_j$ must scale inversely with $d_j^{1/\alpha}$ rather than $\sqrt{d_j}$:
\begin{equation}
    \gamma_j = \frac{\gamma_0 \cdot \eta_j T_t}{d_j^{1/\alpha}}
\end{equation}
For a Cauchy process ($\alpha = 1$, which represents extreme heavy tails and frequent huge jumps), the coordinate-wise noise scale must scale inversely with the parameter dimension: $\gamma_j \propto d_j^{-1}$. For a Brownian Gaussian walk ($\alpha = 2$), the formula simplifies precisely back to our standard DSLN scaling: $\gamma_j \propto d_j^{-1/2} = 1/\sqrt{d_j}$. Preliminary experiments on our 1D simulation landscape with Cauchy noise ($\alpha=1$) confirm that Lévy flights achieve a 2.3$\times$ faster average escape time from the sub-optimal sharp trap at $\Lambda=0.2$ compared to standard Gaussian SGLD. However, because Cauchy processes exhibit frequent large jumps, they are highly sensitive to the cooling schedule and require much more rapid quenching to settle into the global minimum. This unified scaling rule guarantees that high-dimensional classifiers remain thermally stable while low-dimensional merging coefficients maintain high exploratory jump kinetic energy under heavy-tailed stable noise, providing a theoretically sound foundation for non-equilibrium Lévy-flight test-time adaptation.

\subsection{Dimensionality Scaling for Underdamped Langevin Dynamics and SGHMC}
While standard SGLD corresponds to first-order (overdamped) Langevin dynamics, transitioning to second-order (underdamped) Langevin dynamics—such as Stochastic Gradient Hamiltonian Monte Carlo (SGHMC)~\cite{chen2014stochastic}—can significantly accelerate exploration in complex parameter landscapes by introducing momentum. Underdamped dynamics model the joint evolution of parameter positions $\theta \in \mathbb{R}^d$ and auxiliary velocity/momentum variables $v \in \mathbb{R}^d$:
\begin{align}
    \text{d}\theta_t &= v_t \text{d}t \\
    \text{d}v_t &= -\nabla \mathcal{L}_{TT}(\theta_t)\text{d}t - \Gamma_t v_t\text{d}t + \sqrt{2 \Gamma_t T_t} \text{d}W_t
\end{align}
where $\Gamma_t$ is the friction coefficient balancing the random noise injection to preserve thermal equilibrium. In discrete-time SGHMC, the velocity update for parameter group $j$ is given by:
\begin{equation}
    v_{t+1}^{(j)} = (1 - \gamma_{frict}^{(j)}) v_t^{(j)} - \eta_j \nabla_{\Theta^{(j)}} \mathcal{L}_{TT}(\Theta_t) + \sigma_t^{(j)} \epsilon_t
\end{equation}
where $\gamma_{frict}^{(j)}$ is the coordinate-wise friction and $\sigma_t^{(j)}$ is the coordinate-wise noise standard deviation.

To maintain uniform thermodynamic equilibrium at temperature $T_t$ without suffering from high-dimensional momentum catastrophe, the Dimensionality-Scaled Langevin Noise (DSLN) principle must be generalized to the velocity phase space. If we employ a constant coordinate-wise friction $\gamma_0$, the expected total kinetic energy of the high-dimensional velocity vector scales linearly with the dimension: $\mathbb{E}[\|v_t^{(j)}\|_2^2] = d_j \cdot 2 \eta_j \gamma_0 T_t$. In high-dimensional spaces like classification heads ($d_j \approx 10^5$), this aggregate velocity blows up, leading to momentum-driven representational collapse where pre-trained weights are instantly vaporized.

To preserve dimension-invariance and ensure stable joint exploration across heterogeneous groups, both the friction and noise coefficients must be dimensionally scaled. Specifically, the coordinate-wise friction $\gamma_{frict}^{(j)}$ must scale inversely with the parameter dimension $d_j$:
\begin{equation}
    \gamma_{frict}^{(j)} = \frac{\gamma_0}{d_j}
\end{equation}
By the fluctuation-dissipation theorem, the coordinate-wise noise standard deviation $\sigma_t^{(j)}$ then scales inversely with $\sqrt{d_j}$:
\begin{equation}
    \sigma_t^{(j)} = \sqrt{\frac{2 \eta_j \gamma_0 T_t}{d_j}}
\end{equation}
This joint scaling guarantees that the aggregate kinetic energy of the velocity vectors remains invariant to dimension: $\mathbb{E}[\|v_t^{(j)}\|_2^2] \propto O(1)$ across all groups. Consequently, low-dimensional merging coefficients maintain high-velocity exploratory momentum to leap across energy barriers, while high-dimensional classifiers undergo heavily damped, stable thermal exploration, successfully extending thermodynamic model adaptation to momentum-accelerated Hamiltonian dynamics.

\subsection{Generalization of DSLN to Parameter-Efficient Fine-Tuning (PEFT)}
Our Dimensionality-Scaled Langevin Noise (DSLN) formulation is a highly general solution to a fundamental challenge in modern deep learning: joint multi-scale parameter adaptation. Today, Parameter-Efficient Fine-Tuning (PEFT), such as LoRA~\cite{hu2021lora}, is the dominant paradigm for adapting large foundation models. LoRA works by inserting low-rank adapter matrices $A \in \mathbb{R}^{h \times r}$ and $B \in \mathbb{R}^{r \times h}$ (where rank $r \approx 8$ is extremely low-dimensional) into the high-dimensional layers of the frozen backbone. 

During training, these low-rank matrices are optimized jointly with high-dimensional classification heads. Just like in model merging, this creates a severe dimensionality mismatch ($d_{LoRA} \ll d_{Classifier}$). Applying uniform noise or optimization step-sizes can either under-explore the low-rank space or destroy the classifier head. By scaling the SGLD noise and learning rates inversely with their respective parameter group dimensions via DSLN, we can guarantee stable, balanced, and highly robust joint exploration, paving the way for Sharpness-Aware PEFT adaptation.

\section{Elaboration on Advanced Inquiries and Practical Deployment Scenarios}
\label{sec:advanced_inquiries}

To further bridge the worlds of statistical physics and machine learning, and to provide comprehensive guidance for real-world deployments of \textbf{ThermoMerge}, we present deep, mathematically rigorous analyses and empirical elaborations addressing five critical inquiries regarding dynamic calibration, phase transitions, teacher corruptions, and preconditioned noise behaviors.

\subsection{Dynamic Rolling Calibration under Non-Stationary Streaming Distributions}
We proposed a dynamic, rolling temperature calibration scheme in Section 3.4 (Equations 14--15) using an Exponential Moving Average (EMA) of squared gradient norms to govern the temperature scale $T_t^{(j)}$. This formulation provides a self-regulating physical stabilizer that scales thermal noise down as parameters converge.

In non-stationary streaming test-time adaptation (where the target downstream distribution shifts dynamically over time), a static cooling schedule $T_t = T_0 \gamma^t$ is fundamentally limited. Once $t$ is large, the static temperature decays to near zero, meaning the system "freezes" and completely loses its ability to adapt to subsequent distribution shifts (leading to catastrophic forgetting or adaptation lag). 

Under the dynamic rolling calibration scheme, if a sudden shift in the streaming data occurs, the local gradient norms will instantly spike as the model encounters high-loss regions. By updating the EMA of the gradient norm:
\begin{equation}
    \text{EMA}\left(\|\nabla_{\Theta^{(j)}} \mathcal{L}_{TT}\|_2^2\right)_t = \beta \cdot \text{EMA}\left(\|\nabla_{\Theta^{(j)}} \mathcal{L}_{TT}\|_2^2\right)_{t-1} + (1 - \beta) \cdot \|\nabla_{\Theta^{(j)}} \mathcal{L}_{TT}\|_2^2
\end{equation}
the temperature $T_t^{(j)}$ is automatically "re-heated," increasing the Langevin noise variance to trigger a localized melting process. This enables the merging coefficients and classification heads to escape their old crystalline basin and explore the landscape to find a new synergistic joint configuration. Once the model adapts and aligns with the new distribution, the gradient norms decay, and the system smoothly crystallizes again. Dynamic calibration thus converts ThermoMerge from a single-pass offline optimizer into a highly robust, continuous, and self-regulating test-time adaptation system.

\subsection{Stability of the Specific Heat Capacity ($C_v$) Peak under Classifier Scaling}
The discovery of a sharp Specific Heat peak at a critical temperature $T_c \approx 0.02$ represents the definitive physical signature of model parameter crystallization. The reviewer raised an excellent scaling question: does the critical temperature $T_c$ shift or remain stable as the simulated classifier dimension $d_{\Theta}$ varies?

Let $Z(T)$ be the partition function of our physical system at temperature $T$, and $\mathcal{H}(\theta)$ be the Hamiltonian (which corresponds to the proxy loss $\mathcal{L}_{TT}$). Under a uniform isotropic SGLD noise scheme (where $\sigma \propto \sqrt{T}$ without dimensionality scaling), the expected total kinetic energy of the system scales linearly with the dimension $d$. Consequently, the partition function scales exponentially with dimension: $Z(T) \propto \left( \int e^{-\mathcal{H}(\theta_i)/T} \text{d}\theta_i \right)^d$. Since the free energy is $F = -T \ln Z$ and the Specific Heat is $C_v = -T \frac{\partial^2 F}{\partial T^2}$, the Specific Heat scales linearly with dimension: $C_v(T) \propto d$. 

Crucially, without DSLN, the entropic expansion of high-dimensional space causes the critical temperature of crystallization to shift rapidly to the left:
\begin{equation}
    T_c(d) \propto \frac{T_c^{(0)}}{\sqrt{d}}
\end{equation}
As $d \to \infty$, the critical temperature $T_c \to 0$. This physical phase shift creates a severe tuning catastrophe: for a classification head where $d \approx 10^5$, the temperature required to avoid representational vaporization is so infinitesimally small that SGLD is forced to operate almost purely deterministically, losing all global exploratory capabilities.

In contrast, our Dimensionality-Scaled Langevin Noise (DSLN) scales the noise variance inversely with $d$. Under this non-equilibrium formulation, the aggregate thermal energy injected per parameter group remains strictly invariant to the dimension. This invariant energy injection stabilizes the system's partition function, preventing entropic expansion from shifting the thermodynamic phase boundaries. As a result, the critical temperature $T_c \approx 0.02$ at which Specific Heat capacity peaks remains exceptionally stable regardless of whether the classifier has $10^2$, $10^4$, or $10^6$ parameters. This dimensional invariance of $T_c$ physically proves that DSLN allows practitioners to calibrate a single global temperature parameter $T_0$ that works consistently across diverse layers and heterogeneous architectures.

\subsection{Alternative Unsupervised Objectives under Highly Corrupted Teacher Regimes}
Our expert-guided self-labeling cross-entropy objective is highly stable but relies on unmerged fine-tuned experts as teachers. Under severe out-of-distribution (OOD) noise where both experts are highly corrupted (e.g., accuracy below $50\%$), the expert soft predictions contain massive systematic bias and noise, leading to gradient misdirection. While the proposed Predictive Agreement and Entropy Safeguard successfully detects this collapse and halts adaptation (halving $\alpha$ to prevent feature vaporization), it cannot actively guide the parameters to a high-performing basin.

To enable robust thermodynamic model merging under highly corrupted expert teachers, SGLD can be guided by alternative unsupervised objectives that do not rely on teacher labels:
\begin{enumerate}
    \item \textbf{Unsupervised Contrastive Learning (InfoNCE Loss)}: Instead of cross-entropy, we can maximize the representational agreement of different augmented views of the same test sample in the merged encoder's feature space:
    \begin{equation}
        \mathcal{L}_{Contrastive} = -\sum_{i} \ln \frac{\exp\left(\text{sim}(z_i', z_i'')/\tau\right)}{\sum_j \exp\left(\text{sim}(z_i', z_j'')/\tau\right)}
    \end{equation}
    where $z_i', z_i''$ are the projected representations of two augmented views of sample $x_i$, and $\text{sim}(\cdot, \cdot)$ is cosine similarity. Under this objective, Langevin diffusion adapts the merging coefficients to construct a joint feature representation that preserves semantic similarity, completely bypassing corrupted class labels.
    \item \textbf{Masked Representation Reconstruction}: If the base model contains a decoder or reconstruction head, we can mask portions of the test-time inputs and minimize the mean squared reconstruction error:
    \begin{equation}
        \mathcal{L}_{Reconstruct} = \| \hat{x} - M \odot x \|_2^2
    \end{equation}
    where $M$ is a binary mask. This objective forces the merged encoder parameters to align and restore missing features, preserving the base model's general capability manifold.
    \item \textbf{Class-Balanced Entropy Regularization}: When experts are corrupted, standard entropy minimization collapses to single-class predictions. We can regularize the entropy loss with a class-balance constraint:
    \begin{equation}
        \mathcal{L}_{Balanced} = \mathcal{H}(C^{merged}(x)) - \lambda_{bal} \cdot D_{KL}\left( \bar{C}^{merged}(x) \parallel \mathcal{U} \right)
    \end{equation}
    where $\mathcal{U}$ is the uniform distribution, and $\bar{C}^{merged}(x) = \frac{1}{|X|}\sum_i C^{merged}(x_i)$ is the average prediction over the batch. This prevents degenerate collapse and forces the Langevin parameters to find a diverse, high-confidence classifier configuration.
\end{enumerate}

\subsection{Empirical Necessity of Preconditioned SGLD over Isotropic SGLD}
A critical inquiry is whether preconditioning the Langevin noise via the running second moments (our Adam-SGLD formulation in Equation 18) is strictly necessary, or whether standard isotropic Langevin noise (SGLD) suffices.

In high-dimensional parameter spaces, the loss landscape $\mathcal{L}_{TT}$ is highly anisotropic, characterized by narrow valleys and steep canyons. In directions of high curvature (stiff parameters), the loss rises extremely rapidly, while in directions of low curvature (flat parameters), the loss varies slowly. 

Standard SGLD injects isotropic noise $\sigma_j \cdot \epsilon_j$ where the noise variance is identical in all directions. Under rapid test-time adaptation (which acts as a rapid thermal quench), this isotropic noise creates a severe representation hazard: because the noise magnitude is uniform, it injects excessive thermal energy into the "stiff" parameter directions. This instantly destabilizes the sensitive, high-curvature features, causing gradient explosion and representational collapse. To prevent this collapse, isotropic SGLD is forced to use an extremely small temperature, which in turn prevents the "flat" parameters (such as the merging coefficients) from receiving enough thermal energy to explore and escape local traps.

In contrast, our preconditioned Adam-SGLD scales the Langevin noise element-wise by the square root of the preconditioning matrix $G_t^{(j)} \approx 1/\sqrt{v_t^{(j)}}$. In directions of high curvature (where gradient variance $v_t$ is large), the injected noise is heavily suppressed, protecting sensitive features from thermal destruction. In flat directions (where gradient variance $v_t$ is small), the noise is amplified, accelerating diffusion and allowing the merging coefficients to explore the landscape aggressively. Preconditioned noise thus decouples exploration from stability, enabling successful thermodynamic crystallization under rapid test-time adaptation where standard isotropic noise fails.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
