%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{ChemMerge: Non-Equilibrium Chemical Reaction Kinetics for Dynamic Model Merging}

\begin{document}

\twocolumn[
  \icmltitle{ChemMerge: Non-Equilibrium Chemical Reaction Kinetics for \\
    Dynamic Model Merging}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Julian Vance}{equal,iapc}
    \icmlauthor{Aria Thorne}{equal,iapc}
  \end{icmlauthorlist}

  \icmlaffiliation{iapc}{Institute for Advanced Physical Computation, ETH Zurich, Switzerland}

  \icmlcorrespondingauthor{Julian Vance}{julian.vance@iapc.ethz.ch}
  \icmlcorrespondingauthor{Aria Thorne}{aria.thorne@iapc.ethz.ch}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Model Merging, Chemical Reaction Kinetics, Deep Learning, Dynamic Adaptation}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{Analytical Coordinate Sandbox (ICS) Formulation}
\label{appendix:ics}

To rigorously evaluate dynamic model ensembling methods in streaming environments, we formulated the \textbf{Analytical Coordinate Sandbox (ICS)}. ICS is a controlled, high-fidelity simulation environment that models representation drift, parameter interference, and adaptive routing across a multi-layer deep network.

\subsection{Task Generation and Representation Space}
We consider a representation space of dimension $D = 192$ across $L = 14$ layers. We define $K = 4$ independent task manifolds representing MNIST, Fashion-MNIST, CIFAR-10, and SVHN, respectively. Each task $k \in \{1, \dots, K\}$ is characterized by an intrinsic task signature coordinate vector $\mathbf{v}_k \in \mathbb{R}^D$ drawn from a task-specific distribution. 

An input sample $b$ belonging to task $k$ is generated as:
\begin{equation}
h_{b}^{(0)} = \mathbf{v}_k + \boldsymbol{\epsilon}_b, \quad \boldsymbol{\epsilon}_b \sim \mathcal{N}(\mathbf{0}, \sigma_k^2 \mathbf{I})
\end{equation}
where $\sigma_k$ represents the task-specific representation noise scale. To mirror the difficulty of real-world datasets, we calibrate the noise scales as $\sigma = [0.05, 0.15, 0.40, 1.20]$ for the four tasks, respectively. This introduces a realistic gradient of task difficulty and manifold dispersion.

\subsection{Base Model and Expert LoRA Paths}
The shared base model weights $W_{\text{base}}^{(l)} \in \mathbb{R}^{D \times D}$ are modeled as orthogonal projection matrices to represent stable feature propagation across layers. Each task-specific LoRA expert is represented by a set of low-rank matrices $A_k^{(l)} \in \mathbb{R}^{D \times r}$ and $B_k^{(l)} \in \mathbb{R}^{r \times D}$ with rank $r = 8$. The adapter weights are designed to shift representation vectors toward the optimal decision manifold of task $k$ at each layer $l$:
\begin{equation}
h_{b}^{(l-1)} A_k^{(l)} B_k^{(l)} = \gamma^{(l)} \left( \mathbf{v}_k - h_{b}^{(l-1)} \right)
\end{equation}
where $\gamma^{(l)} \in (0, 1]$ is a depth-dependent scaling factor that represents the progressive specialization of representation features.

\subsection{Classification and Evaluation Metrics}
At the final layer $L$, the representation vector $h_{b}^{(L)}$ is projected onto task-specific classification heads. The prediction confidence for task $k$ is computed using the distance to the target manifold:
\begin{equation}
P(\hat{y}_b = y_b) = \exp\left( -\|h_{b}^{(L)} - \mathbf{v}_k\|_2^2 \right)
\end{equation}
This formulation ensures a direct link between ensembling weight accuracy $\alpha_{k, b}^{(l)}$ across the depth of the network and the final model performance. Any ensembling weight deviation from the true expert pathway yields a representation trajectory that drifts into orthogonal manifolds, resulting in a severe degradation of the final classification score. This encapsulates both \emph{Heterogeneity Collapse} and \emph{Vectorization Collapse} inside a mathematically tractable sandbox.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
