\begin{abstract}
Model merging has emerged as a practical paradigm for composing multiple task-specific expert models into a single multi-task network without joint multi-task training. However, state-of-the-art adaptive model merging methods rely on test-time backpropagation and gradient descent over hundreds of steps. This makes them computationally and logistically prohibitive for resource-constrained edge devices (e.g., mobile phones, IoT nodes, embedded CPUs) where gradient tracking, backpropagation latency, and high GPU memory requirements are key bottlenecks. To bridge this gap, we propose \textbf{EdgeMerge}, a completely training-free, forward-only adaptive model merging framework. EdgeMerge extracts fine-grained, channel-wise merging coefficients in a single forward pass using a tiny, unlabeled calibration batch (e.g., 32 samples per task) through three novel techniques: Forward-Only Activation Sampling (FOAS), Scale-Normalized Delta Activation Salience (SNDAS), and Channel-Wise Softmax Gating (CWSG). Under a pragmatic engineering lens, we evaluate EdgeMerge on the 8-task Vision-Language CLIP ViT-B/32 benchmark and explicitly report its performance-resource trade-offs. While our training-free, forward-only approach incurs a substantial accuracy penalty compared to server-grade, gradient-based methods like SyMerge (68.69\% vs. 89.74\%), it matches the absolute peak performance of a thoroughly optimized Task Arithmetic baseline (68.74\%) while requiring no backpropagation and reducing preparation latency from 10 minutes to just 11.95 seconds. Crucially, EdgeMerge demonstrates a much more robust and stable scaling hyperparameter landscape than Task Arithmetic. This work highlights the viability of training-free, forward-only weight-space routing as a robust and hyperparameter-stable solution for resource-constrained edge systems.
\end{abstract}
