\section{Related Work}
\label{sec:related}

\textbf{Model Merging and Weight Soups.} Weight-space averaging of fine-tuned models combines specialized networks without extra inference compute. Model Soups \cite{wortsman2022model} showed that uniform weight averaging improves generalization. Git Re-Basin \cite{ainsworth2022git} leveraged permutation-matching to align internal representations, while Task Arithmetic \cite{ilharco2022editing} edited models via parameter vector addition. However, these static techniques produce frozen weights that cannot adapt dynamically on a sample-by-sample level.

\textbf{Dynamic and Stateful Ensembling.} Dynamic ensembling blends activations or weights on the fly during inference. Mixture of Experts (MoE) \cite{shazeer2017outrageously, fedus2022switch} routes inputs dynamically to specialized sub-networks but incurs significant memory bandwidth overheads. Latent coordinate-space ensembling, evaluated inside the Coordinate Sandbox (ICS), offers a lighter alternative. SABLE \cite{sable2024stateless} introduced training-free centroid-based activation ensembling, ChemMerge \cite{chemmerge2025continuous} used chemical kinetics to smooth ensembling coefficients, and Momentum-Merge \cite{momentummerge2025exponential} simplified kinetics to an EMA. PAC-Kinetics \cite{packinetics2025learning} formulated generalization bounds, while a methodological audit \cite{audit2025deconstructing} showed that regularized linear gating heads are competitively robust but vulnerable to a small-sample bottleneck. Crucially, all these frameworks assume Float32 coordinates and fail under quantization.

\textbf{Neural Network Quantization and STE.} To enable edge deployment, neural network quantization compresses weights and activations to low-precision integers \cite{han2015deep, hubara2017quantized}. Post-Training Quantization (PTQ) \cite{nagel2019data, nagel2020up} computes scales without retraining but can degrade at low bit-widths. Quantization-Aware Training (QAT) \cite{jacob2018quantization, choi2018pact} integrates quantization into training. Since rounding ($\lfloor \cdot \rceil$) has a zero derivative almost everywhere, standard backpropagation fails. The Straight-Through Estimator (STE) \cite{bengio2013estimating, courbariaux2015binaryconnect, yin2019understanding} bypasses this by approximating the derivative as the identity mapping. While STE is widely used for individual networks, we are the first to apply it to routing networks in dynamic activation ensembling.

\textbf{Efficient Edge Serving.} Edge serving is primarily bottlenecked by memory bandwidth, especially when loading multiple expert weight sets dynamically \cite{sheng2023high, kwon2023efficient}. Quantizing weights and activations to INT8 or INT4 addresses this, enabling high-throughput serving on tiny processors \cite{howard2017mobilenets, dettmers2022gpt3, xiao2023smoothquant, dettmers2024qlora}. We bridge the gap between algorithmic ensembling and hardware constraints by designing a quantization-robust coordinate ensembling method operating entirely on discrete, hardware-compatible arithmetic blocks.
