\begin{abstract}
Traditional parameter-space model merging methods rely on the restrictive assumption that task-specific experts must be blended into a single static, input-independent set of weights. In high-conflict multi-task settings and compact model regimes (such as Vision Transformers), this static compromise inevitably leads to \emph{catastrophic representational collapse}, where the merged network's performance degenerates severely. In this work, we propose a radical paradigm shift inspired by quantum mechanics: \textbf{Quantum Wavefunction Superposition Merging (QWS-Merge)}. Instead of searching for a static linear consensus in weight-space, we model fine-tuned expert weights as task eigenstates in a parameter Hilbert space. Merging is represented as a coherent quantum-like superposition of these states, which dynamically ``collapses'' (measures) to a task-specific weight configuration based on the phase-overlap of incoming input features. Under this formulation, a lightweight phase projector projects input representation states onto a learned layer-wise phase-basis, allowing wave-like interference to dynamically assemble optimal parameter paths on-the-fly. This approach is highly parameter-efficient, utilizing only 336 trainable parameters that can be optimized on a tiny offline validation set of 16 samples per task. Our empirical evaluation on a challenging multi-task visual benchmark (MNIST, FashionMNIST, CIFAR-10, SVHN) using a compact 5.7M parameter ViT-Tiny backbone demonstrates that QWS-Merge completely resolves representational collapse, improving the average multi-task accuracy from $49.35\%$ (Uniform Merging) to \textbf{59.32\%}. Furthermore, compared to a classical Linear Router baseline, QWS-Merge exhibits strong regularization properties: under extreme task conflict (SVHN), the Linear Router collapses to $15.30\%$, while QWS-Merge maintains a high \textbf{31.60\%} (preserving $91.5\%$ of expert capacity). Finally, we transparently document and analyze the sensitivity of dynamic merging to batch heterogeneity, exposing a critical ``heterogeneity collapse'' at larger batch sizes. This work establishes quantum-inspired wavefunction superposition as a powerful, regularized, and highly-efficient new frontier for deep parameter fusion.
\end{abstract}
