\begin{abstract}
Test-time dynamic model merging has emerged as a powerful paradigm for combining specialized neural task experts into a single unified multi-task model at inference time.
However, standard dynamic routers suffer from severe vulnerabilities in real-world deployment. In this paper, we expose a critical, previously unstudied phenomenon: the \emph{Batch-Average Smoothing Confounder}. Under standard large-batch evaluation, batch averaging of predicted coefficients acts as an implicit smoothing operator that masks the severe overfitting of the router. When deployed in true, real-time batch-independent or sample-wise vectorized pipelines ($B=1$), this smoothing mask is removed, causing a severe \emph{Vectorization Collapse} where standard dynamic routers exhibit substantial performance degradation (e.g., standard random-initialized L3-Softmax drops to 41.09\% accuracy, nearly 17\% below naive Uniform Merging).
To resolve this, we present a rigorous empirical audit of routing mechanics on a controlled 192-dimensional synthetic Analytical Coordinate Sandbox across 10 independent random seeds. We show that proper initialization and standard regularization---specifically, zero-initialized Softmax routing coupled with weight decay---act as the true foundational drivers of stability, preventing the router from overfitting to data scarcity (64 calibration samples). We show that while the explicit Task-Variance Regularization ($\mathcal{L}_{VR}$) loss penalty provides a principled group-variance limit, it operates in tandem with this powerful implicit zero-initialized prior. Under this unified Prior-Driven Classical Routing Framework, both our well-regularized L3-Softmax baseline (Zero-Initialized Softmax Routing) and its variance-regularized variant (VR-Router) completely resolve vectorization collapse, maintaining a flatline, highly-generalizing joint accuracy of \textbf{59.16\% $\pm$ 1.17\%} and \textbf{59.14\% $\pm$ 1.18\%} respectively across all batch sizes ($B=1$ to $B=512$), outperforming Uniform Merging (58.00\% $\pm$ 1.13\%) and all unregularized dynamic baselines.
\end{abstract}
