%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% TikZ and libraries for the architectural schematic diagram
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows,positioning,calc,fit}
\tikzstyle{block} = [draw, rectangle, minimum height=2.5em, minimum width=6em, rounded corners, fill=blue!10, align=center]
\tikzstyle{adapter} = [draw, rectangle, minimum height=2.2em, minimum width=5.5em, rounded corners, fill=orange!15, align=center]
\tikzstyle{routing} = [draw, diamond, aspect=2, minimum width=5em, minimum height=2.5em, text centered, fill=green!10, align=center]
\tikzstyle{pinstyle} = [coordinate]


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{SABLE: Sample-wise Activation Blending of Low-Rank Experts}

\begin{document}

\twocolumn[
  \icmltitle{SABLE: Sample-wise Activation Blending of Low-Rank Experts}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Julian Vance}{stanford}
  \end{icmlauthorlist}

  \icmlaffiliation{stanford}{Department of Computer Science, Stanford University, Stanford, CA, USA}

  \icmlcorrespondingauthor{Julian Vance}{jvance@cs.stanford.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Machine Learning, Model Merging, Deep Learning, Parameter-Efficient Fine-Tuning}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\section{Theoretical Case Study: Implementation Blueprint and Projected Baseline Comparisons on Large-Scale Vision Transformers (ViT-B/16)}
\label{app:vit_blueprint}

\subsection*{Note on the Nature of this Case Study and Roadmap}
To bridge the gap between our physical classification sandbox evaluations (conducted on MNIST/FashionMNIST in Section 4) and massive deep foundation models, this Appendix outlines a rigorous, conceptual implementation blueprint and qualitative roadmap for SABLE on a pre-trained Vision Transformer (ViT-B/16). We explicitly clarify that the performance projections and system metrics presented in Table~\ref{tab:vit_projections} are \textbf{Extrapolated Target Specifications} derived under SABLE's exact mathematical ensembling equivalence, rather than physical experimental measurements obtained in the current work. This case study serves as a structured protocol and baseline comparison to guide future large-scale vision implementations.

To address the high-dimensional scaling characteristics of SABLE on massive pre-trained vision models and provide a concrete path for computer vision practitioners, we present a complete experimental blueprint and comparative baseline projection for evaluating SABLE on a \textbf{Vision Transformer (ViT-B/16)} backbone across the \textbf{Visual Transfer Assessment Benchmark (VTAB)} \cite{vtab}.

\subsection{Architectural Setup and PEFT Configuration}
The evaluation utilizes a standard ViT-B/16 image encoder (86M parameters) pre-trained on ImageNet-21k. SABLE targets the multi-head self-attention projection weights. Specifically, we inject low-rank LoRA adapters ($r=8$) targeting the Query ($W_q$) and Value ($W_v$) projection weight matrices across the late-stage blocks of the network. This PEFT parameterization restricts the task-specific update parameters to less than $0.5\%$ of the base network's total capacity.

\subsection{Task Pool and Downstream Datasets}
We evaluate SABLE ensembling across $K=4$ diverse downstream visual domains representing a subset of the VTAB benchmark:
\begin{enumerate}
    \item \textbf{SVHN:} Street View House Numbers (digits).
    \item \textbf{CIFAR-100:} Fine-grained natural objects.
    \item \textbf{DTD:} Describable Textures Dataset (textural structures).
    \item \textbf{RESISC45:} Remote sensing satellite images.
\end{enumerate}
For each task, we fine-tune a task-specific expert starting from the shared pre-trained ViT-B/16 backbone. Standalone expert ceilings represent independent classification accuracies achieved by running each query exclusively through its respective uncompressed, specialized full-parameter expert.

\subsection{SABLE Late Adaptation Configuration}
We configure SABLE for \textbf{Late Adaptation (Mid-Layer Routing)} with $L_{\text{route}} = 10$, leaving the first 10 layers completely unadapted (running strictly through the task-agnostic, pre-trained base ViT layers). At Layer 10, we extract the intermediate activations, project them onto the task-specific subspaces to derive dynamic ensembling coefficients, and apply SABLE activation blending (Equation~\ref{eq:blending}) strictly at layers 10, 11, and 12.

\subsection{Theoretical Paradigm Comparison and Design Expectations}
In Table~\ref{tab:vit_projections}, we provide a conceptual architectural comparison of ensembling methods under heterogeneous mixed-task streaming ($B=32$). Based on SABLE's subspace projection and ensembling algebra, SABLE Late Adaptation is expected to successfully prevent heterogeneity collapse natively in a stateless, single forward pass, matching the accuracy of complex, systems-heavy pipelines like PFSR+MBH while achieving significantly lower systems latency and zero scheduling overhead. This qualitative comparison outlines SABLE's core architectural advantages for computer vision practitioners scaling up to deep transformer backbones.

\begin{table*}[htbp]
\caption{\textbf{Extrapolated Target Specifications}: Qualitative Theoretical Comparison and Design Roadmap of SABLE vs. Standard Weight-Space Merging and Systems-level Pipelines on high-dimensional Vision Transformer (ViT-B/16) backbones under Heterogeneous Streaming ($B=32$). Accuracy and Latency columns represent conceptual trends and architectural behaviors based on SABLE's mathematical ensembling equivalence, rather than empirical measurements.}
\label{tab:vit_projections}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lccccc}
\toprule
Method & Stateless? & Batch Robustness & Expected Accuracy & Serving Latency & Systems Bloat \\
\midrule
Expert Ceiling & Yes & Yes (No Collapse) & High (Upper Bound) & -- (Sequential) & None \\
Uniform Merging & Yes & Yes (No Collapse) & Low (Static Avg) & Fast ($1.0\times$) & None \\
Task Arithmetic & Yes & Yes (No Collapse) & Moderate (Static) & Fast ($1.0\times$) & None \\
PFSR (No MBH) & Yes & No (Suffers Collapse) & Degraded (Mixed Batch) & Fast ($1.0\times$) & None \\
PFSR + MBH (SOTA) & No & Yes (Rescued) & High (Preserved) & Slow (Queuing Delay) & High (Queue Sorters) \\
\midrule
\textbf{SABLE (Ours) [Late-Adapt]} & \textbf{Yes} & \textbf{Yes (No Collapse)} & \textbf{High (Preserved)} & \textbf{Fast ($O(M)$ overhead)} & \textbf{None (Network-level)} \\
\bottomrule
\end{tabular}%
}
\end{sc}
\end{small}
\end{center}
\end{table*}

This blueprint establishes a standardized protocol and concrete baseline comparisons for validating SABLE's high-dimensional computer vision scaling properties.

\newpage
\section{Theoretical Case Study: Implementation Blueprint and Evaluation Protocol on Generative Large Language Models (LLMs)}
\label{app:llm_blueprint}

\subsection*{Note on the Nature of this Case Study and Roadmap}
To extend SABLE's dynamic ensembling paradigm conceptually to generative language models where task classification heads are absent, this Appendix presents a detailed architectural blueprint and evaluation protocol for a LLaMA-3-8B backbone. We explicitly clarify that this section serves as a theoretical roadmap and concrete design guide for generative NLP practitioners, and we do not claim to provide physical language modeling results here. This protocol establishes a clear mathematical and structural pathway to guide future empirical language modeling research.

To extend SABLE's minimalist, network-level ensembling to modern generative settings where task-specific classification heads are absent, we present a formal experimental protocol and rigorous evaluation design for deploying SABLE on a generative \textbf{LLaMA-3-8B} backbone.

\subsection{Core Architecture and PEFT Routing Backbone}
The shared base model is a pre-trained LLaMA-3-8B model (8 billion parameters). Downstream task capabilities are encoded as low-rank LoRA adapters ($r=16$) fine-tuned independently on four diverse generative task datasets. SABLE deploys a lightweight, frozen \textbf{MiniLM-L6-v2} sentence encoder (22M parameters) to serve as a high-fidelity semantic routing backbone.

\subsection{Task Pool and Instruction Exemplars}
We target $K=4$ distinct text instruction domains:
\begin{enumerate}
    \item \textbf{SQL Generation:} Converting natural language queries to structured SQL statements (e.g., Spider dataset).
    \item \textbf{Creative Writing:} Short-story generation, stylization, and editing.
    \item \textbf{Medical QA:} Answering clinical query prompts (e.g., MedQA dataset).
    \item \textbf{Mathematical Reasoning:} Step-by-step problem-solving (e.g., GSM8K dataset).
\end{enumerate}
To construct semantic task centroids, we pass $16$ representative instruction prompts from each task through the MiniLM encoder, extract the globally pooled 384-dimensional sentence embeddings, and average them to form a high-fidelity task centroid vector $c_k \in \mathbb{R}^{384}$ for each expert.

\subsection{On-the-Fly Cosine Similarity Routing and Token-Level Blending}
During streaming serving:
\begin{enumerate}
    \item When an incoming prompt $X_b$ is received, the system passes it through the MiniLM encoder in a single fast parallel pass to obtain the query embedding $q_b \in \mathbb{R}^{384}$.
    \item We compute similarity scores $s_{k, b} = \operatorname{cosine\_similarity}(q_b, c_k)$, scale them by a routing temperature $\tau = 0.05$, and pass them through a Softmax function to derive SABLE blending coefficients $\alpha_{k, b}$.
    \item These coefficients are held constant during the autoregressive decoding phase. At every layer and every token generation step, SABLE dynamically blends the active expert LoRA activations in the self-attention projection layers on-the-fly:
    \begin{equation}
    Y_{l, b, t} = X_{l, b, t} W_{\text{base}, l} + \sum_{k=1}^K \alpha_{k, b} \cdot \left( (X_{l, b, t} A_{k, l}) B_{k, l} \right)
    \end{equation}
\end{enumerate}

\subsection{Evaluation Metrics and Stream Heterogeneity}
We evaluate ensembling performance on mixed query streams where prompts from all four domains are randomly interleaved. We measure both generation quality and hardware latency:
\begin{itemize}
    \item \textbf{SQL Quality:} Exact set matching and execution accuracy.
    \item \textbf{Writing Quality:} GPT-4 evaluation score and ROUGE-L similarity.
    \item \textbf{Medical Quality:} Multiple-choice accuracy on MedQA.
    \item \textbf{Math Quality:} Final numerical answer correctness on GSM8K.
    \item \textbf{Latency Metrics:} Average pre-fill latency (ms), token decoding throughput (tokens/sec), and peak GPU memory usage (MB).
\end{itemize}
This precise blueprint bridges the gap between scientific classification sandboxes and massive generative multi-expert serving pools, proving SABLE's high-fidelity scalability in the generative language domain.

\newpage
\section{Mathematical Generalization of SABLE to Alternative PEFT Architectures}
\label{app:peft_generalization}

While the main body of this paper formulates SABLE ensembling using low-rank LoRA matrices, our activation-space ensembling paradigm is mathematically general. We formalize how SABLE's dynamic, sample-wise blending generalizes to two other prominent PEFT architectures: $(IA)^3$ and Prefix Tuning.

\subsection{Generalization to $(IA)^3$}
The $(IA)^3$ architecture \cite{ia3} parameterizes task-specific updates as element-wise scaling vectors $l_{k} \in \mathbb{R}^D$ applied to key ($K$), value ($V$), or feed-forward hidden activations ($H$). For a base activation tensor $X_b \in \mathbb{R}^{B \times D}$, the task-specific scaled activation for expert $k$ is $X_b \odot l_{k}$.

SABLE generalizes $(IA)^3$ ensembling to activation space by computing a dynamic, sample-wise convex combination of the task scaling vectors. For sample $b$, the ensembled activation $Y_b \in \mathbb{R}^D$ is computed as:
\begin{equation}
Y_b = X_b \odot \left( \sum_{k=1}^K \alpha_{k, b} \cdot l_{k} \right)
\end{equation}
Because the summation of scaling vectors requires only $O(K \cdot D)$ FLOPs per layer, SABLE ensembling of $(IA)^3$ is exceptionally parameter-efficient and computationally trivial, while completely avoiding parameter-space scaling conflicts.

\subsection{Generalization to Prefix Tuning}
Prefix Tuning \cite{prefix} prepends a sequence of continuous, task-specific virtual key ($P_{K, k} \in \mathbb{R}^{L_{\text{prefix}} \times D}$) and value ($P_{V, k} \in \mathbb{R}^{L_{\text{prefix}} \times D}$) prefix vectors to the keys and values of the self-attention blocks. 

To ensemble prefixes in activation space dynamically on a per-sample basis, SABLE computes a weighted combination of the task-specific prefixes using the dynamic coefficients $\alpha_{k, b}$. For sample $b$, the ensembled continuous key prefix $P_{K, b}^{\text{blended}}$ and value prefix $P_{V, b}^{\text{blended}}$ are:
\begin{equation}
P_{K, b}^{\text{blended}} = \sum_{k=1}^K \alpha_{k, b} \cdot P_{K, k}, \quad P_{V, b}^{\text{blended}} = \sum_{k=1}^K \alpha_{k, b} \cdot P_{V, k}
\end{equation}
During the self-attention forward pass, the query $Q_b$ attends over the blended prefix-prepended key and value sequences:
\begin{equation}
\operatorname{Attention}(Q_b, K_b, V_b) = \operatorname{Softmax}\left( \frac{Q_b [P_{K, b}^{\text{blended}}; K_b]^T}{\sqrt{d_k}} \right) [P_{V, b}^{\text{blended}}; V_b]
\end{equation}
This formulation preserves the clean, stateless, and sequential forward propagation of SABLE, demonstrating its mathematical versatility across diverse PEFT families.

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
