\section{Conclusion and Future Work}
\label{sec:conclusion}

In this paper, we addressed a critical, real-world bottleneck in edge AI deployment: the hardware-quantization mismatch that occurs when merging specialized task experts for deployment on heterogeneous device fleets. We showed that existing quantization-aware model merging methods (like Q-Merge) overfit to the simulated quantization boundaries used during optimization, resulting in significant cross-schema performance degradation when deployed onto target accelerators running mismatched compilers.

To resolve this, we introduced \textbf{OmniMerge}, a training-free and computationally efficient multi-schema stochastic co-optimization framework. By combining Stochastic Operator Sampling (SOS) and Scale/Zero-Point Noise Perturbation (SZNP), OmniMerge acts as parameter-space data augmentation and smooths the non-differentiable loss landscape. This prevents boundary overfitting and guides the optimization toward flat, robust minima that are invariant to specific post-training quantization (PTQ) schemes.

Our empirical results on a \texttt{ViT-Tiny} backbone across four distinct classification tasks show that OmniMerge completely closes the cross-schema generalization gap under robust 8-bit quantization, outperforming all baseline methods across all 5 target compilers and achieving up to \textbf{50.78\%} average accuracy. 

As a zero-overhead, metadata-free, and training-free method, OmniMerge is exceptionally easy to integrate into existing MLOps pipelines. A single optimization sweep produces a robust merged checkpoint that is ready for deployment across diverse fleets of edge CPUs, GPUs, TPUs, and DSPs.

Future work will investigate extending OmniMerge to modern Large Language Models (LLMs) such as LLaMA. Specifically, we will explore how OmniMerge's stochastic operator sampling (SOS) and scale/zero-point noise perturbations (SZNP) can be adapted to scale to decoder architectures under sub-4-bit block-wise or group-wise configurations (e.g., in AWQ or GPTQ quantizers). In LLMs, high-magnitude outlier weights across specific feature channels present unique challenges, as standard post-training quantization often causes severe clipping or discretization errors in these critical dimensions. By stochastically perturbing the block-level quantization scales and zero-points during coefficient search, OmniMerge can act as a targeted parameter-space regularizer. This forces the test-time ensembling optimizer to identify blending ratios that are intrinsically robust to outlier-driven rounding errors, paving the way for reliable, zero-overhead ensembling of sparse, task-specific LLM experts directly on edge consumer devices.
