\begin{abstract}
Model merging has recently been characterized by complex, overparameterized frameworks relying on test-time optimization and coordinate warping. In this paper, we show that this complexity can be reduced using simple linear algebra. We propose \textbf{Spectral Model Merging via Singular Value Slicing (SVS)}, a non-parametric, training-free, and closed-form model merging operator. SVS performs a Singular Value Decomposition (SVD) on task-specific weight updates (task vectors) and retains only the top $k$ principal singular components, filtering out high-frequency fine-tuning noise. We also discuss \textbf{Barycentric Weight Normalization (BWN)}, an analytical scale-preservation operator designed to preserve weight scale in un-normalized layers. We evaluate SVS on CLIP-ViT-B/32 across four datasets (MNIST, FashionMNIST, CIFAR-10, SVHN). Our results show that SVS successfully consolidates multiple experts into a single multi-task model with zero additional trainable parameters or test-time steps. SVS at rank $k=128$ (retaining only $16.7\%$ of the rank space) matches or exceeds the performance of full-rank Task Arithmetic. Finally, we provide a mathematical proof demonstrating why global weight scaling is neutralized under modern feature normalization, and validate BWN's scale preservation in non-normalized multi-layer perceptrons. SVS offers a simple and computationally efficient baseline for model merging. Our code is available at \url{https://github.com/anonymous/svs}.
\end{abstract}
