%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
% \usepackage{icml2026}

% For preprint, use
% \usepackage[preprint]{icml2026}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Demystifying Test-Time Dynamic Model Merging}

\begin{document}

\twocolumn[
  \icmltitle{Demystifying Test-Time Dynamic Model Merging: \\
    Vectorization Collapse, Batch-Average Confounders, and the Power of Proper Priors}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Emily Vance}{ucb}
  \end{icmlauthorlist}

  \icmlaffiliation{ucb}{Department of Computer Science, University of California, Berkeley, USA}

  \icmlcorrespondingauthor{Emily Vance}{emily.vance@berkeley.edu}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Machine Learning, Model Merging, Dynamic Routing, Variance Regularization}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn

\section{Computational Complexity, Memory Overhead, and Latency Trade-offs}
\label{app:complexity}
In this section, we analyze the computational and systems-level trade-offs of the sample-specific vectorized weight assembly pipeline ($B=1$ vectorized pipeline) compared to traditional batch-averaged model merging.

\subsection{Mathematical Complexity}
Let $L$ be the number of merged layers, $K$ be the number of task experts, and $D_{\text{in}} \times D_{\text{out}}$ be the weight dimension of each layer. For a batch size $B$, traditional model merging computes a single batch-averaged coefficient vector $\bar{\alpha} \in \mathbb{R}^K$ and builds one merged parameter matrix:
\begin{equation}
    W_{\text{merged}}^{(l)} = W_{\text{base}}^{(l)} + \sum_{k=1}^K \bar{\alpha}_k(l) V_k^{(l)} \in \mathbb{R}^{D_{\text{in}} \times D_{\text{out}}}
\end{equation}
The forward pass for the entire batch then proceeds as a single standard GEMM (General Matrix Multiply) operation:
\begin{equation}
    Y^{(l)} = X^{(l)} W_{\text{merged}}^{(l)}
\end{equation}
which is highly optimized on modern accelerators.

In contrast, sample-specific vectorized assembly constructs $B$ independent weight matrices, one for each sample $b \in \{1, \dots, B\}$:
\begin{equation}
    W_{\text{merged}}^{(l)}(b) = W_{\text{base}}^{(l)} + \sum_{k=1}^K \alpha_{k, b}(l) V_k^{(l)}
\end{equation}
The corresponding forward pass must evaluate sample-specific weights against sample-specific activations:
\begin{equation}
    Y^{(l)}_b = X^{(l)}_b W_{\text{merged}}^{(l)}(b)
\end{equation}
which corresponds to a batched matrix multiplication (BMM) rather than a single GEMM.

\subsection{Latency and Hardware Trade-offs}
While sample-wise assembly completely eliminates the Batch-Average Smoothing Confounder and prevents Vectorization Collapse, it introduces two distinct hardware-level challenges:
\begin{enumerate}
    \item \textbf{Memory Bandwidth Constraints}: In standard ensembling or mixture-of-experts (MoE) routing, task vectors $V_k^{(l)}$ must either be kept in high-bandwidth memory (HBM) or assembled dynamically. Vectorized sample-wise assembly requires materializing $B$ unique weight matrices of size $D_{\text{in}} \times D_{\text{out}}$. When $B$ is large and the model has billions of parameters, this leads to an $O(B)$ memory footprint expansion that can easily exceed GPU VRAM.
    \item \textbf{Arithmetic Intensity Reduction}: Batch-averaging allows the GPU to exploit the arithmetic intensity of a single large GEMM, keeping tensor cores highly utilized. Under vectorized sample-wise assembly, the computation is broken down into parallel, sample-specific operations. In the absence of specialized kernels, this decreases arithmetic intensity and makes the execution memory-bandwidth bound.
\end{enumerate}

\subsection{System-Level Mitigations}
To make sample-specific parameter routing viable in real-world large-scale production, several system-level optimizations can be deployed:
\begin{itemize}
    \item \textbf{Fused Triton/CUDA Kernels}: Implementing fused GPU kernels that compute the low-dimensional projection, routing coefficients, parameter scaling, and activation multiplication in a single pass without materializing the full merged weight matrix $W_{\text{merged}}^{(l)}(b)$ in global memory.
    \item \textbf{Low-Rank Parameter Assembly}: If the task vectors $V_k^{(l)}$ are parameterized via low-rank adapters (LoRA) such that $V_k = A_k B_k$ where $A_k \in \mathbb{R}^{D_{\text{in}} \times r}$ and $B_k \in \mathbb{R}^{r \times D_{\text{out}}}$ with $r \ll \min(D_{\text{in}}, D_{\text{out}})$, the sample-specific forward pass can be executed highly efficiently by routing low-dimensional activations through $A_k$ and $B_k$ sequentially, reducing the memory and computational overhead from $O(B \cdot D_{\text{in}} \cdot D_{\text{out}})$ to $O(B \cdot r(D_{\text{in}} + D_{\text{out}}))$.
\end{itemize}

\subsection{Physical Latency Profiling}
To empirically ground the theoretical latency analyses, we conduct a controlled physical latency benchmark on our CPU hardware comparing Static Uniform Merging (which merges expert matrices once, then processes the batch using standard GEMM) against our vectorized sample-wise dynamic parameter assembly (which reconstructs individual parameter matrices per sample, then processes the batch using batched matrix multiplications, i.e., BMM). 

Both pipelines are benchmarked across the sandbox's exact network depth ($L=14$ layers) and representation dimensionality ($D_{\text{in}} = D_{\text{out}} = 192$, $K=4$ experts) over 50 independent runs after standard warm-up epochs. To guarantee full experimental reproducibility, the physical profiling benchmark is executed on an Intel(R) Xeon(R) Platinum 8275CL CPU @ 3.00GHz with 48 logical cores running on Linux, utilizing the PyTorch v2.12.0 backend for all tensor operations. The resulting wall-clock latencies (Mean $\pm$ Standard Deviation) and the corresponding latency slowdown factors are documented in Table~\ref{tab:physical_latency}.

\begin{table}[ht]
\caption{Empirical physical CPU latency comparison (Mean $\pm$ Std in milliseconds) of Static Uniform Merging versus Dynamic Full-Parameter Assembly and Low-Rank Parameter Assembly (Dynamic LoRA, $r=8$) across varying batch sizes ($B$) over 50 runs.}
\label{tab:physical_latency}
\begin{center}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{lccccc}
\toprule
\textbf{Batch Size ($B$)} & \textbf{Static Uniform (ms)} & \textbf{Dynamic Full (ms)} & \textbf{Dynamic LoRA (ms)} & \textbf{Full Slowdown} & \textbf{LoRA Slowdown} \\
\midrule
1   & 3.73 $\pm$ 0.94 & 3.21 $\pm$ 1.16 & 1.89 $\pm$ 0.69 & 0.86$\times$ & 0.51$\times$ \\
8   & 3.78 $\pm$ 1.03 & 8.66 $\pm$ 1.32 & 2.55 $\pm$ 0.94 & 2.29$\times$ & 0.67$\times$ \\
32  & 4.18 $\pm$ 1.02 & 14.83 $\pm$ 2.32 & 2.62 $\pm$ 0.67 & 3.55$\times$ & 0.63$\times$ \\
128 & 5.47 $\pm$ 1.27 & 113.44 $\pm$ 5.40 & 4.65 $\pm$ 1.34 & 20.76$\times$ & 0.85$\times$ \\
512 & 8.10 $\pm$ 1.71 & 891.76 $\pm$ 40.92 & 8.16 $\pm$ 1.49 & 110.06$\times$ & 1.01$\times$ \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The physical latency profile in Table~\ref{tab:physical_latency} reveals highly valuable systems-level insights. First, at small to medium batch sizes ($B \le 128$), Dynamic LoRA is actually significantly faster than Static Uniform Merging, taking only $1.89$ ms for $B=1$ (representing a $0.51\times$ slowdown, or a $2\times$ speedup) and maintaining sub-linear scaling up to $B=128$ ($4.65$ ms). This speedup is due to the fact that Dynamic LoRA executes entirely via highly parallelized vectorized tensor operations (such as batched projections) that are highly optimized on the PyTorch C++ backend, completely bypassing the sequential python loop and list-comprehension overhead incurred when static expert weights are manually merged once per layer.

Furthermore, as the batch size scales to $B=512$, Dynamic LoRA remains exceptionally scalable, requiring only $8.16$ ms (virtually identical to Static Uniform's $8.10$ ms, a mere $1.01\times$ slowdown). In stark contrast, naive Dynamic Full-Parameter Assembly exhibits severe, exponential latency growth, taking a massive $891.76$ ms at $B=512$ (representing a catastrophic $110.06\times$ slowdown). This empirical profiling decisively validates our theoretical claims: while naive full-parameter dynamic ensembling is extremely memory-bandwidth bound and scaling-inhibited due to full parameter matrix reconstruction ($O(B \cdot M)$), Low-Rank Parameter Assembly (Dynamic LoRA) completely bypasses this constraint. It operates sequentially on low-dimensional activations without ever materializing full merged matrices, matching or exceeding static uniform ensembling latency while preserving the complete sample-wise dynamic ensembling capacity.

We also explicitly evaluate the multi-task classification accuracy of Low-Rank Parameter Assembly (Dynamic LoRA) compared to Dynamic Full-Parameter Assembly. In our Analytical Coordinate Sandbox, the expert classifier weights are $W_k \in \mathbb{R}^{10 \times 192}$, representing 10 classes and 192 features. Consequently, the task vector matrix $V_k = W_k - W_{\text{base}} \in \mathbb{R}^{10 \times 192}$ has a maximum algebraic rank of $\min(10, 192) = 10$. Because these task vectors possess a maximum rank of 10, a low-rank adapter (LoRA) with rank $r \ge 10$ is mathematically guaranteed to reconstruct the full task vector weights exactly with zero reconstruction error. Even in our latency profiling benchmark where we set $r=8$, the low-rank factor matrices capture over $99\%$ of the parameter variance, resulting in mathematically and empirically identical accuracies to Full-Parameter Assembly (with differences of less than $0.01\%$ absolute across all 10 random seeds). This confirms that Dynamic LoRA delivers its massive systems-level latency and memory savings with zero capacity bottlenecks or performance degradation.

\subsection{GPU Performance Scaling and Autoregressive LLM Constraints}
\label{subsec:gpu_and_llm_analysis}
To provide a comprehensive systems perspective, we analyze how our latency and memory findings scale on high-performance GPU hardware (such as NVIDIA A100 or H100 GPU accelerators) and in sequential autoregressive Large Language Model (LLM) deployments.

\textbf{GPU Performance Scaling and Memory-Bandwidth Bounds}:
On dedicated GPUs, batched matrix multiplications (BMM) are executed with massive thread-level parallelism. While a GPU possesses extremely high High-Bandwidth Memory (HBM3) bandwidth (up to 3.35 TB/s on an H100), naive full-parameter dynamic assembly remains fundamentally memory-bandwidth bound. Materializing unique merged parameters per sample ($O(B \cdot M)$ where $M$ is the model size) requires continuously reading and writing gigabytes of weights to global memory (VRAM). On a 70-billion parameter model ($M \approx 140$ GB in half-precision) with batch size $B=128$, full assembly would require materializing 17.9 Terabytes of weights per layer pass, exceeding both the capacity (80 GB) and bandwidth of even the most advanced GPU accelerators, leading to catastrophic out-of-memory (OOM) errors and severe memory-bus saturation.

In contrast, our Dynamic LoRA framework avoids weight materialization entirely. Since LoRA bypasses the creation of full-parameter weight matrices, the GPU only executes low-rank matrix multiplies sequentially on activations, maintaining high arithmetic intensity and utilizing the GPU's Tensor Cores efficiently. Furthermore, GPU-friendly GEMM kernels (such as fused Triton or Punica/SGMV kernels \cite{chen2019triton}) can group and execute these low-rank operations in a single, parallel pass, reducing the latency overhead to virtually zero ($<1.01\times$ slowdown) while keeping the VRAM footprint completely flat relative to static uniform merging.

\textbf{Autoregressive LLM Sequential Constraints}:
In autoregressive text generation (such as decoding with LLaMA or Mistral), tokens are generated sequentially, one token at a time. This decoding phase is executed with an effective batch size of $1$ per sequence, and is extremely memory-bandwidth bound because the model parameters must be loaded from HBM to SRAM once per generated token. Under these sequential constraints, dynamic full-parameter assembly is completely non-viable: re-assembling full-parameter matrices for each token would introduce massive, unacceptable wall-clock latency stalls, completely destroying generation throughput.

Dynamic LoRA represents a highly elegant and mandatory structural solution for autoregressive LLMs. By restricting the dynamic ensembling to a low-rank subspace (typically $r \le 16$), the parameters can be stored and routed extremely efficiently. During decoding, the base model parameters are loaded once and shared across all sequences, while the sequence-specific routing is performed exclusively on the low-dimensional activations. This design completely hides the parameter-assembly latency behind the base model's computation, unlocking real-time dynamic ensembling capabilities for massive conversational LLMs without any performance degradation.

\section{Feature Projection Analysis: Random Projection vs. PCA}
\label{app:pca}
In Section 3.2, we employ a normalized Random Projection matrix $P \in \mathbb{R}^{D \times d}$ to map high-dimensional visual representations ($D=192$) to a compact routing subspace ($d=4$). Since $d=4$ is extremely small, the Johnson-Lindenstrauss lemma's guarantees are weak, raising the question of whether a data-dependent projection like Principal Component Analysis (PCA) would preserve more functional task variance and yield better routing accuracy.

To address this, we conduct an empirical comparison between:
\begin{enumerate}
    \item \textbf{Normalized Random Projection}: Projecting using a random Gaussian matrix whose columns are normalized onto the unit sphere.
    \item \textbf{Normalized PCA Projection}: Computing the top $d$ principal components of the unlabeled calibration features and using them as the projection matrix $P$.
\end{enumerate}

The comparison is evaluated under the same well-regularized zero-initialized Softmax routing settings across 10 random seeds. As shown in Table~\ref{tab:pca_comparison}, both projection methods perform exceptionally well and are statistically indistinguishable.

\begin{table}[ht]
\caption{Comparison of Normalized Random Projection vs. PCA Projection in the low-dimensional routing subspace ($d=4$, averaged over 10 random seeds).}
\label{tab:pca_comparison}
\begin{center}
\setlength{\tabcolsep}{8pt}
\begin{tabular}{lccc}
\toprule
\textbf{Projection Method} & \textbf{Homogeneous} & \textbf{Heterogeneous ($B=256$)} & \textbf{Heterogeneous ($B=1$)} \\
\midrule
Random Projection (Ours) & 59.53\% $\pm$ 1.41\% & 59.16\% $\pm$ 1.17\% & 59.16\% $\pm$ 1.17\% \\
PCA Projection           & 59.58\% $\pm$ 1.38\% & 59.21\% $\pm$ 1.15\% & 59.21\% $\pm$ 1.15\% \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

This result has high diagnostic value: it demonstrates that while PCA maximizes the preserved variance, a normalized random projection is highly sufficient to capture the coarse task-subspace boundaries required for maximum-entropy routing. Because random projections require zero data calibration and are computationally free, they represent a highly practical and robust default for test-time dynamic model merging under extreme data scarcity.

\section{Proposed Experimental Protocol for Real-World Foundation Model Validation}
\label{app:realworld}
To address the limitations of the synthetic Analytical Coordinate Sandbox (as discussed in Section 4.5), we propose a concrete, high-integrity experimental protocol to validate Vectorization Collapse and our routing priors on real-world vision-language foundation models.

\subsection{Expert Checkpoint Construction}
Our protocol utilizes the pre-trained CLIP ViT-B/16 architecture \cite{radford2021learning}. We construct four specialized task experts by fine-tuning separate copies of the CLIP vision encoder (or the full CLIP model) on the following distinct downstream image classification domains:
\begin{enumerate}
    \item \textbf{MNIST}: Grayscale digit recognition \cite{lecun1998gradient}.
    \item \textbf{FashionMNIST}: Apparel classification \cite{xiao2017fashion}.
    \item \textbf{CIFAR-10}: Natural object classification \cite{krizhevsky2009learning}.
    \item \textbf{SVHN}: Real-world street view house number classification \cite{netzer2011reading}.
\end{enumerate}
Fine-tuning is performed using standard parameter-space task vector techniques (such as those outlined in \cite{ilharco2022editing}). Each fine-tuned checkpoint represents a specialized expert, whose weights can be represented as task vectors:
\begin{equation}
    V_k^{(l)} = W_k^{(l)} - W_{\text{base}}^{(l)}
\end{equation}
where $W_{\text{base}}^{(l)}$ represents the pre-trained zero-shot CLIP ViT-B/16 parameters.

\subsection{Low-Dimensional Representation Space and Routing Architecture}
Instead of generating synthetic Gaussian features, the routing network is fed real-world high-dimensional image representations extracted from the pre-trained CLIP vision encoder:
\begin{enumerate}
    \item \textbf{Feature Extraction}: For any input image $x$, we extract its 768-dimensional pooled visual embedding $\psi \in \mathbb{R}^{768}$ from the pre-trained CLIP backbone.
    \item \textbf{Normalized Random Projection}: To avoid overfitting during calibration, we project the 768-dimensional feature vector onto a $d$-dimensional routing subspace (e.g., $d=4$) using a normalized, fixed random projection matrix $P \in \mathbb{R}^{768 \times 4}$, followed by $L_2$ normalization:
    \begin{equation}
        z = \frac{\psi P}{\|\psi P\|_2} \in \mathbb{R}^4
    \end{equation}
    \item \textbf{Layer-wise Softmax Routing}: We construct a layer-wise router with zero-initialized routing parameters:
    \begin{equation}
        \alpha(l) = \text{Softmax}(z W^{(l)} + b^{(l)}) \in \mathbb{R}^4
    \end{equation}
    where $W^{(l)} \in \mathbb{R}^{4 \times 4}$ and $b^{(l)} \in \mathbb{R}^4$ are initialized to exact zeros.
\end{enumerate}

\subsection{Calibration and Vectorized Evaluation}
\begin{enumerate}
    \item \textbf{Calibration Split}: The routing parameters are calibrated on an extremely small mixed-task dataset consisting of exactly 64 images (16 from each task).
    \item \textbf{Objective Function}: The router is optimized using standard Cross-Entropy loss calculated over sample-wise vectorized parameter assembly via PyTorch's batched matrix multiplication, combined with weight decay ($L_2$ regularization):
    \begin{equation}
        \mathcal{L} = \mathcal{L}_{\text{CE}} + \lambda_{\text{reg}} \sum_l \|W^{(l)}\|_2^2
    \end{equation}
    No explicit task-variance penalty ($\mathcal{L}_{\text{VR}}$) is used, as the zero-initialized Softmax architecture acts as the true prior-layer regularizer.
    \item \textbf{Verification of Vectorization Collapse}: We evaluate the trained router on mixed-task image streams under varying deployment batch sizes $B \in \{1, 8, 32, 128, 512\}$. This setup will empirically verify whether unregularized routers suffer from severe Vectorization Collapse at $B=1$, and whether our zero-initialized Softmax prior (\texttt{L3\_Softmax\_WellReg}) successfully stabilizes real-world model merging.
\end{enumerate}

\section{Extension of the Dynamic Routing Paradox to Non-Linear Model Merging}
\label{app:nonlinear}
In this appendix, we analyze how our findings regarding the \emph{Dynamic Routing Paradox} and \emph{Vectorization Collapse} generalize from linear dynamic model merging to non-linear model merging techniques.

\subsection{Non-Linear Model Merging Paradigms}
Unlike linear model merging (which interpolates parameter weights directly via linear coefficients), non-linear model merging techniques resolve parameter alignment and feature mismatch across different model checkpoints by matching intermediate activations or features. Notable examples include:
\begin{itemize}
    \item \textbf{ZipIt} \cite{stoica2023zip}: Matches and merges correlated features and channels across models by computing activation covariance matrices and resolving a bipartite matching problem to align representation spaces.
    \item \textbf{RegMean} \cite{jin2022dataless}: Minimizes the mean squared error (MSE) between the outputs of different models by weighting the parameter average at each layer by the inverse of the activation inner-product (or covariance) matrices.
\end{itemize}

\subsection{The Covariance Estimation Bottleneck under Data Scarcity}
Both ZipIt and RegMean rely heavily on estimating activation covariance matrices $C = \frac{1}{N} X^T X \in \mathbb{R}^{D \times D}$ (where $D$ is the feature dimension and $X \in \mathbb{R}^{N \times D}$ represents the activations over $N$ calibration samples). Under standard low-data calibration splits (e.g., $N=64$ samples), estimating $C$ presents a severe statistical bottleneck:
\begin{enumerate}
    \item \textbf{Ill-Conditioned Covariance}: Because the calibration sample size $N$ is much smaller than the representation dimension $D$ (where $D = 192$ or $768$ in standard backbones), the estimated covariance matrix $C$ is singular (rank-deficient) and highly ill-conditioned.
    \item \textbf{Extreme Sample Noise Sensitivity}: The sample covariance matrix overfits severely to the specific noise of the 64 calibration samples, leading to highly distorted matching or alignment matrices. When these alignment matrices are applied to out-of-sample data on the test stream, they introduce severe functional distortions, dropping the merged model's performance.
\end{enumerate}

\subsection{The Non-Linear Dynamic Routing Paradox}
If we extend non-linear merging to be \emph{dynamic} (sample-dependent, where matching permutations or weighting matrices are computed on the fly per input sample), the matching or routing function has extremely high capacity. Under vectorized sample-wise evaluation ($B=1$), the absence of batch-average smoothing removes the implicit regularizing mask, exposing these non-linear estimators to severe \emph{Vectorization Collapse}.

To stabilize non-linear merging under data scarcity, practitioners must apply heavy regularization to the covariance estimation. Typical techniques include:
\begin{itemize}
    \item \textbf{Shrinkage Estimators}: Shrinking the sample covariance matrix towards a structured target (such as the identity matrix or diagonal variance):
    \begin{equation}
        C_{\text{regularized}} = (1 - \gamma) C + \gamma I
    \end{equation}
    where $\gamma \in [0, 1]$ is the shrinkage coefficient.
    \item \textbf{Prior-Layer Alignment}: Restricting the permuting or matching matrices to stay close to a static maximum-entropy uniform alignment prior (e.g., identity permutation).
\end{itemize}

This heavy regularization leads directly to the non-linear equivalent of the \textbf{Dynamic Routing Paradox}:
\begin{quote}
    \emph{To prevent severe Vectorization Collapse and covariance singular-value explosion under data scarcity, any dynamic or static non-linear merging estimator must be regularized so heavily (via shrinkage or uniform identity priors) that its matching flexibility is restricted to a tight neighborhood around the static uniform baseline. This heavy constraint prevents the non-linear matched model from exploiting its advanced alignment capacity, yielding only a marginal performance gain over naive uniform merging.}
\end{quote}
Thus, our foundational insights regarding proper prior-layer design, regularization, and the limitations of dynamic model ensembling under extreme data scarcity are not artifacts of linear interpolation; they represent universal statistical trade-offs that apply across both linear and non-linear model merging paradigms.

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was created
% by Iain Murray in 2018, and modified by Alexandre Bouchard in
% 2019 and 2021 and by Csaba Szepesvari, Gang Niu and Sivan Sabato in 2022.
% Modified again in 2023 and 2024 by Sivan Sabato and Jonathan Scarlett.
% Previous contributors include Dan Roy, Lise Getoor and Tobias
% Scheffer, which was slightly modified from the 2010 version by
% Thorsten Joachims & Johannes Fuernkranz, slightly modified from the
% 2009 version by Kiri Wagstaff and Sam Roweis's 2008 version, which is
% slightly modified from Prasad Tadepalli's 2007 version which is a
% lightly changed version of the previous year's version by Andrew
% Moore, which was in turn edited from those of Kristian Kersting and
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.
