\begin{abstract}
Dynamic weight-space routing has emerged as a powerful paradigm for merging multi-task neural network experts. However, existing methods suffer from two critical, overlooked failure modes: (1) optimization bloat and out-of-distribution (OOD) overfitting due to redundant trainable routing parameterizations (e.g., in wave-inspired or unregularized multi-layer routers), and (2) \emph{heterogeneity collapse}, where mixed-task deployment streams force batch-wise routing coefficients to average out, destroying task specificity. 
In this paper, we present a relentless application of Occam's razor to model-merging streams. Operating under the modern Parameter-Efficient Fine-Tuning (PEFT/LoRA) paradigm to ensure VRAM viability, we introduce \textbf{Parameter-Free Subspace Routing (PFSR)} combined with \textbf{Micro-Batch Homogenization (MBH)}---a zero-shot, completely non-parametric framework that contains \textbf{zero trainable parameters} and requires \textbf{zero calibration split data} in standard classification registries (introducing only a minor, low-resource calibration dependency when enabling density-based OOD rejection or non-classification centroid fallbacks). We explicitly highlight that our framework possesses a strict dependency on the PEFT (LoRA) representation of experts to guarantee spatial VRAM viability, as full-parameter dynamic weight merging on-the-fly would incur prohibitive PCIe host-to-device loading delays. 
PFSR projects high-dimensional intermediate representations from the penultimate layer onto a frozen low-dimensional task coordinate subspace using cosine similarity against pre-trained expert classification weights, deriving routing coefficients directly via temperature-scaled normalization. To shield the model from heterogeneity collapse, MBH dynamically partitions mixed-task streams into homogeneous micro-batches on the fly, allowing specialized model merging without batch-averaging degradation. 
On a high-fidelity synthetic sandbox, PFSR+MBH completely resolves both OOD overfitting and heterogeneity collapse, achieving a high \textbf{75.00\% Joint Mean accuracy} under homogeneous streams and remaining highly robust across all configurations. Our minimalist approach systematically outperforms complex wave-inspired routing pipelines (which collapse to 47.50\%) with zero parameter and optimization overhead.
\end{abstract}
