\section{Conclusion}
\label{sec:conclusion}

In this paper, we presented \textbf{LoRA Subspace Projection Routing (LSPR)}, a minimalist and mathematically elegant approach to dynamic model ensembling and serving. By rejecting the escalating complexity of recent state-of-the-art frameworks—such as offline calibration splits, Unit-Norm and Dispersion Calibration parameters, EM-fitted Gaussian Mixture Models, and classification-head dependencies—LSPR demonstrates that the intrinsic geometric structure of frozen LoRA weights is completely sufficient to solve the routing and OOD rejection problems.

Using a microsecond-level offline QR decomposition of early-layer down-projection matrices, LSPR extracts orthonormal bases representing each task's intrinsic representational subspace. At inference time, routing and OOD rejection are resolved simultaneously in a single parallel pass by projecting early activations onto these task-specific subspaces, yielding scale-invariant and highly robust similarity coordinates. 

Empirically, evaluated under a fully-trained PyTorch multi-task environment with domain-shifted representations, LSPR achieves an outstanding 85.81\% Joint Mean Accuracy (outperforming Uniform Merging by 61.85\% and perfectly recovering the expert ceiling), is completely immune to heterogeneity collapse, detects out-of-distribution queries with an outstanding zero-shot AUROC of 0.9755, and delivers highly efficient physical latency scaling on host CPUs. By validating the power of simple linear algebra, LSPR serves as a compelling proof-of-concept that elegant, deployment-time training-free and data-free mathematical projections can match the performance of highly over-engineered, data-dependent parametric pipelines.

\subsection{Limitations and Future Scaling Roadmap}
\label{subsec:conclusion_roadmap}

While LSPR offers exceptional systems efficiency and geometric routing stability, we acknowledge its key limitations and propose a concrete, step-by-step scaling roadmap to transition LSPR from a controlled proof-of-concept to large-scale, commercial foundation models:

\begin{enumerate}
    \item \textbf{High-Dimensional Transformer Adaptation:} In this work, LSPR was evaluated inside the Isolating Coordinate Sandbox (ICS). When scaling to multi-layer autoregressive LLMs (e.g., Llama-3-8B with $D=4096$), selecting the routing layer $L_{\text{route}}$ is a critical design choice. Because semantic task separation typically emerges in middle layers, we propose placing $L_{\text{route}}$ around Layer 8 to 12. This allows early layers to process shared syntactic features before subspace projection occurs, keeping downstream routing highly separated.
    \item \textbf{Token-Filtered Reconstruction Loss:} On massive pre-training or instruction-tuning corpora, computing the reconstruction loss $\mathcal{L}_{\text{reconstruction}}$ for every token is computationally expensive. To manage this training-time overhead, we propose a \textit{token-filtering strategy}: only compute the joint reconstruction loss on a random 5--10\% sub-sample of sequence tokens, or filter tokens based on their activation magnitudes (discarding padding, punctuation, and uninformative low-norm tokens). This dramatically reduces training-time FLOPs while preserving representational alignment.
    \item \textbf{Quantized Subspace Matching:} For massive multi-task registries (e.g., $K \ge 100$ experts), storing the semi-orthogonal matrices $Q_k \in \mathbb{R}^{D \times r}$ in FP16 could consume non-trivial cache memory. We propose quantizing $Q_k$ to 4-bit or 8-bit integer formats. Since the subspace projection $h_b Q_k$ is a standard inner product, integer-quantized projections can be executed on edge tensor accelerators with negligible loss of routing precision.
    \item \textbf{Large-Scale Benchmarking:} Future work will focus on empirical evaluation on full-sized Transformers (e.g., ViT-B-32 and Llama-3) on standard NLP and vision benchmarks (such as GLUE, SuperGLUE, and VTAB), validating LSPR's zero-shot routing and OOD rejection capabilities in real-world large-scale serving environments.
\end{enumerate}
