The serving of multiple task-specific low-rank experts (e.g., LoRA) on resource-constrained edge hardware requires dynamic model merging and sample-wise routing. Recent state-of-the-art frameworks employ early-layer activation centroids to map queries into low-dimensional coordinate spaces, utilizing diagonal Gaussian Mixture Models (GMMs) for out-of-distribution (OOD) task rejection to avoid routing collapse. In this paper, we deconstruct the hidden methodological assumptions of these coordinate density models. We reveal a severe overfitting vulnerability: fitting diagonal GMMs on low-resource calibration splits ($N \le 64$) yields highly unstable variance estimates that collapse under mild covariate shift (representation noise). 

While small-scale registries with low coordinate dimensions ($K=4$) are highly competitive and easily handled by standard models, as task registries scale to support larger multi-tenant serving networks ($K \ge 16$), unregularized models suffer from severe overfitting due to the curse of dimensionality. To resolve this instability, we propose {\bf SRC-DE} (Shrinkage-Regularized Coordinate Density Estimation), which applies analytical, parameter-free Ledoit-Wolf-style covariance shrinkage as a post-fit regularization step immediately following EM convergence.

We establish a rigorous, symmetric evaluation protocol where representation-level noise is applied equally to both in-distribution (ID) and OOD test features, resolving a critical unequal-noise confounder present in prior evaluations. We also identify a silent caching bug in standard Python statistical software (\texttt{scikit-learn})'s GMM implementation that renders manual covariance regularizations ineffective. Our empirical evaluation across high-conflict vision task registries (MNIST, FashionMNIST, CIFAR-10, SVHN) and controlled scaling simulations demonstrates that unregularized GMMs exhibit severe estimation variance under covariate shift and small calibration splits. Our proposed SRC-DE dramatically stabilizes these density boundaries, reducing estimator variance by over half and achieving up to {\bf +4.63\% absolute AUC improvement} under high-dimensional task registries. By exposing and resolving this low-resource variance collapse, we establish a robust and mathematically sound framework for large-scale dynamic model serving on the edge.
