\section{Conclusion}
\label{sec:conclusion}

In this paper, we introduced {\bf Barycentric Proximity-Anchored Merging (BPAM)} and presented a transparent, critical deconstructive audit of test-time adaptive model merging. Guided by the core philosophy of **The Minimalist**, we systematically isolated and analyzed the factors that drive performance gains in test-time adaptation, mapping the boundaries of parameter frugality.

Our findings reveal a critical scientific lesson for the model-merging community. While high-capacity weight-space adaptation methods (such as AdaMerging) achieve substantial, genuine weight-space alignment gains due to their layer-wise parameters, extremely low-parameter regimes (like BPAM's 8 global task-scalars) are too constrained to resolve fine-grained parameter conflicts on their own. Under this restricted regime, auxiliary classification head adaptation becomes the primary driver of performance, shifting the optimization burden from weight alignment to decision boundary adaptation. By evaluating BPAM across three distinct spatial configurations, we showed that restricting merging to a single projection bottleneck (BPAM-Restricted) collapses performance to 51.38\%, proving that whole-model coordinate blending is essential. When the whole model is merged under frozen heads, our true 8-parameter variant (BPAM-Static) achieves 69.21\% average accuracy, matching linear Task Arithmetic. Concurrently tuning classification heads (BPAM-Full) yields 75.22\% average accuracy.

Ultimately, our work highlights that elegant mathematical constraints—specifically, a Convex Barycentric Simplex Projection and a closed-form Mean-Field Proximity Penalty—can deliver robust multi-task weight blending with a near-zero parameter footprint, serving as a vital boundary baseline for the community. We hope our transparent, deconstructive approach encourages a shift in future research toward mapping the precise trade-offs between parameter density and multi-task capability, steering the community toward mathematically grounded optimization boundaries.
