\begin{abstract}
Test-time model merging offers a powerful paradigm for combining multiple task-specific experts into a single cohesive network without additional training. However, when adapting merging coefficients on small, unlabeled test streams, models frequently suffer from the \emph{Overfitting-Optimizer Paradox}---where high-dimensional, unconstrained optimization leads to severe transductive collapse and representational decay on out-of-distribution data. Prior solutions address this by introducing complex spatial regularizers, elastic weight constraints, or continuous polynomial trajectories. Under the guiding principle of Occam's razor, we argue that such methods are unnecessarily convoluted and heavily dependent on delicate hyperparameter tuning. In this paper, we propose \textbf{Pruned Gradient Merging (PG-Merge)}, a training-free, non-parametric, and highly elegant alternative. PG-Merge introduces a simple, dynamic sparse gradient mask that freezes the vast majority ($85\%$) of merging coefficients, adapting only the most critical, high-sensitivity coordinates at each test-time step. We show that this simple restriction of optimization degrees of freedom acts as an extremely effective low-pass filter, preventing high-frequency parameter drift and local overfitting. Our comprehensive evaluation on a Vision Transformer (\texttt{vit\_tiny}) across multiple vision tasks (MNIST, FashionMNIST, CIFAR-10, and SVHN) demonstrates that PG-Merge substantially outperforms unregularized AdaMerging and matches or exceeds complex state-of-the-art regularizers, all while adding zero hyperparameter bloat or computational overhead.
\end{abstract}