\section{Conclusion and Future Work}
\label{sec:conclusion}

In this paper, we introduced \textbf{Pruned Gradient Merging (PG-Merge)}, a training-free, non-parametric, and highly elegant active model merging framework. PG-Merge directly confronts the Overfitting-Optimizer Paradox in test-time model fusion, where high-dimensional merging coefficients overfit to local, unlabeled test-stream statistics. Guided by the principle of Occam's razor, we demonstrated that the complex, hyperparameter-heavy spatial regularizers and geometric polynomial restrictions proposed in prior SOTA methods are entirely redundant. Instead, a simple binary sparse gradient mask that freezes $85\%$ of the coefficients and allows only the top-$15\%$ most sensitive parameters to update is fully sufficient to prevent overfitting.

Our extensive experiments on a Vision Transformer (\texttt{vit\_tiny}) across four visual datasets (MNIST, FashionMNIST, CIFAR-10, and SVHN) show that PG-Merge dramatically outperforms unconstrained AdaMerging and matches or exceeds SOTA regularizers, without adding any optimization bloat or extra hyperparameters to tune. 

In future work, we plan to evaluate the scalability of PG-Merge on larger foundation networks, including large language models (LLMs) and diffusion models. Given its training-free nature, extreme computational efficiency, and ease of implementation, we believe PG-Merge provides a clean, robust, and highly practical blueprint for real-world multi-task model fusion on-the-fly. We hope this work encourages the community to favor elegant, simplified solutions over unnecessary architectural and optimization complexity.