\begin{abstract}
Dynamic model merging has emerged as a key mechanism for adapting multi-task neural architectures to streaming, heterogeneous workloads on edge hardware without retraining. 
However, existing dynamic ensembling methods are fundamentally bottlenecked by their stateless, layer-wise decoupled execution. 
Under noisy, rapidly switching sample streams, these approaches cause high-frequency ensembling coefficient oscillations (routing weight jitter), representation saturation, and sharp activation spikes across layers. 
Conversely, systems-level scheduling wrappers like Micro-Batch Homogenization (MBH) safeguard representation stability but introduce a prohibitive $O(K)$ sequential latency penalty on edge hardware. 
In this work, we present \textbf{ChemMerge}, a training-free, continuous-time paradigm that completely rethinks dynamic model ensembling through the lens of non-equilibrium chemical reaction kinetics. 
By modeling representation flow through the network's depth as a multi-component chemical reactor, we treat task experts as reactive species and early-layer representations as catalytic enzymes. 
ChemMerge tracks continuous, sample-wise expert concentration states that evolve across sequential layers governed by a system of first-order kinetics differential equations. 
Our multi-seed evaluation within a high-fidelity Analytical Coordinate Sandbox (ICS) (simulating MNIST, Fashion-MNIST, CIFAR-10, and SVHN streams) demonstrates that ChemMerge recovers \textbf{98.81\%} of the Expert Oracle ceiling, achieving ensembling accuracy statistically comparable to state-of-the-art activation blending approaches while outperforming stateless nearest-centroid baselines by up to \textbf{+8.22\%} under a stateless, constant $O(1)$ edge latency. 
Crucially, on real pre-trained Vision Transformers ($\text{ViT-B/16}$), we perform a routing-only validation on frozen activations and show that ChemMerge reduces layer-to-layer ensembling weight routing jitter by up to \textbf{9.9$\times$} compared to stateless nearest-centroid routing and over \textbf{2.15$\times$} compared to SABLE (under identical sensitivities). 
While absolute classification accuracies are statistically comparable between the best-performing ensembling methods, ChemMerge solves the critical accuracy-stability trade-off, bringing immense physical stability to deep representation trajectories without any sequential latency overhead.
\end{abstract}
