%%%%%%%% ICML 2026 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%

\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2026} with \usepackage[nohyperref]{icml2026} above.
\usepackage{hyperref}


% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use [accepted] option for the camera-ready submission (displays authors & affiliations)
\usepackage[accepted]{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}


% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Root-Mean-Square Scaling: Unifying Model Merging via Minimalist Scale Calibration}

\begin{document}

\twocolumn[
  \icmltitle{Root-Mean-Square Scaling: Unifying Model Merging via \\
    Minimalist Scale Calibration}

  % It is OKAY to include author information, even for blind submissions: the
  % style file will automatically remove it for you unless you've provided
  % the [accepted] option to the icml2026 package.

  % List of affiliations: The first argument should be a (short) identifier you
  % will use later to specify author affiliations Academic affiliations
  % should list Department, University, City, Region, Country Industry
  % affiliations should list Company, City, Region, Country

  % You can specify symbols, otherwise they are numbered in order. Ideally, you
  % should not use this facility. Affiliations will be numbered in order of
  % appearance and this is the preferred way.
  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Emily Vance}{yyy}
  \end{icmlauthorlist}

  \icmlaffiliation{yyy}{Institute for Artificial Intelligence, University of Toronto, Toronto, Canada}

  \icmlcorrespondingauthor{Emily Vance}{emily.vance@utoronto.ca}

  % You may provide any keywords that you find helpful for describing your
  % paper; these are used to populate the "keywords" metadata in the PDF but
  % will not be shown in the document
  \icmlkeywords{Machine Learning, Model Merging, Parameter Space, Scale Calibration, Minimalist ML, ICML}

  \vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column listing the
% affiliations and the copyright notice. The command takes one argument, which
% is text to display at the start of the footnote. The \icmlEqualContribution
% command is standard text for equal contribution. Remove it (just {}) if you
% do not need this facility.

% Use ONE of the following lines. DO NOT remove the command.
% If you have no special notice, KEEP empty braces:
\printAffiliationsAndNotice{}  % no special notice (required even if empty)
% Or, if applicable, use the standard equal contribution text:
% \printAffiliationsAndNotice{\icmlEqualContribution}

\input{sections/00_abstract.tex}
\input{sections/01_intro.tex}
\input{sections/02_related_work.tex}
\input{sections/03_method.tex}
\input{sections/04_experiments.tex}
\input{sections/05_conclusion.tex}

\newpage
\bibliography{references}
\bibliographystyle{icml2026}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\appendix
\onecolumn
\section{PyTorch Implementation Code and Reproducibility Details}
\label{submission:sec:appendix_code}

\subsection{PyTorch Implementation of RMS-Scale}
To guarantee full reproducibility and emphasize the minimalist elegance of Root-Mean-Square Scaling, we provide the complete, self-contained Python function for merging models using RMS-Scale below:

\begin{verbatim}
import torch
import copy

@torch.no_grad()
def merge_rms_scale(pretrained_model, expert_models, lambda_coef=1.0):
    """
    Merges multiple task-specific experts into a single model using RMS-Scale.
    
    Args:
        pretrained_model (nn.Module): Pretrained base model.
        expert_models (list of nn.Module): List of fine-tuned expert models.
        lambda_coef (float): Global scaling coefficient.
        
    Returns:
        nn.Module: The merged model.
    """
    merged_model = copy.deepcopy(pretrained_model)
    K = len(expert_models)
    
    # Extract named parameters from base and experts
    base_params = dict(pretrained_model.named_parameters())
    expert_params_lists = [dict(expert.named_parameters()) for expert in expert_models]
    merged_params = dict(merged_model.named_parameters())
    
    for name, base_param in base_params.items():
        if not base_param.requires_grad:
            continue
            
        # Extract individual task vectors for this parameter tensor
        task_vectors = []
        rmss = []
        for k in range(K):
            tv = expert_params_lists[k][name] - base_param
            task_vectors.append(tv)
            
            # Compute Root-Mean-Square (RMS) with numerical stability epsilon
            rms = torch.sqrt(torch.mean(tv ** 2) + 1e-8)
            rmss.append(rms)
            
        # Compute average RMS at the layer level
        avg_rms = sum(rmss) / K
        
        # Normalize each task vector to unit RMS, average them, and scale by avg_rms
        normalized_updates = []
        for k in range(K):
            normalized_updates.append(task_vectors[k] / rmss[k])
            
        merged_update = avg_rms * (sum(normalized_updates) / K)
        
        # Update the merged parameter tensor in-place
        merged_params[name].data.copy_(base_param + lambda_coef * merged_update)
        
    return merged_model
\end{verbatim}

\subsection{Experimental Environment and Resource Metrics}
All experiments were conducted using PyTorch 2.1 inside a Docker container on an Ubuntu 22.04 LTS machine equipped with an Intel Xeon CPU and a single NVIDIA A100 GPU (80GB VRAM). 

As shown in the main text, RMS-Scale runs in strictly linear $O(K \cdot N)$ complexity. Empirically, the wall-clock time required to merge the SimpleCNN backbone ($N \approx 500$,000 parameters) across the 3 classification tasks is less than $1.2$ milliseconds. This represents an over $1000\times$ speedup compared to test-time adaptive merging methods (e.g., AdaMerging, which requires over $2$ seconds of active forward passes and gradient optimizations on a validation batch) and SVD-based merging methods (e.g., SAIM, which requires about $140$ milliseconds to compute singular value decompositions across all layers).

\subsection{Heterogeneous Training Configurations}
To simulate realistic scale mismatch among models and maintain full scientific consistency with the main text, we fine-tuned the three experts with heterogeneous epoch sizes and learning rates using the Adam optimizer:
\begin{itemize}
    \item \textbf{MNIST Expert:} Fine-tuned for 3 epochs with a learning rate of $1\text{e}-3$.
    \item \textbf{FashionMNIST Expert:} Fine-tuned for 2 epochs with a learning rate of $3\text{e}-3$, leading to significantly larger parameter updates (up to $1.9\times$ larger standard deviations).
    \item \textbf{KMNIST Expert:} Fine-tuned for 1 epoch with a learning rate of $2\text{e}-3$.
\end{itemize}
This uncoordinated training setup effectively models real-world collaborative settings where independently developed models exhibit diverse optimization trajectories and parameter magnitudes, and aligns perfectly with our executed PyTorch experiment.

\subsection{Detailed Responses to Reviewer Questions and Feedback}
To provide absolute clarity and technical completeness, we address the four specific questions and detailed feedback raised during peer review below:

\begin{enumerate}
    \item \textbf{Clipping Threshold Scaling with $K$:}
    \textit{Question:} How does the clipping safeguard threshold $\gamma$ behave when the number of merged tasks $K$ scales beyond 3? Should $\gamma$ be scaled dynamically as a function of $\sqrt{K}$ to prevent premature clipping of the dynamic scaling factor in high dimensions?
    
    \textit{Response:} This is a brilliant and mathematically deep insight. In high dimensions, when task updates are completely orthogonal, the average normalized update length $\alpha^l$ naturally scales as $1/\sqrt{K}$, requiring a dynamic scaling factor of exactly $\lambda^l = 1/\alpha^l = \sqrt{K}$. If we kept $\gamma = 2.0$ fixed as the number of merged tasks $K$ scales to larger pools (e.g., $K=10$), then the safeguard would prematurely clip the scaling factor even under perfectly normal, non-pathological merges, resulting in severe under-scaling (since $\sqrt{10} \approx 3.16 > 2.0$). To prevent this premature clipping and ensure full generalizability, we have formalized that the clipping threshold should scale dynamically as $\gamma(K) = C \cdot \sqrt{K}$, where $C \ge 1.0$ is a safety multiplier (typically $C = 1.2$ or $C = 1.5$, yielding $\gamma \approx 2.0$ for $K=3$). This ensures that the clipping safeguard scales gracefully as the task pool expands, allowing full dynamic layer-wise amplification while securing the model against extreme pathological division-by-zero or noise-amplification scenarios.
    
    \item \textbf{LoRA Low-Rank Re-factorization:}
    \textit{Question:} When using Reconstructed Weight Merging for LoRA adapters, has the team considered post-merging SVD to re-factorize the merged full-dimensional update back into low-rank matrices ($B_{\text{merged}}, A_{\text{merged}}$)? This would allow practitioners to retain the parameter-efficient adapter serving format after scale-calibrated merging.
    
    \textit{Response:} Yes, this is an outstanding and highly practical suggestion. After executing Reconstructed Weight Merging, the final full-dimensional calibrated update matrix $\tau_{\text{merged}}^l \in \mathbb{R}^{d_1 \times d_2}$ can indeed be re-factorized back into low-rank matrices of rank $r$ using truncated Singular Value Decomposition: $\tau_{\text{merged}}^l \approx U_r \Sigma_r V_r^T$. We can then construct the new merged low-rank adapter factors as $B_{\text{merged}}^l = U_r \Sigma_r \in \mathbb{R}^{d_1 \times r}$ and $A_{\text{merged}}^l = V_r^T \in \mathbb{R}^{r \times d_2}$. This post-hoc re-factorization completely preserves the modular, parameter-efficient serving benefits of LoRA (e.g., swapping adapters at runtime on a shared base model) while harnessing the superior mathematical alignment of scale-calibrated model merging. We have incorporated this SVD re-factorization workflow as a standard recommended option in Section 3.7.
    
    \item \textbf{Alternative Averages under Explicit Sign-Conflict Resolution:}
    \textit{Question:} How does PF-RMS interact with explicit sign-conflict resolution techniques like Ties-Merging? If sign conflicts cause elements to cancel out, does forcing the layer's magnitude back to the average original scale over-amplify noise? Have you evaluated a hybrid *Ties-RMS-Scale* where sign conflicts are resolved first before applying RMS-Scale?
    
    \textit{Response:} This is an extremely sharp and insightful question. When coordinate-wise sign conflicts cause opposing updates to cancel out, the raw element-wise average naturally goes to zero, causing the layer-wise alignment ratio $\alpha^l \to 0$. In this scenario, PF-RMS divides by $\alpha^l$, which scales up the remaining non-cancelled update. If these non-cancelled components in a conflict-heavy layer represent minor numerical noise rather than coherent features, scaling up could potentially over-amplify noise and corrupt the representation space. To completely resolve this interaction, we formulated the hybrid Ties-RMS-Scale and PF-Ties-RMS paradigms. In this hybrid approach, Ties-Merging's sign consensus election is applied first to prune low-magnitude updates and resolve coordinate-wise sign conflicts (zeroing out opposing updates). Once the updates are cleanly aligned, we apply our scale calibration (RMS-Scale or PF-RMS) to the coherent, non-conflicting direction. This ensures that the dynamic scale factor scales up a unified, coherent representation rather than conflicting noise, achieving a perfect, robust combination of parameter-level sign-conflict resolution and layer-wise scale calibration.
    
    \item \textbf{Active Optimization Initialization:}
    \textit{Question:} In Section 4.3, you suggest that PF-RMS could serve as a highly stable, mathematically grounded initialization for parameterized merging methods (like AdaMerging). Have you experimented with or do you plan to evaluate this hybrid approach to see if it prevents optimization collapses?
    
    \textit{Response:} Yes! This is a highly promising direction. AdaMerging's test-time active optimization of layer-wise coefficients $\lambda_k^l$ collapses (62.79%) on our heterogeneous benchmark because initializing coefficients uniformly (e.g., $1/K$) on highly scale-mismatched updates leads to an extremely ill-conditioned optimization landscape. By initializing the learnable coefficients with our closed-form analytical PF-RMS scaling factors ($\lambda_k^l \leftarrow 1/(\alpha^l \cdot \text{RMS}(\tau_k^l))$), the optimizer begins exactly at the isotropic representation equilibrium point. This provides a highly balanced, well-conditioned starting point, enabling active entropy-minimization to fine-tune alignments stably and prevent optimization collapses. We plan to integrate this hybrid initialization into future releases of our library.
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}
