\begin{abstract}
Test-time model ensembling offers a flexible paradigm for serving multiple expert models under sequential, heterogeneous query streams. However, existing stateless dynamic routers suffer from a severe \emph{routing jitter paradox}, where high-frequency switching between experts in response to query-level noise destabilizes representation flows across layers. Conversely, stateful heuristic routers reduce jitter but lack mathematical stability or generalization guarantees. In this work, we propose \textbf{PAC-Kinetics}, the first learning-theoretic framework for stateful, sequential model ensembling. By formulating representation dynamics as a continuous-time non-equilibrium chemical kinetics system, we derive a stateful discrete recurrence that acts as a robust low-pass filter over sequential representations. We prove that this dynamical system is globally asymptotically stable and input-to-state stable (ISS). Crucially, to model sequential temporal dependencies, we derive a novel, parameter-space PAC-Bayesian generalization bound for stationary $\beta$-mixing stochastic processes. We directly minimize this bound to optimize all routing and kinetics parameters. On both orthogonal and overlapping manifold configurations in our Analytical Coordinate Sandbox, PAC-Kinetics slashes routing jitter by over \textbf{11.2$\times$} on orthogonal streams and up to \textbf{16.0$\times$} on overlapping streams compared to stateless SABLE, while matching or exceeding oracle-level joint accuracy (up to \textbf{95.07\%}). We explicitly address the fundamental stateful-stateless trade-off (where stateful memory acts as a liability under uncorrelated streams) by incorporating an Adaptive Online Kinetics mechanism that dynamically suppresses routing lag under rapid switches, successfully balancing stability and responsiveness and establishing a rigorous mathematical foundation for stable test-time model serving.
\end{abstract}
