\section{Related Work}
\label{sec:related}

\subsection{Parameter-Efficient Fine-Tuning and Multi-Task Serving}
The rapid proliferation of large-scale pre-trained models has made full-parameter fine-tuning for every downstream task computationally prohibitive. Parameter-Efficient Fine-Tuning (PEFT) techniques—such as Low-Rank Adaptation (LoRA)~\cite{hu2021lora}, adapters~\cite{houlsby2019parameter}, and prefix-tuning~\cite{li2021prefix}—alleviate this burden by updating a tiny fraction of the parameters while keeping the backbone frozen. Under edge serving conditions, multiple specialized downstream experts must be deployed concurrently on a shared hardware budget~\cite{chen2023}. Mixture-of-Experts (MoE)~\cite{shazeer2017outrageously, fedus2021switch} and dynamic parameter merging~\cite{wortsman2022robust} provide frameworks for dynamically combining these specialists. However, traditional MoE systems route tokens independently, ignoring sequence-level task structure and suffering from token-routing instability (jitter)~\cite{malkin2023}.

\subsection{Stateless vs. Stateful Dynamic Model Merging}
To enable seamless multi-task edge serving, recent works have explored test-time dynamic model merging, which interpolates expert parameters or activations layer-by-layer on-the-fly.
\textbf{SABLE}~\cite{sable2025} introduces nearest-centroid routing in raw feature space, utilizing a small calibration set to construct layer-wise centroids for each task. Although SABLE is highly plastic and training-free, it evaluates each layer and sample statelessly. As a result, representation-space noise causes extreme layer-to-layer ensembling weight oscillations (routing jitter), degrading overall accuracy.

To resolve routing jitter, state-of-the-art methods introduce stateful propagation of routing coefficients:
\begin{enumerate}
    \item \textbf{Momentum-Merge (Spatial-only)}~\cite{momentummerge}: deconstructs the spatial routing path, proving that depth-wise representation flow can be stabilized by applying a spatial Exponential Moving Average (EMA) to the ensembling weights across layers. However, Momentum-Merge resets its state completely for every new sample, remaining highly vulnerable to temporal stream-level noise.
    \item \textbf{PAC-Kinetics (Temporal-only)}~\cite{pac_kinetics}: addresses sequence-level transitions using a first-order state-space model that propagates ensembling weights temporally across sequential samples. Crucially, PAC-Kinetics relies on learning-theoretic optimization of PAC-Bayesian bounds to fit state matrices. To avoid prohibitive depth-wise overhead, it computes a single routing vector at an early layer and forces all subsequent adapted layers to use the identical weights, completely losing depth-wise local representational alignment and centroid calibration.
    \item \textbf{ChemMerge (Biochemical Stateful Merging)}~\cite{chemmerge}: attempts a unified approach by modeling routing trajectory as non-equilibrium chemical kinetics with Arrhenius rate equations, virtual integration steps, and activation energies. While ChemMerge smooths trajectories in both depth and time, it introduces severe computational complexity: solving continuous-time ODEs online at edge devices incurs high latency, and calibrating five or more uninterpretable parameters is highly sensitive and complex.
\end{enumerate}

\subsection{Our Minimalist Perspective}
In this work, we take a step back from the prevailing trend of building increasingly parameterized dynamical models (e.g., biochemical ODEs or PAC-Bayesian learned recurrences). We argue that the success of these systems resides primarily in their underlying bilinear smoothing properties rather than their physical or theoretical packaging. 

By applying Occam's razor, we show that we can unify the benefits of spatial and temporal smoothing into a single, training-free, 2D bilinear recursive filter (2D-STEM). Our framework is mathematically transparent, has zero online optimization loops, and achieves competitive noise reduction and task-switching accuracy by replacing continuous-time numerical solvers with a single-line arithmetic update.
