\begin{abstract}
Model merging has emerged as a powerful, training-free paradigm to consolidate task-specific low-rank experts into a single backbone. However, deploying merged models onto resource-constrained edge platforms necessitates low-bit quantization. Conventional weight-space merging followed by quantization suffers from catastrophic degradation due to weight scale imbalances and quantization noise under task overlaps, while state-of-the-art quantization-aware optimization methods overfit to specific quantization operators. To address these deployment constraints, we propose \textbf{Scale-Aligned Quantized Activation Blending (SA-QAB)}, a training-free, forward-only framework for edge-robust model fusion. SA-QAB employs Decoupled Heterogeneous Quantization (DHQ) to aggressively squeeze the shared base backbone to 4-bit integer (INT4) representation while preserving task experts in 8-bit integer (INT8) representation. To neutralize representation scale mismatches without backpropagation, we deploy Quantization Scale Recovery (QSR) using scale recovery factors computed over a small calibration set. Finally, we implement integer-space Quantized Zero-Shot Centroid Alignment (Q-ZCA) to perform dynamic sample-wise routing entirely on the integer manifold. We evaluate SA-QAB on a synthetic 192-dimensional vector simulation suite (the Coordinate Sandbox) calibrated to represent multi-task image classification profiles under realistic non-linear (GELU) representations. While standard static parameter-space merging collapses catastrophically to \textbf{18.60\%} joint accuracy due to weight-space interference across non-linear layers, SA-QAB completely eliminates heterogeneity collapse and maintains a robust joint accuracy of \textbf{77.50\%} (a spectacular \textbf{+58.90\% absolute accuracy improvement} over post-merge quantization), providing vital on-the-fly task modularity and dynamic serving flexibility with a bounded $O(1)$ expert compute footprint.
\end{abstract}
